---
title: "Inference & t-tests"
subtitle: "PS50008A Week 14 | Lecture"
author: "Gordon Wright"
date: "2026-02-04"
format:
  revealjs:
    theme: [default, ../../assets/slides.scss]
    slide-number: c/t
    progress: true
    transition: fade
    width: 1050
    height: 700
    center: false
    center-title-slide: true
execute:
  echo: false
  warning: false
  message: false
---

## Learning Outcomes

By the end of this lecture you will be able to:

- Explain the logic of null hypothesis significance testing
- Interpret p-values correctly
- Distinguish between independent and paired t-tests
- Choose the appropriate t-test for different research designs

# The Logic of Hypothesis Testing {background-color="#275882"}

## The Fundamental Question

<!-- TODO: Introduce the core problem -->

::: {.fragment}
When we observe a difference in our data...

**Is it a real effect, or just random chance?**
:::

## Null Hypothesis Significance Testing (NHST)

<!-- TODO: Define NHST framework -->

::: {.definition-box}
**NHST**: A statistical framework that tests whether observed data are consistent with a null hypothesis of "no effect" or "no difference".
:::

## The Two Hypotheses

<!-- TODO: Explain H0 and H1 -->

:::: {.columns}

::: {.column width="50%"}
**Null Hypothesis (H0)**

- Assumes no effect/difference
- The "skeptic's position"
- What we try to reject

*Example: "There is no difference in memory scores between groups."*
:::

::: {.column width="50%"}
**Alternative Hypothesis (H1)**

- Assumes there IS an effect
- The researcher's prediction
- What we hope to support

*Example: "Memory scores differ between groups."*
:::

::::

## The Logic: Proof by Contradiction

<!-- TODO: Explain the reasoning -->

::: {.incremental}
1. Assume the null hypothesis is true
2. Calculate how likely our data would be under H0
3. If our data would be very unlikely under H0...
4. We reject H0 and accept H1
:::

::: {.fragment}
::: {.example-box}
Like a court trial: We assume innocence (H0) until evidence (data) proves guilt beyond reasonable doubt.
:::
:::

## The Sampling Distribution

<!-- TODO: Introduce sampling distributions -->

::: {.definition-box}
**Sampling Distribution**: The distribution of a statistic (e.g., the mean) across all possible samples of a given size from a population.
:::

::: {.fragment}
<!-- TODO: Add visualisation -->
:::

# p-values and Significance {background-color="#275882"}

## What is a p-value?

<!-- TODO: Define p-value properly -->

::: {.definition-box}
**p-value**: The probability of obtaining results at least as extreme as those observed, assuming the null hypothesis is true.
:::

::: {.fragment}
::: {.callout-warning}
## NOT what a p-value means
- The probability that H0 is true
- The probability the results occurred by chance
- The probability of replication
:::
:::

## Interpreting p-values

<!-- TODO: Add interpretation guide -->

::: {.example-box}
**p = 0.03 means:**

"If there truly were no effect, we would observe results this extreme or more extreme only 3% of the time."
:::

## The Significance Threshold (alpha)

<!-- TODO: Discuss alpha level -->

::: {.definition-box}
**Alpha (alpha)**: The predetermined threshold for rejecting H0. Conventionally set at 0.05.
:::

::: {.fragment}
| p-value | Decision |
|---------|----------|
| p < .05 | Reject H0 ("significant") |
| p >= .05 | Fail to reject H0 ("not significant") |
:::

## Problems with p-values

<!-- TODO: Discuss limitations -->

::: {.incremental}
- Arbitrary threshold (why .05?)
- Dichotomous thinking (significant vs not)
- Affected by sample size
- Often misinterpreted
- Does not indicate effect size
:::

## Types of Errors

<!-- TODO: Add Type I and Type II errors -->

|  | H0 True | H0 False |
|--|---------|----------|
| **Reject H0** | Type I Error (false positive) | Correct! |
| **Fail to reject H0** | Correct! | Type II Error (false negative) |

::: {.fragment}
- **Type I error rate** = alpha (usually .05)
- **Type II error rate** = beta (related to power)
:::

# Independent t-tests (Between-Subjects) {background-color="#275882"}

## When to Use Independent t-tests

<!-- TODO: Define use cases -->

::: {.definition-box}
**Independent samples t-test**: Compares means from two separate, unrelated groups.
:::

::: {.fragment}
**Use when:**

- Two different groups of participants
- Each participant is in only one condition
- Comparing a continuous outcome variable
:::

## The Between-Subjects Design

<!-- TODO: Add design illustration -->

::: {.example-box}
**Example:** Does caffeine affect reaction time?

- Group 1: 30 participants receive caffeine
- Group 2: 30 different participants receive placebo
- DV: Reaction time (ms)
:::

## The t-statistic Formula

<!-- TODO: Add formula explanation -->

$$t = \frac{\bar{X}_1 - \bar{X}_2}{SE_{difference}}$$

::: {.fragment}
Where $SE_{difference}$ accounts for variability in both groups.
:::

## Assumptions of Independent t-test

<!-- TODO: List assumptions -->

::: {.incremental}
- Independence of observations
- Continuous dependent variable
- Normal distribution (or large n)
- Homogeneity of variance (Levene's test)
:::

## Example: Independent t-test

<!-- TODO: Add worked example -->

```{r}
#| eval: false
# TODO: Add R code for independent t-test
```

::: {.exercise-box}
**Research question:** Do students who study with music perform differently on exams than those who study in silence?
:::

# Paired t-tests (Within-Subjects) {background-color="#275882"}

## When to Use Paired t-tests

<!-- TODO: Define paired t-tests -->

::: {.definition-box}
**Paired samples t-test**: Compares means from two related measurements (same participants measured twice, or matched pairs).
:::

::: {.fragment}
**Use when:**

- Same participants in both conditions
- Pre-test/post-test designs
- Matched pairs of participants
:::

## The Within-Subjects Design

<!-- TODO: Add design illustration -->

::: {.example-box}
**Example:** Does a relaxation technique reduce anxiety?

- Same 30 participants
- Measure 1: Anxiety before technique
- Measure 2: Anxiety after technique
- Compare within each person
:::

## Why Paired Tests are More Powerful

<!-- TODO: Explain power advantage -->

::: {.incremental}
- Controls for individual differences
- Each person serves as their own control
- Reduces error variance
- Often needs fewer participants
:::

## The Paired t-test Formula

<!-- TODO: Add formula -->

$$t = \frac{\bar{D}}{SE_{\bar{D}}}$$

::: {.fragment}
Where $\bar{D}$ is the mean of the difference scores, and $SE_{\bar{D}}$ is the standard error of those differences.
:::

## Assumptions of Paired t-test

<!-- TODO: List assumptions -->

::: {.incremental}
- Pairs are independent of each other
- Differences are continuous
- Differences are normally distributed (or large n)
:::

## Example: Paired t-test

<!-- TODO: Add worked example -->

```{r}
#| eval: false
# TODO: Add R code for paired t-test
```

::: {.exercise-box}
**Research question:** Do participants' mood ratings improve after watching a comedy video?
:::

# Choosing the Right Test {background-color="#275882"}

## Decision Flowchart

<!-- TODO: Add decision guide -->

::: {.incremental}
1. **How many groups/conditions?** (2 for t-tests)
2. **Same or different participants?**
   - Different groups -> Independent t-test
   - Same participants -> Paired t-test
3. **Check assumptions**
4. **Report results properly**
:::

## Reporting t-test Results

<!-- TODO: Add APA format guidance -->

::: {.example-box}
**APA format:**

"An independent samples t-test revealed a significant difference in reaction times between the caffeine group (M = 245, SD = 32) and placebo group (M = 278, SD = 41), t(58) = 3.52, p = .001, d = 0.91."
:::

## What to Report

<!-- TODO: List components -->

::: {.incremental}
- Descriptive statistics (M, SD) for each group
- t-value and degrees of freedom: t(df)
- p-value (exact, unless p < .001)
- Effect size (Cohen's d)
- Confidence interval (optional but recommended)
:::

# Summary {background-color="#275882"}

## Key Takeaways

::: {.incremental}
- **NHST** tests whether data are consistent with "no effect"
- **p-values** indicate probability of data given H0 is true
- **Independent t-tests** compare two separate groups
- **Paired t-tests** compare two measurements from the same participants
- Always report **effect sizes** alongside p-values
:::

## Next Week

<!-- TODO: Update with next week's topic -->

**Week 15: Correlation**

- Relationships between variables
- Visualising with scatterplots
- Pearson's r and Spearman's rho
- Why correlation does not imply causation

## Questions?

::: {.r-fit-text}
Questions?
:::

See you in the lab!
