---
title: "t-tests in R"
subtitle: "PS50008A Week 14 Lab Activity"
author: "Gordon Wright"
date: "2026-02-04"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(webexercises)
library(tidyverse)
library(knitr)

# Load DataLab data
memory <- read_csv("../../_planning/DATALAB_MATERIALS/data/processed/memory_clean.csv", show_col_types = FALSE)
merged <- read_csv("../../_planning/DATALAB_MATERIALS/data/processed/foundations_merged.csv", show_col_types = FALSE)

navy <- "#275882"
orange <- "#EA8439"
theme_set(theme_minimal(base_size = 14))
```

## Overview

::: {.callout-important}
## Focus for Today
This lab introduces the **t-test** - the workhorse of psychological research for comparing means:

- Running independent samples t-tests (between-subjects)
- Running paired samples t-tests (within-subjects)
- Interpreting R output from t.test()
- Choosing the right test for your design
:::

---

## Activity 1: Independent t-test (Lab A vs Lab B Memory) (20 min)

### When to Use an Independent t-test

An **independent samples t-test** compares means from **two different groups** of participants.

```{r}
#| label: lab-data-prep
memory_labs <- memory |>
  filter(!is.na(Q11) & !is.na(mem_total_1)) |>
  mutate(Lab = factor(Q11, levels = c(1, 2), labels = c("Lab A (Abstract)", "Lab B (Concrete)")))

lab_stats <- memory_labs |>
  group_by(Lab) |>
  summarise(
    n = n(),
    Mean = mean(mem_total_1, na.rm = TRUE),
    SD = sd(mem_total_1, na.rm = TRUE)
  )
```

```{r}
#| label: lab-comparison-plot
#| fig-height: 4
#| fig-align: center
ggplot(memory_labs, aes(x = Lab, y = mem_total_1, fill = Lab)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.5) +
  geom_jitter(aes(colour = Lab), width = 0.15, size = 3, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, colour = "black") +
  scale_fill_manual(values = c(navy, orange)) +
  scale_colour_manual(values = c(navy, orange)) +
  labs(x = NULL, y = "Words Recalled (out of 10)",
       title = "Memory Performance: Lab A vs Lab B",
       subtitle = "Black diamonds = group means") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

### Conceptual Questions

**1. This comparison (Lab A vs Lab B) uses an independent t-test because:**

`r mcq(c("The same participants are in both groups", answer = "Different participants are in each group", "We are comparing more than two groups", "There is no control group"))`

**2. What are the two groups being compared?**

`r mcq(c(answer = "Lab A (abstract words) vs Lab B (concrete words)", "Before vs After", "High vs Low performers", "Morning vs Afternoon sessions"))`

**3. What is the dependent variable in this analysis?**

`r mcq(c("Word type (abstract/concrete)", "Lab group", answer = "Number of words recalled", "Participant ID"))`

### Understanding the R Output

Here's what R output from `t.test()` looks like:

```
	Welch Two Sample t-test

data:  mem_total_1 by Lab
t = -2.45, df = 38.2, p-value = 0.019
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -2.15  -0.21
sample estimates:
mean in group Lab A  mean in group Lab B
               5.40                6.58
```

**4. Looking at the output above, what is the t-statistic?**

`r mcq(c("38.2", "0.019", answer = "-2.45", "-2.15"))`

**5. What is the p-value?**

`r mcq(c("-2.45", "38.2", answer = "0.019", "6.58"))`

**6. At alpha = 0.05, would we reject the null hypothesis?**

`r mcq(c(answer = "Yes, because p < 0.05", "No, because p > 0.05", "Can't tell from this output", "Yes, because t is negative"))`

**7. The 95% CI for the difference is [-2.15, -0.21]. What does this tell us?**

`r mcq(c(answer = "The CI doesn't include 0, supporting a real difference", "The CI includes 0, so there's no difference", "Lab B scored lower than Lab A", "The difference is not significant"))`

True or False: A negative t-value means the first group has a lower mean than the second. `r torf(TRUE)`

---

## Activity 2: Paired t-test (Pre vs Post Mood) (20 min)

### When to Use a Paired t-test

A **paired samples t-test** compares means from the **same participants** measured twice.

```{r}
#| label: mood-data-prep
mood_data <- merged |>
  select(participant_id, mood_pre_1, post_mood_post_1) |>
  filter(!is.na(mood_pre_1) & !is.na(post_mood_post_1)) |>
  mutate(change = post_mood_post_1 - mood_pre_1)

mood_long <- mood_data |>
  pivot_longer(cols = c(mood_pre_1, post_mood_post_1),
               names_to = "Time", values_to = "Mood") |>
  mutate(Time = factor(Time, levels = c("mood_pre_1", "post_mood_post_1"),
                       labels = c("Before", "After")))
```

```{r}
#| label: mood-comparison-plot
#| fig-height: 4
#| fig-align: center
ggplot(mood_long, aes(x = Time, y = Mood, fill = Time)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.5) +
  geom_line(data = mood_data |>
              pivot_longer(cols = c(mood_pre_1, post_mood_post_1),
                          names_to = "Time", values_to = "Mood") |>
              mutate(Time = factor(Time, levels = c("mood_pre_1", "post_mood_post_1"),
                                  labels = c("Before", "After"))),
            aes(group = participant_id), alpha = 0.3, colour = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, colour = "black") +
  scale_fill_manual(values = c(navy, orange)) +
  labs(x = NULL, y = "Mood Rating (1-10)",
       title = "Mood Before vs After DataLab",
       subtitle = "Lines connect the same person") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

### Conceptual Questions

**1. This comparison uses a paired t-test because:**

`r mcq(c(answer = "The same participants are measured at both time points", "Different participants are in each group", "We are comparing more than two groups", "The data is paired with another dataset"))`

**2. What is the advantage of a paired design over an independent design?**

`r mcq(c("Easier to recruit participants", "Faster to run", answer = "Each person is their own control, reducing individual differences", "No assumptions required"))`

**3. In R, how do we specify a paired t-test?**

`r mcq(c("t.test(x, y, type = 'paired')", answer = "t.test(x, y, paired = TRUE)", "paired.t.test(x, y)", "t.test.paired(x, y)"))`

### Understanding Paired Output

```
	Paired t-test

data:  mood_pre and mood_post
t = -3.12, df = 42, p-value = 0.003
alternative hypothesis: true mean difference is not equal to 0
95 percent confidence interval:
 -1.52  -0.32
sample estimates:
mean difference
          -0.92
```

**4. The mean difference of -0.92 means:**

`r mcq(c("Pre scores were 0.92 higher than post", answer = "Post scores were 0.92 higher than pre", "The groups are 0.92 apart on average", "92% of people improved"))`

**5. Is this difference statistically significant at alpha = 0.05?**

`r mcq(c(answer = "Yes, p = 0.003 < 0.05", "No, p = 0.003 > 0.05", "Can't tell", "Only marginally significant"))`

**6. The degrees of freedom (df) for a paired t-test equals:**

`r mcq(c("n", answer = "n - 1 (number of pairs minus 1)", "n - 2", "n1 + n2 - 2"))`

True or False: In a paired t-test, we analyse the difference scores, not the raw scores. `r torf(TRUE)`

---

## Activity 3: Interpreting Output (20 min)

### Reading and Reporting t-test Results

Psychologists report t-test results in a specific format. Let's practice!

**Standard format:** t(df) = value, p = value

### Practice Examples

**Example 1:**
```
t = 2.34, df = 28, p-value = 0.027
```

**1. How would you report this result?**

`r mcq(c("t = 2.34, p = 0.027", answer = "t(28) = 2.34, p = .027", "t(2.34) = 28, p = 0.027", "df = 28, t = 2.34"))`

**2. Is this result significant at alpha = 0.05?**

`r mcq(c(answer = "Yes", "No", "Can't tell", "Marginally significant"))`

**Example 2:**
```
t = 1.52, df = 45, p-value = 0.136
```

**3. Is this result significant at alpha = 0.05?**

`r mcq(c("Yes", answer = "No", "Can't tell", "Marginally significant"))`

**4. What should you conclude?**

`r mcq(c("The groups are definitely the same", answer = "We cannot reject the null hypothesis - insufficient evidence of a difference", "The experiment failed", "We need to use a different test"))`

### Assumptions of the t-test

**5. Which assumption is about the shape of the data distribution?**

`r mcq(c(answer = "Normality assumption", "Independence assumption", "Homogeneity of variance", "Randomness assumption"))`

**6. Welch's t-test is preferred when:**

`r mcq(c("Sample sizes are equal", "Data is normally distributed", answer = "Variances may be unequal between groups", "We want a paired test"))`

**7. Independence means:**

`r mcq(c(answer = "One person's score doesn't influence another's", "The groups are completely separate", "Variables are uncorrelated", "The test is unbiased"))`

True or False: t-tests are robust to mild violations of normality, especially with larger samples. `r torf(TRUE)`

True or False: If p = 0.06, the result is "almost significant" and we should report it as meaningful. `r torf(FALSE)`

---

## Activity 4: Choosing the Right Test (20 min)

### Decision Tree Practice

For each scenario, identify whether to use an **independent** or **paired** t-test.

::: {.webex-check .webex-box}

### Scenario 1
A researcher measures anxiety levels in 30 participants before and after a relaxation intervention.

**Which t-test?** `r mcq(c("Independent t-test", answer = "Paired t-test"))`

**Why?** `r mcq(c("Different people in each condition", answer = "Same people measured twice"))`
:::

::: {.webex-check .webex-box}

### Scenario 2
Comparing exam scores between students who used Method A vs Method B to study.

**Which t-test?** `r mcq(c(answer = "Independent t-test", "Paired t-test"))`

**Why?** `r mcq(c(answer = "Different people in each condition", "Same people measured twice"))`
:::

::: {.webex-check .webex-box}

### Scenario 3
Testing whether people rate faces as more attractive when the same faces are presented smiling vs neutral.

**Which t-test?** `r mcq(c("Independent t-test", answer = "Paired t-test"))`

**Why?** `r mcq(c("Different people rate different faces", answer = "Same people rate both versions of faces"))`
:::

::: {.webex-check .webex-box}

### Scenario 4
Comparing reaction times between left-handed and right-handed participants.

**Which t-test?** `r mcq(c(answer = "Independent t-test", "Paired t-test"))`

**Why?** `r mcq(c(answer = "Handedness is a between-subjects variable", "People switch hands during the experiment"))`
:::

::: {.webex-check .webex-box}

### Scenario 5
Each participant completes a task under quiet conditions and under noisy conditions.

**Which t-test?** `r mcq(c("Independent t-test", answer = "Paired t-test"))`

**Why?** `r mcq(c("Different people in each condition", answer = "Same people tested in both conditions"))`
:::

### The Key Question

**To decide which t-test to use, ask yourself:**

`r mcq(c("How many variables are there?", "Is the data normally distributed?", answer = "Are the SAME or DIFFERENT people in each condition?", "What is the sample size?"))`

---

## Summary

::: {.callout-important}
## Key Takeaways

**Independent t-test:**
- Compares means from TWO DIFFERENT groups
- Use for between-subjects designs
- df = n1 + n2 - 2

**Paired t-test:**
- Compares means from the SAME people measured twice
- Use for within-subjects designs
- df = n - 1 (number of pairs minus 1)
- More powerful (controls for individual differences)

**Reporting:**
- Format: t(df) = value, p = value
- Include effect size (Cohen's d) when possible
- Report 95% CI for the difference

**Choosing:**
- Ask: "Same people or different people?"
:::

---

## Next Steps

1. Complete the [Week 14 Self-Assessment](self-assessment.qmd)
2. Review the [Lecture Reading](reading.qmd)
3. Prepare for Week 15: Correlation in R
