---
title: "Week 13 Self-Assessment"
subtitle: "Correlation & Relationships"
---

```{r, echo=FALSE}
library(webexercises)
```

## Instructions

This quiz deepens your understanding of correlational relationships. Week 12 introduced correlation as a design type — this week you'll explore how to interpret correlations more thoroughly.

---

## Part 1: Correlation Basics

### Question 1: Direction and Strength

A correlation coefficient of **r = -.75** indicates:

Direction: `r mcq(c("Positive", answer = "Negative", "No relationship", "Cannot tell"))`

Strength: `r mcq(c("Weak", "Moderate", answer = "Strong", "Perfect"))`

### Question 2: Interpreting r Values

Match each correlation coefficient to its interpretation:

**r = .15** → `r mcq(c(answer = "Weak positive", "Moderate positive", "Strong positive", "No relationship"))`

**r = -.62** → `r mcq(c("Weak negative", answer = "Moderate to strong negative", "Strong positive", "Perfect negative"))`

**r = .03** → `r mcq(c("Weak positive", "Moderate positive", answer = "Essentially no relationship", "Strong negative"))`

**r = .89** → `r mcq(c("Weak positive", "Moderate positive", answer = "Strong positive", "Perfect positive"))`

### Question 3: Range of r

The correlation coefficient (r) can range from:

`r mcq(c("0 to 1", "0 to 100", answer = "-1 to +1", "-100 to +100"))`

---

## Part 2: Correlation ≠ Causation

### Question 4: The Third Variable Problem

A researcher finds that ice cream sales and drowning rates are positively correlated. The most likely explanation is:

`r mcq(c("Ice cream causes drowning", "Drowning causes people to buy ice cream", answer = "A third variable (warm weather) affects both", "This is a coincidence with no explanation"))`

### Question 5: Identifying Third Variables

For each correlation, identify the most likely third variable:

**Shoe size and reading ability are positively correlated in children.**

`r mcq(c("Reading makes feet grow", "Big feet help with reading", answer = "Age (older children have bigger feet AND read better)", "This is coincidence"))`

**Number of firefighters at a scene and damage caused are positively correlated.**

`r mcq(c("Firefighters cause damage", "Damage attracts firefighters", answer = "Fire severity (bigger fires need more firefighters AND cause more damage)", "This is coincidence"))`

### Question 6: Why Correlation Cannot Establish Causation

```{r}
opts_cause <- c(
  "Correlations are always weak",
  answer = "Correlations show association but cannot rule out third variables or determine direction of effect",
  "Correlations only work with small samples",
  "Correlations are only for categorical variables"
)
```

Why can't we conclude causation from a correlational study?

`r longmcq(opts_cause)`

---

## Part 3: Scatterplot Interpretation

### Question 7: Reading Scatterplots

A scatterplot shows points forming a pattern from lower-left to upper-right. This indicates:

`r mcq(c(answer = "Positive correlation", "Negative correlation", "No correlation", "Curvilinear relationship"))`

A scatterplot shows points forming a pattern from upper-left to lower-right. This indicates:

`r mcq(c("Positive correlation", answer = "Negative correlation", "No correlation", "Curvilinear relationship"))`

A scatterplot shows points scattered randomly with no clear pattern. This indicates:

`r mcq(c("Positive correlation", "Negative correlation", answer = "No correlation (or very weak)", "Perfect correlation"))`

### Question 8: Outliers

What effect do outliers typically have on Pearson's r?

`r mcq(c("No effect", "Always increase r", "Always decrease r", answer = "Can substantially change r in either direction"))`

---

## Part 4: Correlation Tests

### Question 9: Pearson vs Spearman

Pearson's r is appropriate when:

`r mcq(c(answer = "Both variables are continuous and normally distributed", "One variable is categorical", "Data are ranks or ordinal", "Sample size is very small"))`

Spearman's rho (ρ) is appropriate when:

`r mcq(c("Both variables must be normally distributed", "One variable is categorical", answer = "Data are ordinal/ranked OR continuous data violates normality", "Sample size is very large"))`

### Question 10: DataLab Example

In our DataLab, we might correlate **confidence ratings** with **throwing accuracy**.

What type of relationship might we expect?

`r mcq(c(answer = "Positive (higher confidence = better accuracy)", "Negative (higher confidence = worse accuracy)", "No relationship expected", "Cannot predict"))`

What test would we use if both variables are roughly normally distributed?

`r mcq(c(answer = "Pearson's r", "Spearman's rho", "Independent t-test", "Paired t-test"))`

---

## Part 5: APA Reporting

### Question 11: Reporting Correlations

The correct APA format for reporting a significant positive correlation is:

`r mcq(c("r = .45, significant", "r(48) = .45, p = .001", answer = "r(48) = .45, p = .001 (or r = .45, p = .001)", "correlation = 0.45, sig = 0.001"))`

### Question 12: Fill in the Blank

When r = .49, the coefficient of determination (r²) = `r fitb(c(".24", "0.24", ".2401", "24%"), ignore_case = TRUE)`

This means that `r fitb(c("24", "24%"), ignore_case = TRUE)` percent of the variance in one variable is explained by the other.

---

## Part 6: MCQ Exam Practice

::: {.webex-check .webex-box}

### Investigation 1

A researcher measures hours of sleep and exam performance in 50 students to see if they are related.

1. **Design**: `r mcq(c("Between-subjects", "Within-subjects", "Matched pairs", answer = "Correlational"))`

2. **IV**: `r mcq(c("Hours of sleep", "Exam performance", "Number of students", answer = "No IV (correlational design)"))`

3. **Variables measured**: `r mcq(c("Just sleep hours", "Just exam scores", answer = "Both sleep hours and exam scores", "Neither"))`

4. **Appropriate test**: `r mcq(c("Independent t-test", "Paired t-test", answer = "Pearson's r (or Spearman's rho)", "Chi-square"))`

5. **If r = .35, this is**: `r mcq(c("A strong relationship", answer = "A moderate relationship", "No relationship", "A perfect relationship"))`

:::

::: {.webex-check .webex-box}

### Investigation 2

A researcher finds that self-esteem scores correlate positively with academic grades (r = .42) in university students.

1. **Can we conclude high self-esteem CAUSES better grades?** `r torf(FALSE)`

2. **Name a possible third variable**: `r fitb(c("motivation", "intelligence", "effort", "socioeconomic status", "family support", "IQ"), ignore_case = TRUE)`

3. **What percentage of variance is shared?** `r fitb(c("17.64", "18", "17.6", "about 18"), ignore_case = TRUE)`%

:::

---

## Summary Checklist

Before moving on, make sure you can:

- [ ] Interpret correlation coefficients (direction and strength)
- [ ] Explain why correlation does not imply causation
- [ ] Identify potential third variables in correlational findings
- [ ] Read and interpret scatterplots
- [ ] Choose between Pearson's r and Spearman's rho
- [ ] Calculate and interpret r² (coefficient of determination)
- [ ] Report correlations in APA format

::: {.callout-important}
## Research Plan Connection
If your Research Plan uses a correlational design, you need to clearly state that you CANNOT make causal claims. This is assessed in the MCQ exam.
:::
