---
title: "Week 19 Self-Assessment"
subtitle: "Results & Lab Report Writing"
---

```{r, echo=FALSE}
library(webexercises)
```

## Instructions

This quiz focuses on writing the Results section of your lab report and understanding how to present statistical findings in APA format.

---

## Part 1: Results Section Structure

### Question 1: What Belongs in Results?

Which of the following belongs in the Results section?

Descriptive statistics (M, SD) `r torf(TRUE)`

Statistical test results `r torf(TRUE)`

Interpretation of what the results mean for theory `r torf(FALSE)`

Effect sizes `r torf(TRUE)`

Literature review `r torf(FALSE)`

### Question 2: Results vs Discussion

The Results section should:

`r mcq(c("Explain why you found what you found", answer = "Report what you found without extensive interpretation", "Include your hypothesis from the Introduction", "Describe your participants"))`

The Discussion section should:

`r mcq(c("Report statistical values", answer = "Interpret results and relate them to theory and previous research", "Describe the procedure", "List materials used"))`

---

## Part 2: Reporting Descriptive Statistics

### Question 3: APA Format for Descriptives

The correct APA format for reporting a mean of 45.30 and standard deviation of 8.25 is:

`r mcq(c("mean=45.30, std=8.25", "Average: 45.30 (8.25)", answer = "M = 45.30, SD = 8.25", "45.30 +/- 8.25"))`

### Question 4: Decimal Places

In APA format, how many decimal places are typically reported for means and standard deviations?

`r mcq(c("0", "1", answer = "2", "As many as possible"))`

### Question 5: Descriptives Table

A well-formatted descriptive statistics table should include:

`r mcq(c("Only means", "Only standard deviations", answer = "Group names, n, M, and SD for each condition", "Raw data for all participants"))`

---

## Part 3: Reporting t-Tests

### Question 6: Independent t-Test Reporting

The correct APA format for an independent t-test is:

`r mcq(c("t = 2.45, significant", answer = "t(48) = 2.45, p = .018, d = 0.71", "t-test showed significant results", "Independent samples showed difference"))`

What does the number in parentheses represent?

`r mcq(c("Sample size", answer = "Degrees of freedom", "Effect size", "p-value"))`

### Question 7: Paired t-Test Reporting

A paired t-test with 25 participants would have degrees of freedom of:

`r fitb(c("24"), ignore_case = TRUE)`

The correct reporting format is:

`r mcq(c("t(25) = ...", answer = "t(24) = ...", "t(50) = ...", "t = ... (n=25)"))`

### Question 8: Reporting Direction

When t(28) = -2.89, p = .007, the negative t-value indicates:

`r mcq(c("An error in analysis", "The result is not significant", answer = "The second mean was higher than the first (or first lower than second)", "Effect size is negative"))`

---

## Part 4: Reporting Correlations

### Question 9: Correlation Format

The correct APA format for reporting a correlation is:

`r mcq(c("r = .45, significant", answer = "r(48) = .45, p = .001", "Correlation was found (r = .45)", "Pearson's correlation = 0.45, p-value = 0.001"))`

### Question 10: Degrees of Freedom for Correlation

For a correlation with 50 participants, df =

`r fitb(c("48"), ignore_case = TRUE)`

(Hint: df = N - 2 for correlations)

### Question 11: Direction and Strength

How would you describe r = -.62 in words?

`r mcq(c("A weak negative correlation", answer = "A moderate-to-strong negative correlation", "A strong positive correlation", "No correlation"))`

---

## Part 5: Putting It All Together

### Question 12: Complete Results Paragraph

Which is the best written Results paragraph?

```{r}
opts_results <- c(
  "The results were significant and showed what we expected.",
  answer = "An independent samples t-test revealed that participants in the music condition (M = 72.40, SD = 8.50) recalled significantly more words than those in the silence condition (M = 65.20, SD = 9.10), t(48) = 2.89, p = .006, d = 0.82.",
  "We ran a t-test and it worked. The music group was better.",
  "The hypothesis was supported because p < .05."
)
```

`r longmcq(opts_results)`

### Question 13: Results Writing Practice

Fill in the blanks for this Results paragraph:

"A `r fitb(c("paired", "paired samples"), ignore_case = TRUE)` t-test revealed that anxiety scores were significantly `r fitb(c("lower", "reduced", "decreased"), ignore_case = TRUE)` after the intervention (M = 32.50, SD = 7.20) compared to before (M = 45.80, SD = 8.40), t(29) = -4.56, p < .001, d = `r fitb(c("0.83", ".83", "0.8"), ignore_case = TRUE)`."

---

## Part 6: Common Mistakes

### Question 14: Avoid These Errors

Which of these is INCORRECT for a Results section?

`r mcq(c("Reporting exact p-values (e.g., p = .023)", answer = "Writing 'The results prove our hypothesis'", "Including effect sizes", "Presenting descriptive statistics before inferential statistics"))`

### Question 15: p-Value Reporting

Which is the correct way to report a very small p-value?

`r mcq(c("p = .000", answer = "p < .001", "p = 0", "p is significant"))`

### Question 16: Leading Zeros

In APA format, which is correct?

`r mcq(c("p = 0.032", answer = "p = .032", "p = .03", "P = 0.032"))`

Why no leading zero?

`r mcq(c("It saves space", answer = "Values that cannot exceed 1 (like p and r) don't need a leading zero", "It's arbitrary convention", "Both are acceptable"))`

---

## Part 7: Practice Scenarios

::: {.webex-check .webex-box}

### Scenario 1

You conducted an independent t-test comparing test scores between a morning class (n=25, M=78.4, SD=10.2) and afternoon class (n=25, M=72.1, SD=11.5). Results: t(48) = 2.05, p = .046, d = 0.58.

1. **Is this significant at α = .05?** `r torf(TRUE)`

2. **Effect size is:** `r mcq(c("Small", answer = "Medium", "Large", "Negligible"))`

3. **Degrees of freedom are:** `r fitb(c("48"), ignore_case = TRUE)`

4. **The morning class scored:** `r mcq(c(answer = "Higher", "Lower", "The same", "Cannot tell"))`

:::

::: {.webex-check .webex-box}

### Scenario 2

You conducted a correlation between study hours and exam grades (n=60). Results: r(58) = .42, p < .001.

1. **Degrees of freedom are:** `r fitb(c("58"), ignore_case = TRUE)`

2. **The correlation is:** `r mcq(c("Negative", answer = "Positive", "Zero", "Perfect"))`

3. **Strength is:** `r mcq(c("Weak", answer = "Moderate", "Strong", "Very weak"))`

4. **Shared variance (r²) is approximately:** `r fitb(c("17.64", "18", "17.6"), ignore_case = TRUE)`%

:::

---

## Summary Checklist

Before writing your Results section, verify:

- [ ] Descriptive statistics reported first (M, SD for each condition)
- [ ] Correct APA format used for all statistics
- [ ] Degrees of freedom reported in parentheses
- [ ] Exact p-values reported (or p < .001 for very small values)
- [ ] Effect sizes included
- [ ] No leading zeros for values that can't exceed 1
- [ ] Results described without extensive interpretation
- [ ] Statistical notation italicised (t, p, r, d, M, SD)

::: {.callout-important}
## Lab Report Connection
Your Results section should follow these guidelines exactly. Review the assessment brief for specific requirements and word limits.
:::
