---
title: "Regression"
subtitle: "PS50008A Week 16 | Lecture"
author: "Gordon Wright"
date: "2026-02-25"
format:
  revealjs:
    theme: [default, ../../assets/slides.scss]
    slide-number: c/t
    progress: true
    transition: fade
    width: 1050
    height: 700
    center: false
    center-title-slide: true
execute:
  echo: false
  warning: false
  message: false
---

## Learning Outcomes

By the end of this lecture you will be able to:

- Explain how regression extends correlation to prediction
- Interpret the line of best fit and its equation
- Understand R^2 and explained variance
- Use a regression equation to make predictions

# From Correlation to Prediction {background-color="#275882"}

## Beyond Describing Relationships

<!-- TODO: Introduce prediction concept -->

::: {.incremental}
- **Correlation** tells us variables are related
- **Regression** allows us to predict one from the other
- Moving from description to prediction
:::

::: {.fragment}
::: {.example-box}
**Correlation:** Study hours and exam scores are positively related (r = .72)

**Regression:** If a student studies for X hours, we predict their exam score will be Y.
:::
:::

## What is Regression?

<!-- TODO: Define regression -->

::: {.definition-box}
**Simple Linear Regression**: A statistical method that models the relationship between a predictor variable (X) and an outcome variable (Y) using a straight line.
:::

## Terminology

<!-- TODO: Define key terms -->

:::: {.columns}

::: {.column width="50%"}
**Predictor Variable (X)**

- Also called: independent variable, explanatory variable
- Used to predict Y
- Goes on X-axis
:::

::: {.column width="50%"}
**Outcome Variable (Y)**

- Also called: dependent variable, response variable
- What we're predicting
- Goes on Y-axis
:::

::::

## The Regression Question

<!-- TODO: Frame the core question -->

::: {.fragment}
Given a value of X, what is our best prediction of Y?
:::

::: {.fragment}
::: {.example-box}
"If a student studies for 10 hours, what exam score do we predict?"
:::
:::

## Why Linear?

<!-- TODO: Explain linearity assumption -->

::: {.incremental}
- We assume a straight-line relationship
- Simplest model
- Often a good approximation
- Can extend to non-linear later
:::

# Line of Best Fit {background-color="#275882"}

## The Regression Line

<!-- TODO: Introduce regression line concept -->

::: {.definition-box}
**Line of Best Fit**: The straight line that minimises the total distance between the line and all data points.
:::

::: {.fragment}
<!-- TODO: Add scatterplot with regression line -->
:::

## The Regression Equation

<!-- TODO: Present the equation -->

$$\hat{Y} = b_0 + b_1X$$

::: {.fragment}
Where:

- $\hat{Y}$ (Y-hat) = predicted value of Y
- $b_0$ = intercept (Y when X = 0)
- $b_1$ = slope (change in Y per unit X)
- $X$ = value of predictor variable
:::

## Understanding the Intercept (b0)

<!-- TODO: Explain intercept -->

::: {.definition-box}
**Intercept (b0)**: The predicted value of Y when X equals zero. Where the line crosses the Y-axis.
:::

::: {.fragment}
::: {.callout-note}
The intercept may not always be meaningful (e.g., "exam score when study hours = 0").
:::
:::

## Understanding the Slope (b1)

<!-- TODO: Explain slope -->

::: {.definition-box}
**Slope (b1)**: The change in Y for each one-unit increase in X. The steepness and direction of the line.
:::

::: {.fragment}
::: {.example-box}
If b1 = 5.2:

"For each additional hour of study, exam scores increase by 5.2 points on average."
:::
:::

## How the Line is Fitted

<!-- TODO: Introduce least squares -->

::: {.definition-box}
**Ordinary Least Squares (OLS)**: Finds the line that minimises the sum of squared residuals (vertical distances between points and line).
:::

::: {.fragment}
<!-- TODO: Add visual showing residuals -->
:::

## Residuals

<!-- TODO: Define residuals -->

::: {.definition-box}
**Residual**: The difference between an observed Y value and the predicted Y value. The "leftover" variation.
:::

$$Residual = Y - \hat{Y}$$

::: {.fragment}
- Positive residual: actual > predicted
- Negative residual: actual < predicted
:::

## Visualising Residuals

<!-- TODO: Add residual plot -->

```{r}
#| eval: false
# TODO: Add R code showing scatterplot with
# vertical lines to regression line (residuals)
```

# R^2 and Explained Variance {background-color="#275882"}

## How Good is Our Model?

<!-- TODO: Introduce model fit concept -->

::: {.incremental}
- We have a line, but how well does it fit?
- How much of Y's variation does X explain?
- We need a measure of model quality
:::

## Total Variance in Y

<!-- TODO: Explain variance decomposition -->

::: {.fragment}
The total variance in Y can be split into:

1. **Explained variance**: Variation predicted by X
2. **Unexplained variance**: Leftover (residual) variation
:::

$$Total = Explained + Unexplained$$

## Coefficient of Determination (R^2)

<!-- TODO: Define R-squared -->

::: {.definition-box}
**R^2 (R-squared)**: The proportion of variance in Y that is explained by X. Ranges from 0 to 1.
:::

$$R^2 = \frac{Explained\ Variance}{Total\ Variance}$$

## Interpreting R^2

<!-- TODO: Add interpretation examples -->

| R^2 value | Interpretation |
|-----------|----------------|
| 0.00 | X explains none of Y's variance |
| 0.25 | X explains 25% of Y's variance |
| 0.50 | X explains half of Y's variance |
| 1.00 | X explains all of Y's variance (perfect) |

## R^2 in Simple Regression

<!-- TODO: Connect to correlation -->

::: {.callout-tip}
## Key insight
In simple linear regression (one predictor), R^2 equals the square of the correlation coefficient.

If r = 0.72, then R^2 = 0.52
:::

## What R^2 Doesn't Tell You

<!-- TODO: Limitations of R^2 -->

::: {.incremental}
- Whether the relationship is causal
- Whether the model is "correct"
- Whether predictions will be accurate for new data
- The importance of the predictor
:::

## Residual Analysis

<!-- TODO: Checking assumptions -->

::: {.example-box}
**Checking model assumptions via residuals:**

- Should be randomly scattered (no patterns)
- Should be roughly normally distributed
- Should have constant variance (homoscedasticity)
:::

# Making Predictions {background-color="#275882"}

## Using the Regression Equation

<!-- TODO: Prediction walkthrough -->

::: {.example-box}
**Given:** $\hat{Y} = 30 + 5X$

**Question:** Predict exam score for someone who studied 8 hours.

**Solution:** $\hat{Y} = 30 + 5(8) = 30 + 40 = 70$

**Prediction:** 70 points
:::

## Worked Example

<!-- TODO: Add complete worked example -->

```{r}
#| eval: false
# TODO: Add R code showing:
# 1. Scatterplot of data
# 2. Regression line
# 3. Equation extraction
# 4. Making a prediction
```

## Interpolation vs Extrapolation

<!-- TODO: Discuss prediction limits -->

:::: {.columns}

::: {.column width="50%"}
**Interpolation**

- Predicting within data range
- More reliable
- Model tested here
:::

::: {.column width="50%"}
**Extrapolation**

- Predicting beyond data range
- Risky!
- Relationship may change
:::

::::

## The Dangers of Extrapolation

<!-- TODO: Add cautionary examples -->

::: {.callout-warning}
## Warning
Never extrapolate far beyond your data. The linear relationship may not hold.
:::

::: {.fragment}
::: {.example-box}
If study hours in your data range from 2-15 hours, don't predict scores for someone who studied 100 hours!
:::
:::

## Prediction Intervals

<!-- TODO: Introduce uncertainty in predictions -->

::: {.definition-box}
**Prediction Interval**: A range of values within which a new observation is likely to fall, accounting for uncertainty in the prediction.
:::

::: {.fragment}
Wider than confidence intervals for the mean.
:::

## Reporting Regression Results

<!-- TODO: APA format guidance -->

::: {.example-box}
**APA format:**

"A simple linear regression was conducted to predict exam scores from study hours. Study hours significantly predicted exam scores, b = 5.2, t(48) = 7.89, p < .001. Study hours explained 52% of the variance in exam scores, R^2 = .52, F(1, 48) = 62.25, p < .001."
:::

## What to Report

<!-- TODO: Reporting checklist -->

::: {.incremental}
- The regression equation (with coefficients)
- Statistical significance of predictors
- R^2 and its interpretation
- F-test for overall model
- Assumptions checked
:::

# Summary {background-color="#275882"}

## Key Takeaways

::: {.incremental}
- **Regression** extends correlation to prediction
- **Line of best fit** minimises squared residuals
- **Equation:** $\hat{Y} = b_0 + b_1X$ (intercept + slope * X)
- **R^2** tells us proportion of variance explained
- **Predictions** should stay within the data range
:::

## Looking Ahead

<!-- TODO: Module wrap-up preview -->

::: {.fragment}
You now have the core statistical toolkit:

- Describing data (Week 12)
- Estimation & effect sizes (Week 13)
- Hypothesis testing & t-tests (Week 14)
- Correlation (Week 15)
- Regression (Week 16)
:::

## Applying Your Knowledge

<!-- TODO: Connection to assessments -->

::: {.callout-tip}
## Research Plan Reminder
Your research plan is due soon! These statistical concepts will help you:

- Choose appropriate analyses
- Justify your analytical approach
- Interpret your future results
:::

## Questions?

::: {.r-fit-text}
Questions?
:::

See you in the lab!
