---
title: "Week 11 Self-Assessment"
subtitle: "Introduction to Research Methods & DataLab"
---

```{r, echo=FALSE}
library(webexercises)
```

## Instructions

Test your understanding of this week's material. Answer each question to check your learning.

---

## Conceptual Questions

### Question 1: Why Research Methods?

Why is "common sense" unreliable for understanding human behaviour?

`r mcq(c("It's always wrong", answer = "It suffers from biases like hindsight and confirmation bias", "Scientists don't like common sense", "It only works for simple behaviours"))`

`r hide("Show explanation")`
Common sense is unreliable because of:

- **Hindsight bias** — events seem obvious after they occur
- **Confirmation bias** — we notice supporting evidence and ignore contradicting evidence
- **Anecdotal evidence** — individual cases can mislead
- Common sense can justify contradictory conclusions (e.g., "opposites attract" vs "birds of a feather flock together")
`r unhide()`

### Question 2: Hallmarks of Science

```{r}
opts_hallmarks <- c(
  "Popular, funded, published, and cited",
  answer = "Empirical, systematic, falsifiable, and transparent",
  "Theoretical, mathematical, complex, and reviewed",
  "Objective, simple, fast, and certain"
)
```

What are the four hallmarks of scientific claims?

`r longmcq(opts_hallmarks)`

### Question 3: Variable Types

A **continuous** variable can take any value within a range. `r torf(TRUE)`

A participant's **experimental condition** (A or B) is an example of a categorical variable. `r torf(TRUE)`

**Reaction time in milliseconds** is an example of a categorical variable. `r torf(FALSE)`

### Question 4: Replication

What does it mean for a finding to be "replicable"?

`r mcq(c("The finding was published in a journal", "The researcher repeated the study themselves", answer = "Other researchers get the same results following the same procedures", "The finding agrees with common sense"))`

---

## Practical Questions

### Question 5: Measurement Error

In the reaction time task (ruler drop), which of the following could cause measurement error?

Inconsistent ruler height `r torf(TRUE)`

Participant knowing when the drop will happen `r torf(TRUE)`

Using a standardised ruler `r torf(FALSE)`

### Question 6: Being a Participant and Tester

Why did you act as both participant AND tester in DataLab?

`r mcq(c("To save time", "Because we didn't have enough researchers", answer = "To understand both perspectives and the importance of standardised procedures", "To collect more data"))`

### Question 7: Signal Detection

In the staring task, what is a "hit"?

`r mcq(c("Saying 'not staring' when they were staring", answer = "Correctly detecting when someone IS staring", "Saying 'staring' when they were NOT staring", "Missing all trials"))`

What is a "false alarm"?

`r mcq(c("Correctly detecting when someone is staring", "Saying 'not staring' when they were staring", answer = "Saying 'staring' when they were NOT staring", "Missing all trials"))`

---

## Fill in the Blank

Complete these statements:

The four steps of the scientific method are: observe, `r fitb(c("hypothesise", "hypothesize", "hypothesis"), ignore_case = TRUE)`, test, and conclude.

A variable that can only take whole number values (like "number of words recalled") is called a `r fitb(c("discrete", "count"), ignore_case = TRUE)` variable.

---

## Summary Checklist

Before moving on, check you understand:

- [ ] Why psychology uses research methods rather than relying on common sense
- [ ] The four hallmarks of scientific claims
- [ ] The difference between continuous, categorical, and discrete variables
- [ ] The steps of the scientific method
- [ ] What "replicable" means and why it matters
