---
title: "Week 12 Self-Assessment"
subtitle: "Research Design Framework"
---

```{r, echo=FALSE}
library(webexercises)
```

## Instructions

This is the most important quiz of the term. Week 12 introduces ALL the terminology you need for your Research Plan. Make sure you can answer these questions confidently.

---

## Part 1: Variables

### Question 1: IV and DV Definitions

The variable that the researcher **manipulates or selects** is called the:

`r mcq(c("Dependent variable", answer = "Independent variable", "Control variable", "Confounding variable"))`

The variable that the researcher **measures** is called the:

`r mcq(c(answer = "Dependent variable", "Independent variable", "Control variable", "Confounding variable"))`

### Question 2: Identifying Variables

A researcher compares memory scores between participants who slept 8 hours versus 4 hours.

What is the IV? `r mcq(c(answer = "Sleep duration (8 hrs vs 4 hrs)", "Memory score", "Participant age", "Time of testing"))`

What is the DV? `r mcq(c("Sleep duration", answer = "Memory score", "Number of participants", "Room temperature"))`

### Question 3: DataLab Variables

In our **Room A vs Room B** comparison of reaction times:

The IV is: `r mcq(c(answer = "Room (A vs B)", "Reaction time", "Participant", "Time of day"))`

The DV is: `r mcq(c("Room assignment", answer = "Reaction time", "Number of trials", "Catch distance"))`

---

## Part 2: Design Types

### Question 4: Design Definitions

Match each design to its definition:

**Different participants** in each condition = `r mcq(c(answer = "Between-subjects", "Within-subjects", "Matched pairs", "Correlational"))`

**Same participants** tested in all conditions = `r mcq(c("Between-subjects", answer = "Within-subjects", "Matched pairs", "Correlational"))`

Participants **paired on characteristics**, one from each pair in each condition = `r mcq(c("Between-subjects", "Within-subjects", answer = "Matched pairs", "Correlational"))`

**No manipulation**, just measuring relationships = `r mcq(c("Between-subjects", "Within-subjects", "Matched pairs", answer = "Correlational"))`

### Question 5: DataLab Design Examples

What type of design is **Room A vs Room B** (different students in each room)?

`r mcq(c(answer = "Between-subjects", "Within-subjects", "Matched pairs", "Correlational"))`

What type of design is **Pre-mood vs Post-mood** (same students measured twice)?

`r mcq(c("Between-subjects", answer = "Within-subjects", "Matched pairs", "Correlational"))`

What type of design is **Confidence rating vs Throwing accuracy** (measuring relationship)?

`r mcq(c("Between-subjects", "Within-subjects", "Matched pairs", answer = "Correlational"))`

### Question 6: No IV

In a correlational design, there is no IV. `r torf(TRUE)`

---

## Part 3: Hypotheses

### Question 7: Hypothesis Types

A hypothesis that predicts a difference but NOT the direction (e.g., "scores will differ") is:

`r mcq(c("One-tailed", answer = "Two-tailed", "Null", "Alternative"))`

A hypothesis that predicts the direction (e.g., "Group A will score higher than Group B") is:

`r mcq(c(answer = "One-tailed", "Two-tailed", "Null", "Alternative"))`

### Question 8: Null vs Experimental

The **null hypothesis** predicts:

`r mcq(c("A significant effect", answer = "No effect or no difference", "The direction of the effect", "That the study will fail"))`

The **experimental hypothesis** (also called the alternative hypothesis) predicts:

`r mcq(c(answer = "An effect or difference exists", "No effect or difference", "The study is flawed", "Random variation"))`

### Question 9: Writing Hypotheses

Which is a correctly written **one-tailed** hypothesis?

```{r}
opts_hyp <- c(
  "Reaction times will differ between Room A and Room B",
  answer = "Participants in Room A will have faster reaction times than participants in Room B",
  "There will be no difference in reaction times between rooms",
  "Reaction times may or may not differ between rooms"
)
```

`r longmcq(opts_hyp)`

---

## Part 4: Test Selection

### Question 10: Design to Test Matching

Fill in the correct test for each design:

Between-subjects design → `r fitb(c("independent t-test", "independent t test", "independent samples t-test", "independent samples t test"), ignore_case = TRUE)`

Within-subjects design → `r fitb(c("paired t-test", "paired t test", "paired samples t-test", "related t-test"), ignore_case = TRUE)`

Correlational design → `r fitb(c("Pearson's r", "Pearsons r", "pearson", "correlation"), ignore_case = TRUE)`

### Question 11: Test Selection MCQ

A researcher compares anxiety levels before and after a relaxation intervention (same participants).

Which test should they use?

`r mcq(c("Independent t-test", answer = "Paired t-test", "Pearson's r", "Mann-Whitney U"))`

---

## Part 5: Descriptive Statistics

### Question 12: Calculation

Calculate the **mean** of these scores: 10, 15, 20, 25, 30

Mean = `r fitb(20)`

### Question 13: Standard Deviation

The standard deviation tells us:

`r mcq(c("The middle value", "The most common value", answer = "How spread out values are around the mean", "The difference between highest and lowest"))`

### Question 14: APA Format

In APA format, how do you report a mean of 25.50 and standard deviation of 4.30?

`r mcq(c("Mean = 25.50, SD = 4.30", "mean: 25.50, sd: 4.30", answer = "M = 25.50, SD = 4.30", "Average = 25.50 (4.30)"))`

---

## Part 6: MCQ Exam Practice

::: {.webex-check .webex-box}

### Investigation 1

A researcher wants to know if background music affects concentration. Half the participants complete a task with music playing, half complete it in silence. Concentration is measured by number of errors.

1. **Design**: `r mcq(c(answer = "Between-subjects", "Within-subjects", "Matched pairs", "Correlational"))`

2. **IV**: `r mcq(c(answer = "Music condition (music vs silence)", "Number of errors", "Concentration level", "No IV"))`

3. **DV**: `r mcq(c("Music condition", answer = "Number of errors", "Background noise level", "Participant preference"))`

4. **Appropriate test**: `r mcq(c(answer = "Independent t-test", "Paired t-test", "Pearson's r", "Chi-square"))`

:::

::: {.webex-check .webex-box}

### Investigation 2

A researcher measures students' anxiety before an exam and their exam performance to see if they are related.

1. **Design**: `r mcq(c("Between-subjects", "Within-subjects", "Matched pairs", answer = "Correlational"))`

2. **IV**: `r mcq(c("Anxiety level", "Exam performance", "Student year", answer = "No IV"))`

3. **DV**: `r mcq(c("Anxiety level", "Exam performance", answer = "Both variables are measured (no DV)", "Student motivation"))`

4. **Appropriate test**: `r mcq(c("Independent t-test", "Paired t-test", answer = "Pearson's r", "One-way ANOVA"))`

:::

::: {.webex-check .webex-box}

### Investigation 3

Participants rate their mood before and after watching a comedy video.

1. **Design**: `r mcq(c("Between-subjects", answer = "Within-subjects", "Matched pairs", "Correlational"))`

2. **IV**: `r mcq(c(answer = "Time (before vs after video)", "Mood rating", "Video content", "No IV"))`

3. **DV**: `r mcq(c("Time point", answer = "Mood rating", "Video length", "Participant age"))`

4. **Appropriate test**: `r mcq(c("Independent t-test", answer = "Paired t-test", "Pearson's r", "Mann-Whitney U"))`

:::

---

## Summary Checklist

Before moving on, make sure you can:

- [ ] Define IV (independent variable) and DV (dependent variable)
- [ ] Identify IV and DV in any research scenario
- [ ] Define all four design types (between, within, matched, correlational)
- [ ] Recognise when there is no IV (correlational design)
- [ ] Write one-tailed and two-tailed hypotheses
- [ ] Distinguish experimental from null hypotheses
- [ ] Select the correct test for each design type
- [ ] Calculate mean and interpret SD
- [ ] Report statistics in APA format

::: {.callout-important}
## Research Plan Connection
You need ALL of these skills for your Research Plan (due Week 16). If you're unsure about any concept, review the Week 12 lecture materials before continuing.
:::
