---
title: "Week 13 Self-Assessment"
subtitle: "Confidence Intervals, Bootstrap, and Effect Sizes"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
```

## Instructions

Test your understanding of confidence intervals, bootstrap methods, and effect sizes. These concepts are fundamental to understanding uncertainty in our estimates and the practical significance of research findings.

---

## Part 1: Conceptual Questions

### Question 1: What is a Confidence Interval?

A 95% confidence interval tells us:

`r mcq(c("There is a 95% chance the true population value is within this range", answer = "If we repeated this study many times, 95% of the CIs would contain the true population value", "95% of our sample data falls within this range", "We are 95% sure our hypothesis is correct"))`

### Question 2: Interpreting CI Width

A wider confidence interval indicates:

`r mcq(c("Greater precision in our estimate", answer = "Greater uncertainty in our estimate", "A larger effect size", "A significant result"))`

What factors can lead to a narrower (more precise) confidence interval?

`r mcq(c(answer = "Larger sample size and smaller variability", "Smaller sample size and larger variability", "Higher confidence level (e.g., 99% instead of 95%)", "More extreme scores in the data"))`

### Question 3: CI and Statistical Significance

If a 95% CI for the difference between two group means does NOT include zero, this suggests:

`r mcq(c("The result is not significant", answer = "The result is statistically significant at p < .05", "The effect size is large", "We need a larger sample"))`

If a 95% CI for a correlation coefficient includes zero, this suggests:

`r mcq(c(answer = "The correlation is not statistically significant at p < .05", "The correlation is significant", "We should use a different test", "The sample is too large"))`

### Question 4: Bootstrap Methods

The bootstrap method involves:

`r mcq(c("Collecting new participants", "Using a larger sample", answer = "Resampling with replacement from the original data to estimate sampling variability", "Removing outliers from the data"))`

Why might we use bootstrap methods instead of traditional formulas?

`r mcq(c("Bootstrap is always more accurate", answer = "Bootstrap makes fewer assumptions about the data distribution", "Bootstrap is required for all modern statistics", "Bootstrap gives smaller p-values"))`

### Question 5: Bootstrap Confidence Intervals

A bootstrap 95% CI is constructed by:

`r mcq(c("Adding and subtracting 2 standard errors from the mean", answer = "Finding the 2.5th and 97.5th percentiles of the bootstrap distribution", "Using the normal distribution tables", "Calculating the exact probability"))`

How many bootstrap samples are typically used?

`r mcq(c("10", "100", answer = "1000 or more", "Exactly 95"))`

### Question 6: Effect Sizes

An effect size tells us:

`r mcq(c("Whether the result is statistically significant", answer = "The magnitude or practical importance of an effect", "The probability of replication", "The sample size needed"))`

Why report effect sizes alongside p-values?

`r mcq(c("Because journals require it", answer = "Because statistical significance doesn't tell us about practical importance", "Because effect sizes are always larger than p-values", "Because p-values are unreliable"))`

---

## Part 2: Interpretation Questions

### Question 7: Interpreting a CI

A study reports: "Mean reaction time = 450ms, 95% CI [420, 480]"

The point estimate is: `r mcq(c("420ms", answer = "450ms", "480ms", "60ms"))`

The margin of error is approximately: `r mcq(c("420ms", "450ms", answer = "30ms", "480ms"))`

We can be reasonably confident the true population mean is between: `r mcq(c(answer = "420ms and 480ms", "0ms and 450ms", "450ms and 900ms", "We cannot say"))`

### Question 8: Comparing Groups with CIs

Study A finds: Group 1 mean = 50, 95% CI [45, 55]; Group 2 mean = 60, 95% CI [55, 65]

The CIs do NOT overlap. This suggests the difference is:

`r mcq(c("Not significant", answer = "Likely statistically significant", "Due to chance", "Meaningless"))`

Study B finds: Group 1 mean = 50, 95% CI [40, 60]; Group 2 mean = 55, 95% CI [45, 65]

The CIs DO overlap substantially. This suggests:

`r mcq(c("The groups are definitely different", answer = "The difference may not be statistically significant", "We should increase the confidence level", "The study failed"))`

### Question 9: Cohen's d Interpretation

Cohen's d = 0.20 is considered: `r mcq(c(answer = "Small", "Medium", "Large", "Negligible"))`

Cohen's d = 0.50 is considered: `r mcq(c("Small", answer = "Medium", "Large", "Very large"))`

Cohen's d = 0.80 is considered: `r mcq(c("Small", "Medium", answer = "Large", "Massive"))`

Cohen's d = 1.20 would be considered: `r mcq(c("Small", "Medium", answer = "Large (or very large)", "Impossible"))`

### Question 10: Interpreting Effect Size in Context

A medication reduces anxiety scores with d = 0.25 (small effect). A therapy reduces anxiety with d = 0.75 (medium-large effect).

Which treatment shows a larger practical effect? `r mcq(c("Medication", answer = "Therapy", "They are the same", "Cannot compare"))`

If the medication is much cheaper and has no side effects, should we ignore it because the effect is "small"?

`r mcq(c("Yes, small effects are worthless", answer = "No, small effects can still be meaningful in context", "Yes, only large effects matter", "We should only consider statistical significance"))`

---

## Part 3: Application Questions

### Question 11: Scenario - Understanding Uncertainty

A researcher measures spelling test scores in 30 children and reports: M = 75, 95% CI [70, 80].

Which statement is the best interpretation?

```{r}
opts_ci <- c(
  "95% of children scored between 70 and 80",
  "There is a 95% probability the true mean is between 70 and 80",
  answer = "We are 95% confident that the interval from 70 to 80 captures the true population mean",
  "The mean could be any value from 0 to 100"
)
```

`r longmcq(opts_ci)`

### Question 12: Scenario - Choosing Methods

A researcher has data that is heavily skewed (not normally distributed). They want to estimate the confidence interval for the median.

Which approach is most appropriate?

`r mcq(c("Use the standard formula assuming normality", answer = "Use bootstrap methods", "Ignore the skewness", "Only report the mean"))`

### Question 13: Scenario - Practical Significance

A pharmaceutical company tests a new drug for depression. They find a statistically significant improvement (p = .02) with d = 0.15.

How should they interpret this?

```{r}
opts_drug <- c(
  "The drug is highly effective and should be approved immediately",
  "The result is not significant, so the drug doesn't work",
  answer = "The effect is statistically significant but very small - clinical utility is questionable",
  "Effect sizes don't matter for drug trials"
)
```

`r longmcq(opts_drug)`

---

## Part 4: MCQ Exam Practice

::: {.webex-check .webex-box}

### Investigation 1

A researcher measures memory scores in a sample of 40 participants and calculates: M = 65, SD = 12, 95% CI [61.17, 68.83].

1. **What does the CI tell us about precision?** `r mcq(c("Very precise (narrow interval)", answer = "Moderately precise", "Very imprecise (wide interval)", "Cannot determine"))`

2. **If they had tested 100 participants instead, the CI would likely be:** `r mcq(c("Wider", answer = "Narrower", "The same", "Impossible to predict"))`

3. **If the population mean is actually 64, would this CI capture it?** `r torf(TRUE)`

4. **The margin of error is approximately:** `r mcq(c("65", "12", answer = "About 4", "61"))`

:::

::: {.webex-check .webex-box}

### Investigation 2

A study compares two teaching methods. Results show: Method A (M = 78, SD = 10) vs Method B (M = 72, SD = 11), with Cohen's d = 0.57 and p = .03.

1. **The effect size is:** `r mcq(c("Small", answer = "Medium", "Large", "Not significant"))`

2. **Is the result statistically significant at alpha = .05?** `r torf(TRUE)`

3. **Method A shows scores that are on average how much higher?** `r mcq(c("0.57 points higher", answer = "6 points higher", "57% higher", "Cannot calculate"))`

4. **The d = 0.57 means the difference is about:** `r mcq(c("0.57 points", answer = "About half a standard deviation", "57 standard deviations", "Not meaningful"))`

5. **Should we conclude Method A is practically better?** `r mcq(c("No, the effect is too small", answer = "Possibly - a medium effect in education can be meaningful", "Yes, definitely superior", "Cannot say without more data"))`

:::

---

## Summary Checklist

Before moving on, make sure you can:

- [ ] Explain what a confidence interval represents
- [ ] Interpret the width of a confidence interval
- [ ] Understand how sample size affects CI width
- [ ] Explain the basic logic of bootstrap methods
- [ ] Interpret Cohen's d values (small, medium, large)
- [ ] Distinguish statistical significance from practical significance
- [ ] Explain why effect sizes are important to report

::: {.callout-important}
## Key Takeaway
Statistical significance (p-value) tells us whether an effect is likely real, while effect size tells us whether it matters practically. Always consider both when interpreting research findings.
:::
