---
title: "Week 16 Self-Assessment"
subtitle: "Research Plan Tips"
---

```{r, echo=FALSE}
library(webexercises)
```

## Instructions

Your Research Plan is due this week. This quiz helps you check that you've covered all the key components. Use it as a final checklist before submitting.

---

## Part 1: Research Plan Components

### Question 1: Essential Elements

Which of the following MUST be included in your Research Plan?

Experimental hypothesis `r torf(TRUE)`

Identified IV and DV `r torf(TRUE)`

Design type (between/within/correlational) `r torf(TRUE)`

Full literature review `r torf(FALSE)`

Method section outline `r torf(TRUE)`

Statistical test selection `r torf(TRUE)`

### Question 2: Hypothesis Requirements

Your experimental hypothesis should be:

`r mcq(c("Vague and general", answer = "Specific, testable, and directional (for one-tailed) or non-directional (for two-tailed)", "Based only on your intuition", "Identical to the null hypothesis"))`

A one-tailed hypothesis must:

`r mcq(c("Predict that variables are related", answer = "Specify the direction of the predicted effect", "Be longer than a two-tailed hypothesis", "Include statistical notation"))`

---

## Part 2: Variable Identification

### Question 3: IV Characteristics

The IV in an experimental study is:

`r mcq(c(answer = "Manipulated by the researcher", "Measured as the outcome", "Always a continuous variable", "The same as the DV"))`

In a correlational design, there is:

`r mcq(c("One IV and one DV", "Two IVs", answer = "No IV — both variables are measured", "Only a DV"))`

### Question 4: DV Characteristics

The DV should be:

`r mcq(c(answer = "Operationally defined (how exactly will you measure it?)", "Vague and flexible", "The same for all studies", "Categorical only"))`

### Question 5: Practice - Identify Variables

A researcher wants to test if caffeine improves reaction time. Participants are randomly assigned to drink coffee or water, then complete a reaction time task.

IV: `r mcq(c(answer = "Drink type (coffee vs water)", "Reaction time", "Participant assignment", "Task difficulty"))`

DV: `r mcq(c("Caffeine content", answer = "Reaction time", "Drink preference", "Group membership"))`

---

## Part 3: Design Specification

### Question 6: Design Selection

```{r}
opts_design <- c(
  "It determines the statistical test to use",
  "It clarifies how participants are allocated",
  "It affects the type of conclusions you can draw",
  answer = "All of the above"
)
```

Why is specifying your design type important?

`r longmcq(opts_design)`

### Question 7: Design Justification

If you use a within-subjects design, you should justify it by noting:

`r mcq(c(answer = "It controls for individual differences between participants", "It requires more participants", "It's always better than between-subjects", "It doesn't need justification"))`

If you use a between-subjects design, a key consideration is:

`r mcq(c(answer = "Random assignment to conditions to avoid confounds", "Testing participants multiple times", "Correlation between variables", "Using the same participants twice"))`

---

## Part 4: Test Selection

### Question 8: Matching Design to Test

Fill in the appropriate test:

Between-subjects design (comparing two groups) → `r fitb(c("independent t-test", "independent t test", "independent samples t-test"), ignore_case = TRUE)`

Within-subjects design (comparing same people twice) → `r fitb(c("paired t-test", "paired t test", "paired samples t-test", "related t-test"), ignore_case = TRUE)`

Correlational design (measuring relationship) → `r fitb(c("Pearson's r", "Pearsons r", "correlation", "pearson"), ignore_case = TRUE)`

### Question 9: Test Justification

Your Research Plan should explain why your chosen test is appropriate. Which is the best justification for using an independent t-test?

```{r}
opts_justify <- c(
  "Because it's the easiest test",
  answer = "Because we have two separate groups (between-subjects) and continuous DV data",
  "Because it always gives significant results",
  "Because the lecturer said to use it"
)
```

`r longmcq(opts_justify)`

---

## Part 5: Method Section Preview

### Question 10: Method Components

The Method section of a research report typically includes:

Participants `r torf(TRUE)`

Design `r torf(TRUE)`

Full statistical results `r torf(FALSE)`

Materials `r torf(TRUE)`

Procedure `r torf(TRUE)`

### Question 11: Participants Section

What should the Participants section include?

`r mcq(c(answer = "Sample size, demographics, and how participants were recruited", "The statistical test used", "The results of the study", "The hypothesis"))`

### Question 12: Procedure Section

The Procedure section should describe:

`r mcq(c("Only what participants did wrong", answer = "Step-by-step what happened to participants during the study", "The statistical analysis", "The theory behind the study"))`

---

## Part 6: Common Mistakes to Avoid

### Question 13: Hypothesis Errors

Which is a correctly written **experimental hypothesis**?

`r mcq(c("Caffeine might affect memory", answer = "Participants who consume caffeine will recall more words than those who consume water", "There is no effect of caffeine on memory", "Memory and caffeine are variables"))`

### Question 14: Design-Test Mismatch

A researcher claims to have a "between-subjects correlational design". This is:

`r mcq(c("A valid combination", answer = "Incorrect — correlational designs measure relationships, not group differences", "The best type of design", "Required for the Research Plan"))`

### Question 15: Data Table

Your Research Plan should include a summary data table. This table should:

`r mcq(c("Contain made-up data", "Show individual participant scores", answer = "Show the structure for presenting means and SDs (without actual data yet)", "Include all raw data"))`

---

## Part 7: Final Checklist

::: {.webex-check .webex-box}

### Research Plan Self-Check

Before submitting, confirm you have:

1. **A clear experimental hypothesis?** (one-tailed or two-tailed, justified) `r mcq(c(answer = "Yes", "No", "Unsure"))`

2. **Identified your IV?** (what you manipulate or compare) `r mcq(c(answer = "Yes", "No", "Not applicable"))`

3. **Identified your DV?** (what you measure) `r mcq(c(answer = "Yes", "No", "Unsure"))`

4. **Specified your design?** (between/within/correlational) `r mcq(c(answer = "Yes", "No", "Unsure"))`

5. **Selected your statistical test?** (and justified it) `r mcq(c(answer = "Yes", "No", "Unsure"))`

6. **Outlined your Method section?** (Participants, Design, Materials, Procedure) `r mcq(c(answer = "Yes", "No", "Partially"))`

7. **Included a data summary table template?** `r mcq(c(answer = "Yes", "No", "What's that?"))`

:::

---

## Summary Checklist

Before submitting your Research Plan, verify:

- [ ] Experimental hypothesis is specific and testable
- [ ] IV and DV are clearly identified and operationally defined
- [ ] Design type is specified and appropriate for your research question
- [ ] Statistical test matches your design
- [ ] Method section outlines Participants, Design, Materials, Procedure
- [ ] Data summary table shows expected structure (without fabricated data)

::: {.callout-important}
## Research Plan Due This Week
Submit your Research Plan by the deadline. This assessment is worth 20% of your module grade. Review the assessment brief on Canvas for formatting requirements.
:::
