---
title: "Week 19 Self-Assessment"
subtitle: "APA Reporting Format"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
```

## Instructions

Test your understanding of APA format for reporting statistical results. Correct reporting is essential for scientific communication and is assessed in your lab report and exam.

---

## Part 1: Conceptual Questions

### Question 1: Why APA Format?

APA format for statistics is important because:

`r mcq(c("It makes papers look professional", answer = "It provides a standardised way to communicate findings that all researchers understand", "Journals require it for publication", "All of the above"))`

### Question 2: General Formatting Rules

In APA format, statistical symbols (like *t*, *p*, *r*, *M*, *SD*) should be:

`r mcq(c("Bold", answer = "Italicised", "Underlined", "In quotation marks"))`

How many decimal places are typically reported for means and standard deviations?

`r mcq(c("0", "1", answer = "2", "As many as available"))`

### Question 3: Leading Zeros

For values that cannot exceed 1 (like *p*, *r*, and proportions), APA format:

`r mcq(c("Always uses a leading zero (0.05)", answer = "Does not use a leading zero (.05)", "Uses both interchangeably", "Uses percent instead"))`

Which is correct for reporting a p-value?

`r mcq(c("p = 0.032", answer = "p = .032", "P = .032", "p-value = 0.032"))`

### Question 4: Very Small p-Values

When a p-value is very small (e.g., .00004), the correct way to report it is:

`r mcq(c("p = .00004", "p = .000", answer = "p < .001", "p is significant"))`

### Question 5: Degrees of Freedom

Degrees of freedom should be reported:

`r mcq(c("At the end of the sentence", answer = "In parentheses after the test statistic symbol", "Only if requested", "In a separate table"))`

### Question 6: Rounding Rules

Values should generally be rounded to:

`r mcq(c("Whole numbers", answer = "Two decimal places", "Four decimal places", "The nearest tenth"))`

When should you use three decimal places?

`r mcq(c("For all values", answer = "For very small p-values (e.g., p = .002) or when more precision is needed", "Never", "Only for correlations"))`

---

## Part 2: Interpretation - Reporting t-Tests

### Question 7: Independent t-Test Format

The correct APA format for an independent t-test is:

`r mcq(c("t = 2.45, p = .018", answer = "t(48) = 2.45, p = .018, d = 0.71", "t-test: 2.45, significant", "t(n=50) = 2.45, sig = .018"))`

### Question 8: Paired t-Test Format

A paired t-test with 25 participants would be reported as:

`r mcq(c("t(25) = 3.12, p = .004", answer = "t(24) = 3.12, p = .004, d = 0.62", "t = 3.12, df = 25, p = .004", "paired t = 3.12"))`

Why df = 24 for n = 25 in a paired test?

`r mcq(c("It's a mistake", answer = "For paired tests, df = n - 1", "Half the sample is in each condition", "One participant was excluded"))`

### Question 9: Negative t-Values

When reporting t(28) = -2.89, should the negative sign be included?

`r mcq(c("No, always report absolute values", answer = "Yes, it indicates the direction of the effect", "Only if specifically asked", "Negative values are errors"))`

### Question 10: Complete t-Test Reporting

Which is the most complete and correct report?

```{r}
opts_t <- c(
  "The t-test was significant (p < .05)",
  "t = 2.45, p = .018",
  answer = "An independent samples t-test revealed a significant difference, t(48) = 2.45, p = .018, d = 0.71",
  "There was a significant effect of condition on the DV"
)
```

`r longmcq(opts_t)`

---

## Part 3: Interpretation - Reporting Correlations

### Question 11: Correlation Format

The correct APA format for a correlation is:

`r mcq(c("r = .45, significant", answer = "r(48) = .45, p = .001", "correlation = 0.45, p-value = 0.001", "Pearson r: .45*"))`

### Question 12: Degrees of Freedom for Correlations

For a correlation with n = 50, df =

`r mcq(c("50", "49", answer = "48", "100"))`

The formula is: `r mcq(c("df = n", "df = n - 1", answer = "df = n - 2", "df = n/2"))`

### Question 13: Reporting Correlation Strength

When describing r = -.62 in text, you might write:

`r mcq(c("A weak negative correlation", answer = "A moderate-to-strong negative correlation", "A strong positive correlation", "No correlation"))`

### Question 14: R-Squared Reporting

If you report r2, how many decimal places?

`r mcq(c("0", "1", answer = "2 (e.g., r2 = .36)", "4"))`

Reporting both r and r2:

`r mcq(c("Is redundant and unnecessary", answer = "Can be useful as r2 shows proportion of variance explained", "Is required in all cases", "Is incorrect"))`

---

## Part 3: Interpretation - Reporting Descriptive Statistics

### Question 15: Means and SDs

The correct format for reporting a mean and standard deviation is:

`r mcq(c("mean = 45.30, std = 8.25", "Average: 45.30 (8.25)", answer = "M = 45.30, SD = 8.25", "45.30 +/- 8.25"))`

### Question 16: Reporting Groups

When comparing two groups, descriptive statistics should:

`r mcq(c("Only be in a table", answer = "Be reported for each group, either in text or in a table", "Only include the combined mean", "Be omitted if there's a significant difference"))`

Example: "The experimental group (*M* = 72.40, *SD* = 8.50) scored higher than the control group (*M* = 65.20, *SD* = 9.10)."

Is this correct? `r torf(TRUE)`

---

## Part 4: Application Questions

### Question 17: Correcting Errors

Identify what is WRONG with each report:

"t(48) = 2.45, p = 0.018" - Error: `r mcq(c("Wrong degrees of freedom", answer = "Leading zero on p-value", "Missing effect size", "All of these"))`

"p = .000" - Error: `r mcq(c("Wrong symbol", answer = "Should be p < .001 (p cannot equal exactly zero)", "Too many zeros", "Needs italics"))`

"The correlation was r = .45 (p = significant)" - Error: `r mcq(c("Correlation is wrong", answer = "Should report exact p-value, not just 'significant'", "r should be italicised", "Multiple errors"))`

### Question 18: Writing a Complete Results Sentence

Fill in the blanks for correct APA format:

"An independent samples t-test revealed that participants in the music condition (`r mcq(c(answer = "M = 72.40, SD = 8.50", "mean = 72.40, std = 8.50", "72.40 (8.50)", "M: 72.40"))`) recalled significantly more words than those in the silence condition (`r mcq(c(answer = "M = 65.20, SD = 9.10", "mean = 65.20", "65.20 average", "M = 65.20 standard deviation = 9.10"))`), `r mcq(c("t = 2.89, p = .006", answer = "t(48) = 2.89, p = .006, d = 0.82", "t-test significant", "t(48) = 2.89, significance = .006"))`."

### Question 19: Results vs Discussion Content

Which belongs in the Results section?

`r mcq(c(answer = "t(48) = 2.89, p = .006, d = 0.82", "This finding supports Smith's (2020) theory", "The study had several limitations", "Future research should examine..."))`

Which belongs in the Discussion section?

`r mcq(c("M = 45.30, SD = 8.25", answer = "These results are consistent with previous research showing...", "r(48) = .45, p = .001", "The experimental group scored higher"))`

---

## Part 5: MCQ Exam Practice

::: {.webex-check .webex-box}

### Scenario 1: Report the t-Test

Independent t-test results: Group A (n=25): M=78.4, SD=10.2; Group B (n=25): M=72.1, SD=11.5; t=2.05, p=.046, d=0.58

1. **Degrees of freedom:** `r mcq(c("50", answer = "48", "25", "24"))`

2. **Correct p-value format:** `r mcq(c("p = 0.046", answer = "p = .046", "p < .05", "sig = .046"))`

3. **Effect size is:** `r mcq(c("Small", answer = "Medium", "Large", "Not reported"))`

4. **Complete correct report:**

```{r}
opts_report <- c(
  "t = 2.05, p = .046",
  "t(50) = 2.05, p = 0.046, d = 0.58",
  answer = "t(48) = 2.05, p = .046, d = 0.58",
  "The groups were significantly different"
)
```

`r longmcq(opts_report)`

:::

::: {.webex-check .webex-box}

### Scenario 2: Report the Correlation

Correlation results: n=60, r=.42, p=.0008

1. **Degrees of freedom:** `r mcq(c("60", "59", answer = "58", "42"))`

2. **How should p be reported?** `r mcq(c("p = .0008", "p = 0.001", answer = "p < .001", "p = .001"))`

3. **Correct report:** `r mcq(c("r = .42, p < .001", answer = "r(58) = .42, p < .001", "correlation = .42, significant", "r(60) = .42, p = .0008"))`

4. **Describing the relationship:** `r mcq(c("Weak", answer = "Moderate positive", "Strong", "Negative"))`

:::

::: {.webex-check .webex-box}

### Scenario 3: Spot the Errors

A student writes: "We found a significant effect (t = 2.34, p = 0.023). The mean for Group 1 was 56.7 and Group 2 was 48.2."

Identify the errors:

1. **Missing from t-test report:** `r mcq(c(answer = "Degrees of freedom and effect size", "The p-value", "Significance statement", "Group labels"))`

2. **p-value format error:** `r mcq(c(answer = "Should not have leading zero (should be p = .023)", "Should be p < .05", "Should be rounded", "No error"))`

3. **Descriptive statistics error:** `r mcq(c(answer = "Missing SDs and proper formatting (M = , SD = )", "Means are wrong", "Should only report one group", "Numbers should be whole"))`

:::

---

## Summary Checklist

Before writing your Results section, verify:

- [ ] Statistical symbols are italicised (*t*, *p*, *r*, *M*, *SD*, *d*)
- [ ] Degrees of freedom in parentheses after test statistic
- [ ] No leading zeros for values that cannot exceed 1 (*p*, *r*)
- [ ] Two decimal places for most values
- [ ] *p* < .001 for very small p-values (never *p* = .000)
- [ ] Effect sizes reported
- [ ] Descriptive statistics (M, SD) for each group/condition
- [ ] Results reported without extensive interpretation

::: {.callout-important}
## Common Mistakes to Avoid
1. Writing "p = 0.023" instead of "p = .023"
2. Writing "p = .000" instead of "p < .001"
3. Forgetting degrees of freedom: "t = 2.45" instead of "t(48) = 2.45"
4. Omitting effect sizes
5. Not italicising statistical symbols
:::
