---
title: "Estimation & Confidence Intervals"
subtitle: "PS50008A Week 13 | Lecture"
author: "Gordon Wright"
date: "2026-01-28"
format:
  revealjs:
    theme: [default, ../../assets/slides.scss]
    slide-number: c/t
    progress: true
    transition: fade
    width: 1050
    height: 700
    center: false
    center-title-slide: true
execute:
  echo: false
  warning: false
  message: false
---

## Learning Outcomes

By the end of this lecture you will be able to:

- Explain the difference between point estimates and interval estimates
- Interpret confidence intervals correctly
- Understand the bootstrap method for estimation
- Calculate and interpret effect sizes using Cohen's d

# Point Estimates {background-color="#275882"}

## What is a Point Estimate?

<!-- TODO: Define point estimates -->

::: {.definition-box}
**Point Estimate**: A single value calculated from sample data that serves as a "best guess" for an unknown population parameter.
:::

::: {.fragment}
**Examples of point estimates:**

- Sample mean as estimate of population mean
- Sample proportion as estimate of population proportion
- Sample standard deviation as estimate of population SD
:::

## Sample Statistics as Estimators

<!-- TODO: Add content about sample statistics -->

:::: {.columns}

::: {.column width="50%"}
**Population Parameters**

- Fixed but unknown
- Greek letters (mu, sigma)
- What we want to know
:::

::: {.column width="50%"}
**Sample Statistics**

- Calculated from data
- Roman letters (x-bar, s)
- Our best estimates
:::

::::

::: {.fragment}
<!-- TODO: Add example with real data -->
:::

## The Problem with Point Estimates

<!-- TODO: Discuss limitations -->

::: {.incremental}
- A single number gives no indication of uncertainty
- Different samples give different estimates
- How confident can we be in our estimate?
:::

::: {.fragment}
::: {.example-box}
**Example:** If our sample mean is 75, is the population mean likely to be 74? 70? 50?
:::
:::

# Interval Estimates & Confidence Intervals {background-color="#275882"}

## From Points to Intervals

<!-- TODO: Introduce interval estimation -->

::: {.definition-box}
**Interval Estimate**: A range of values that is likely to contain the population parameter.
:::

::: {.fragment}
Instead of saying "the mean is 75", we say "the mean is between 72 and 78".
:::

## What is a Confidence Interval?

<!-- TODO: Define CIs properly -->

::: {.definition-box}
**Confidence Interval (CI)**: An interval estimate combined with a probability statement about how often such intervals capture the true parameter.
:::

::: {.fragment}
**Common confidence levels:**

- 95% CI (most common)
- 99% CI (more conservative)
- 90% CI (less common)
:::

## Interpreting Confidence Intervals

<!-- TODO: Add correct vs incorrect interpretations -->

::: {.callout-warning}
## Common Misconception
A 95% CI does NOT mean there is a 95% probability that the true parameter lies within this specific interval.
:::

::: {.fragment}
::: {.callout-tip}
## Correct Interpretation
If we repeated our study many times, 95% of the confidence intervals we calculate would contain the true population parameter.
:::
:::

## Visualising Confidence Intervals

<!-- TODO: Add visual demonstration -->

```{r}
#| eval: false
# TODO: Add R code to visualise multiple CIs
# Show how most but not all capture the true parameter
```

## Width of Confidence Intervals

<!-- TODO: Explain factors affecting CI width -->

**Factors affecting CI width:**

::: {.incremental}
- **Sample size**: Larger n = narrower CI
- **Variability**: More variation = wider CI
- **Confidence level**: Higher confidence = wider CI
:::

# Bootstrap Method {background-color="#275882"}

## Why Bootstrap?

<!-- TODO: Introduce bootstrap rationale -->

::: {.incremental}
- Traditional CIs assume normal distribution
- What if our data isn't normal?
- What if we have small samples?
- Bootstrap provides a flexible alternative
:::

## The Bootstrap Idea

<!-- TODO: Explain resampling concept -->

::: {.definition-box}
**Bootstrap**: A resampling method that treats your sample as if it were the population, drawing many new samples (with replacement) to estimate sampling variability.
:::

::: {.fragment}
<!-- TODO: Add step-by-step explanation -->
:::

## Bootstrap Steps

<!-- TODO: Outline the procedure -->

::: {.incremental}
1. Take your original sample (n observations)
2. Resample n observations WITH replacement
3. Calculate the statistic of interest
4. Repeat steps 2-3 many times (e.g., 10,000)
5. Use the distribution of statistics to build CI
:::

## Bootstrap in Action

<!-- TODO: Add worked example -->

```{r}
#| eval: false
# TODO: Add R code demonstrating bootstrap
```

::: {.example-box}
**Example:** Bootstrapping the mean of reaction times
:::

## Percentile Bootstrap CI

<!-- TODO: Show how to construct CI -->

::: {.fragment}
For a 95% CI:

- Find the 2.5th percentile of bootstrap statistics
- Find the 97.5th percentile of bootstrap statistics
- These form your interval
:::

# Effect Sizes (Cohen's d) {background-color="#275882"}

## Why Effect Sizes Matter

<!-- TODO: Introduce effect sizes -->

::: {.incremental}
- Statistical significance tells us IF there's an effect
- Effect size tells us HOW BIG the effect is
- Essential for practical importance
- Allows comparison across studies
:::

## What is Cohen's d?

<!-- TODO: Define Cohen's d -->

::: {.definition-box}
**Cohen's d**: A standardised measure of the difference between two means, expressed in standard deviation units.
:::

$$d = \frac{\bar{X}_1 - \bar{X}_2}{s_{pooled}}$$

## Interpreting Cohen's d

<!-- TODO: Add Cohen's conventions -->

**Cohen's (1988) conventions:**

| d value | Interpretation |
|---------|----------------|
| 0.2     | Small effect   |
| 0.5     | Medium effect  |
| 0.8     | Large effect   |

::: {.fragment}
::: {.callout-note}
These are rough guidelines, not strict rules. Context matters!
:::
:::

## Visualising Effect Sizes

<!-- TODO: Add visual comparison of effect sizes -->

```{r}
#| eval: false
# TODO: Add R code showing overlapping distributions
# for small, medium, and large effects
```

## Effect Size and Sample Size

<!-- TODO: Discuss relationship -->

::: {.example-box}
**Key insight:** A small effect can be statistically significant with a large enough sample. A large effect might not reach significance with a small sample.
:::

::: {.fragment}
This is why reporting both p-values AND effect sizes is crucial.
:::

# Summary {background-color="#275882"}

## Key Takeaways

::: {.incremental}
- **Point estimates** are single-value guesses about population parameters
- **Confidence intervals** provide a range plus uncertainty information
- **Bootstrap** offers a flexible, distribution-free approach to CIs
- **Cohen's d** quantifies effect magnitude in standardised units
:::

## Next Week

<!-- TODO: Update with next week's topic -->

**Week 14: Inference & t-tests**

- The logic of hypothesis testing
- Understanding p-values
- Comparing means between groups

## Questions?

::: {.r-fit-text}
Questions?
:::

See you in the lab!
