---
title: "Correlation"
subtitle: "PS50008A Week 15 | Lecture"
author: "Gordon Wright"
date: "2026-02-11"
format:
  revealjs:
    theme: [default, ../../assets/slides.scss]
    slide-number: c/t
    progress: true
    transition: fade
    width: 1050
    embed-resources: true
    height: 700
    center: false
    center-title-slide: true
execute:
  echo: false
  warning: false
  message: false
---

## Learning Outcomes

By the end of this lecture you will be able to:

- Describe relationships between continuous variables
- Create and interpret scatterplots
- Calculate and interpret Pearson's r and Spearman's rho
- Explain why correlation does not imply causation

# Relationships Between Variables {background-color="#275882"}

## Beyond Group Differences

<!-- Introduce correlational thinking -->

::: {.incremental}
- So far: comparing groups (t-tests)
- Now: examining relationships between variables
- Both variables are measured, not manipulated
:::

::: {.fragment}
::: {.example-box}
**Question:** Is there a relationship between study hours and exam scores?
:::
:::

## What is Correlation?

<!-- Define correlation -->

::: {.definition-box}
**Correlation**: A statistical measure that quantifies the strength and direction of the linear relationship between two continuous variables.
:::

## Direction of Relationships

```{r}
#| fig-width: 10
#| fig-height: 4.5
library(ggplot2)
set.seed(42)
n <- 50

pos_data <- data.frame(
  x = rnorm(n, 50, 10),
  type = "Positive"
)
pos_data$y <- 0.8 * pos_data$x + rnorm(n, 0, 5)

neg_data <- data.frame(
  x = rnorm(n, 50, 10),
  type = "Negative"
)
neg_data$y <- 100 - 0.8 * neg_data$x + rnorm(n, 0, 5)

plot_data <- rbind(pos_data, neg_data)
plot_data$type <- factor(plot_data$type, levels = c("Positive", "Negative"))

ggplot(plot_data, aes(x = x, y = y)) +
  geom_point(colour = "#275882", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#EA8439", linewidth = 1.5) +
  facet_wrap(~type, scales = "free_y") +
  theme_minimal(base_size = 18) +
  labs(x = "Variable X", y = "Variable Y") +
  theme(
    strip.text = element_text(face = "bold", size = 20),
    panel.grid.minor = element_blank()
  )
```

:::: {.columns}

::: {.column width="50%"}
**Positive**: As X ↑, Y ↑

*Example: Height and weight*
:::

::: {.column width="50%"}
**Negative**: As X ↑, Y ↓

*Example: Stress and sleep quality*
:::

::::

## Strength of Relationships

```{r}
#| fig-width: 10
#| fig-height: 5
set.seed(123)
n <- 40

make_corr_data <- function(r, label) {
  x <- rnorm(n, 50, 10)
  y <- r * scale(x) + sqrt(1 - r^2) * rnorm(n)
  data.frame(x = x, y = y * 10 + 50, strength = label)
}

strength_data <- rbind(
  make_corr_data(0.95, "Strong (r = 0.95)"),
  make_corr_data(0.50, "Moderate (r = 0.50)"),
  make_corr_data(0.15, "Weak (r = 0.15)")
)
strength_data$strength <- factor(strength_data$strength,
  levels = c("Strong (r = 0.95)", "Moderate (r = 0.50)", "Weak (r = 0.15)"))

ggplot(strength_data, aes(x = x, y = y)) +
  geom_point(colour = "#275882", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#EA8439", linewidth = 1.5) +
  facet_wrap(~strength) +
  theme_minimal(base_size = 18) +
  labs(x = "Variable X", y = "Variable Y") +
  theme(
    strip.text = element_text(face = "bold", size = 16),
    panel.grid.minor = element_blank()
  )
```

## Types of Relationships

```{r}
#| fig-width: 10
#| fig-height: 5
set.seed(456)
n <- 50

x_lin <- seq(20, 80, length.out = n)
y_lin <- 0.8 * x_lin + rnorm(n, 0, 5)

x_curve <- seq(20, 80, length.out = n)
y_curve <- -0.03 * (x_curve - 50)^2 + 60 + rnorm(n, 0, 3)

x_none <- rnorm(n, 50, 12)
y_none <- rnorm(n, 50, 12)

type_data <- rbind(
  data.frame(x = x_lin, y = y_lin, type = paste0("Linear (r = ", round(cor(x_lin, y_lin), 2), ")")),
  data.frame(x = x_curve, y = y_curve, type = paste0("Curvilinear (r = ", round(cor(x_curve, y_curve), 2), ")")),
  data.frame(x = x_none, y = y_none, type = paste0("No Relationship (r = ", round(cor(x_none, y_none), 2), ")"))
)
type_data$type <- factor(type_data$type, levels = unique(type_data$type))

ggplot(type_data, aes(x = x, y = y)) +
  geom_point(colour = "#275882", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#EA8439", linewidth = 1.5) +
  facet_wrap(~type) +
  theme_minimal(base_size = 18) +
  labs(x = "Variable X", y = "Variable Y") +
  theme(
    strip.text = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

::: {.callout-warning}
Correlation measures LINEAR relationships only. The curvilinear data has a strong pattern but r ≈ 0!
:::

# Scatterplots {background-color="#275882"}

## Visualising Relationships

::: {.definition-box}
**Scatterplot**: A graph displaying the relationship between two continuous variables, with each observation represented as a point.
:::

::: {.fragment}
- X-axis: One variable (often predictor)
- Y-axis: Other variable (often outcome)
- Each point: One observation
:::

## Creating Good Scatterplots

::: {.incremental}
- Label axes clearly (with units)
- Use appropriate scales
- Include a trend line (with caution)
- Look for patterns, outliers, clusters
:::

## Example: Study Hours vs Exam Scores

```{r}
#| fig-width: 8
#| fig-height: 5
set.seed(789)
n <- 60

study_data <- data.frame(
  study_hours = pmax(1, rnorm(n, 15, 6)),
  noise = rnorm(n, 0, 8)
)
study_data$exam_score <- pmin(100, pmax(30, 35 + 3.2 * study_data$study_hours + study_data$noise))

r_val <- round(cor(study_data$study_hours, study_data$exam_score), 2)

ggplot(study_data, aes(x = study_hours, y = exam_score)) +
  geom_point(colour = "#275882", size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = "#EA8439", fill = "#EA8439", alpha = 0.2, linewidth = 1.5) +
  theme_minimal(base_size = 20) +
  labs(
    x = "Study Hours per Week",
    y = "Exam Score (%)",
    title = paste0("r = ", r_val)
  ) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) +
  scale_y_continuous(limits = c(30, 100))
```

## What to Look For

::: {.example-box}
**When examining a scatterplot, ask:**

1. Is there a relationship? (or random scatter?)
2. What direction? (positive/negative)
3. How strong? (tight/loose clustering)
4. Is it linear? (or curved?)
5. Any outliers? (unusual points)
:::

## The Impact of Outliers

```{r}
#| fig-width: 10
#| fig-height: 4.5
set.seed(321)
n <- 30

base_x <- rnorm(n, 50, 8)
base_y <- 0.7 * scale(base_x) * 8 + 50 + rnorm(n, 0, 4)

no_outlier <- data.frame(x = base_x, y = base_y)
with_outlier <- rbind(no_outlier, data.frame(x = 80, y = 25))

r_without <- round(cor(no_outlier$x, no_outlier$y), 2)
r_with <- round(cor(with_outlier$x, with_outlier$y), 2)

no_outlier$group <- paste0("Without Outlier (r = ", r_without, ")")
with_outlier$group <- paste0("With Outlier (r = ", r_with, ")")
with_outlier$is_outlier <- c(rep(FALSE, n), TRUE)
no_outlier$is_outlier <- FALSE

outlier_data <- rbind(no_outlier, with_outlier)
outlier_data$group <- factor(outlier_data$group, levels = unique(outlier_data$group))

ggplot(outlier_data, aes(x = x, y = y)) +
  geom_point(aes(colour = is_outlier, size = is_outlier), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#EA8439", linewidth = 1.5) +
  facet_wrap(~group) +
  scale_colour_manual(values = c("FALSE" = "#275882", "TRUE" = "red")) +
  scale_size_manual(values = c("FALSE" = 3, "TRUE" = 5)) +
  theme_minimal(base_size = 18) +
  labs(x = "Variable X", y = "Variable Y") +
  theme(
    strip.text = element_text(face = "bold", size = 16),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

::: {.callout-warning}
A single outlier dramatically changed r from 0.70 to 0.35! Always visualise first.
:::

# Pearson's r and Spearman's rho {background-color="#275882"}

## Pearson's Correlation Coefficient

<!-- Define Pearson's r -->

::: {.definition-box}
**Pearson's r**: A measure of the linear relationship between two continuous variables. Ranges from -1 to +1.
:::

$$r = \frac{\sum(X - \bar{X})(Y - \bar{Y})}{(n-1)s_X s_Y}$$

## Interpreting Pearson's r

<!-- Add interpretation guide -->

| r value | Interpretation |
|---------|----------------|
| +/- 1.0 | Perfect correlation |
| +/- 0.7 to 0.9 | Strong |
| +/- 0.4 to 0.6 | Moderate |
| +/- 0.1 to 0.3 | Weak |
| 0 | No linear relationship |

::: {.fragment}
::: {.callout-note}
These are rough guidelines. Context and field norms matter!
:::
:::

## Coefficient of Determination (r^2)

<!-- Explain r-squared -->

::: {.definition-box}
**r^2 (r-squared)**: The proportion of variance in Y that is explained by X. Found by squaring r.
:::

::: {.fragment}
::: {.example-box}
If r = 0.6, then r^2 = 0.36

"36% of the variance in exam scores is explained by study hours."
:::
:::

## Assumptions of Pearson's r

<!-- List assumptions -->

::: {.incremental}
- Both variables are continuous
- Linear relationship (check scatterplot!)
- Bivariate normality (roughly)
- No extreme outliers
- Independence of observations
:::

## When Pearson's r Fails

<!-- Introduce need for alternatives -->

::: {.callout-warning}
Pearson's r is inappropriate when:

- Data are ordinal (ranked)
- Relationship is non-linear
- There are extreme outliers
- Data are severely non-normal
:::

## Spearman's Rho (rho)

<!-- Define Spearman's rho -->

::: {.definition-box}
**Spearman's rho**: A non-parametric correlation based on ranks rather than raw values. More robust to outliers and non-linearity.
:::

::: {.fragment}
- Converts data to ranks
- Then calculates Pearson's r on ranks
- Ranges from -1 to +1 (same interpretation)
:::

## Pearson vs Spearman

<!-- Comparison table -->

:::: {.columns}

::: {.column width="50%"}
**Use Pearson's r when:**

- Both variables continuous
- Linear relationship
- Roughly normal data
- No major outliers
:::

::: {.column width="50%"}
**Use Spearman's rho when:**

- Ordinal data
- Non-linear but monotonic
- Non-normal data
- Outliers present
:::

::::

## Testing Significance

<!-- Add significance testing -->

::: {.example-box}
**Null hypothesis:** r = 0 (no correlation in population)

**Test:** Is our observed r significantly different from 0?

**Report:** r(df) = value, p = value
:::

# Correlation Does Not Equal Causation {background-color="#275882"}

## The Third Variable Problem

<!-- Explain confounds -->

::: {.definition-box}
**Third Variable (Confound)**: An unmeasured variable that causes both X and Y, creating a spurious correlation between them.
:::

::: {.fragment}
::: {.example-box}
**Ice cream sales and drowning deaths are correlated.**

Third variable: Hot weather causes both!
:::
:::

## Directionality Problem

<!-- Explain directionality -->

::: {.incremental}
- Correlation shows X and Y are related
- But which causes which?
- X -> Y?
- Y -> X?
- Or both directions?
:::

::: {.fragment}
::: {.example-box}
**Depression and social isolation are correlated.**

Does depression cause isolation, or isolation cause depression, or both?
:::
:::

## Famous Spurious Correlations

<!-- Add entertaining examples -->

::: {.example-box}
**Real correlations that aren't causal:**

- Number of pirates and global temperature (r = -0.9)
- Nicolas Cage films and pool drownings (r = 0.87)
- Cheese consumption and death by bedsheet (r = 0.95)
:::

::: {.fragment}
[tylervigen.com/spurious-correlations](https://tylervigen.com/spurious-correlations)
:::

## When CAN We Infer Causation?

<!-- Discuss causal inference criteria -->

::: {.incremental}
- **Temporal precedence**: Cause before effect
- **Covariation**: They vary together
- **Elimination of alternatives**: No confounds
:::

::: {.fragment}
::: {.callout-tip}
Experimental designs (with random assignment) can establish causation. Correlational designs cannot.
:::
:::

## Reporting Correlations

<!-- APA format guide -->

::: {.example-box}
**APA format:**

"There was a strong positive correlation between study hours and exam scores, r(48) = .72, p < .001. Study time explained 52% of the variance in exam performance."
:::

# Summary {background-color="#275882"}

## Key Takeaways

::: {.incremental}
- **Correlation** measures strength and direction of linear relationships
- **Scatterplots** visualise relationships - always look first!
- **Pearson's r** for continuous, linear, normal data
- **Spearman's rho** for ordinal, non-linear, or non-normal data
- **Correlation NEVER proves causation** (third variables, directionality)
:::

## Next Week

<!-- Update with next week's topic -->
**Reading Week**


## Questions?

::: {.r-fit-text}
Questions?
:::

See you in the lab!
