---
title: "Week 12 Lab Activity"
subtitle: "Design Detectives"
author: "Gordon Wright"
date: "2026-01-23"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(webexercises)
library(tidyverse)
library(knitr)

# Load DataLab data
memory <- read_csv("../../_planning/DATALAB_MATERIALS/data/processed/memory_clean.csv", show_col_types = FALSE)
merged <- read_csv("../../_planning/DATALAB_MATERIALS/data/processed/foundations_merged.csv", show_col_types = FALSE)
frontallob <- read_csv("../../_planning/DATALAB_MATERIALS/data/processed/task5_frontallob_clean.csv", show_col_types = FALSE)
sixthsense <- read_csv("../../_planning/DATALAB_MATERIALS/data/processed/task3_sixthsense_clean.csv", show_col_types = FALSE)

# Brand colours
navy <- "#275882"
orange <- "#EA8439"

# Set theme
theme_set(theme_minimal(base_size = 14))
```

## Overview

::: {.callout-important}
## Focus for Today
This is NOT a heavy R coding session. Today focuses on **conceptual understanding**:

- Seeing your REAL data
- Identifying design types, IVs, and DVs
- Understanding what means and SDs tell us
- Writing hypotheses
:::

---

## Your Data Is Real

Last week you generated this data. Let's look at it!

### Memory Scores — Every Dot Is One Person

```{r}
#| label: memory-jitter
#| fig-height: 5
#| fig-align: center
memory_clean <- memory |> filter(!is.na(mem_total_1))

mean_score <- mean(memory_clean$mem_total_1, na.rm = TRUE)
sd_score <- sd(memory_clean$mem_total_1, na.rm = TRUE)

ggplot(memory_clean, aes(x = "", y = mem_total_1)) +
  geom_jitter(width = 0.3, size = 4, alpha = 0.7, colour = navy) +
  geom_hline(yintercept = mean_score, colour = orange, linewidth = 2, linetype = "solid") +
  geom_hline(yintercept = mean_score + sd_score, colour = orange, linewidth = 1, linetype = "dashed") +
  geom_hline(yintercept = mean_score - sd_score, colour = orange, linewidth = 1, linetype = "dashed") +
  annotate("text", x = 1.4, y = mean_score, label = paste0("Mean = ", round(mean_score, 1)),
           colour = orange, size = 5, hjust = 0) +
  annotate("text", x = 1.4, y = mean_score + sd_score, label = paste0("+1 SD"),
           colour = orange, size = 4, hjust = 0) +
  annotate("text", x = 1.4, y = mean_score - sd_score, label = paste0("-1 SD"),
           colour = orange, size = 4, hjust = 0) +
  labs(x = NULL, y = "Words Recalled (out of 10)",
       title = "Memory Scores",
       subtitle = paste0("n = ", nrow(memory_clean), " participants | SD = ", round(sd_score, 2))) +
  scale_y_continuous(breaks = 0:10, limits = c(0, 10)) +
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, colour = "gray50"))
```

**Can you spot yourself?** The solid orange line is the mean. The dashed lines show ±1 SD — about 68% of people fall between them.

---

### The Distribution — How Scores Spread Out

```{r}
#| label: memory-histogram
#| fig-height: 4
#| fig-align: center
ggplot(memory_clean, aes(x = mem_total_1)) +
  geom_histogram(binwidth = 1, fill = navy, colour = "white", alpha = 0.8) +
  geom_vline(xintercept = mean_score, colour = orange, linewidth = 2) +
  labs(x = "Words Recalled (out of 10)", y = "Number of People",
       title = "Distribution of Memory Scores",
       subtitle = paste0("Mean = ", round(mean_score, 1), " | SD = ", round(sd_score, 2),
                        " | Range = ", min(memory_clean$mem_total_1), "-", max(memory_clean$mem_total_1))) +
  scale_x_continuous(breaks = 0:10) +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, colour = "gray50"))
```

::: {.callout-note}
## Quick Check
Looking at the histogram:

Is the distribution roughly symmetric or skewed? `r mcq(c(answer = "Roughly symmetric", "Skewed left", "Skewed right", "Uniform"))`

Are most scores clustered near the mean or spread across the range? `r mcq(c(answer = "Clustered near the mean", "Spread evenly", "All at one end", "Bimodal"))`
:::

---

## Activity 1: The Big Reveal — Lab A vs Lab B (20 min)

### What You Didn't Know

There was a **hidden manipulation** in the memory task:

- **Lab A** memorised **abstract** words: justice, freedom, wisdom...
- **Lab B** memorised **concrete** words: chair, river, mountain...

Let's see if it made a difference!

```{r}
#| label: lab-comparison
#| fig-height: 5
#| fig-align: center
memory_labs <- memory |>
  filter(!is.na(Q11) & !is.na(mem_total_1)) |>
  mutate(Lab = factor(Q11, levels = c(1, 2), labels = c("Lab A\n(Abstract)", "Lab B\n(Concrete)")))

lab_stats <- memory_labs |>
  group_by(Lab) |>
  summarise(
    n = n(),
    Mean = mean(mem_total_1, na.rm = TRUE),
    SD = sd(mem_total_1, na.rm = TRUE),
    Min = min(mem_total_1, na.rm = TRUE),
    Max = max(mem_total_1, na.rm = TRUE)
  )

ggplot(memory_labs, aes(x = Lab, y = mem_total_1, fill = Lab)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.5) +
  geom_jitter(aes(colour = Lab), width = 0.15, size = 3, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, colour = "black") +
  scale_fill_manual(values = c(navy, orange)) +
  scale_colour_manual(values = c(navy, orange)) +
  labs(x = NULL, y = "Words Recalled (out of 10)",
       title = "Memory Performance by Lab",
       subtitle = "Black diamonds = group means") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, colour = "gray50"))
```

### The Numbers

```{r}
#| label: lab-table
kable(lab_stats, digits = 2, col.names = c("Lab", "n", "Mean", "SD", "Min", "Max"),
      caption = "Descriptive Statistics by Lab")
```

```{r}
diff_means <- abs(lab_stats$Mean[2] - lab_stats$Mean[1])
```

**The difference in means is `r round(diff_means, 2)` words.**

---

### Design Detective Questions

Now identify the design elements:

**1. What did the researcher MANIPULATE?**

`r mcq(c(answer = "Word type (abstract vs concrete)", "Number of words", "Time allowed", "Nothing was manipulated"))`

**2. What variable is this called?**

`r mcq(c(answer = "Independent Variable (IV)", "Dependent Variable (DV)", "Control Variable", "Confounding Variable"))`

**3. What did the researcher MEASURE?**

`r mcq(c("Word type", answer = "Number of words recalled", "Time to recall", "Participant happiness"))`

**4. What variable is this called?**

`r mcq(c("Independent Variable (IV)", answer = "Dependent Variable (DV)", "Control Variable", "Confounding Variable"))`

**5. Were the SAME people in both labs, or DIFFERENT people?**

`r mcq(c("Same people in both labs", answer = "Different people in each lab"))`

**6. What type of design is this?**

`r mcq(c(answer = "Between-subjects", "Within-subjects", "Correlational", "Matched pairs"))`

::: {.callout-tip collapse="true"}
## Check Your Understanding

- **IV** = Word type (abstract vs concrete) — this was manipulated
- **DV** = Words recalled — this was measured
- **Design** = Between-subjects — different people in Lab A vs Lab B
- You were in ONE lab only, not both!
:::

---

## Activity 2: Same Person, Different Conditions (20 min)

### Pre-Post Mood

You rated your mood BEFORE and AFTER the session. Let's see the change!

```{r}
#| label: mood-change
#| fig-height: 5
#| fig-align: center
mood_data <- merged |>
  select(participant_id, mood_pre_1, post_mood_post_1) |>
  filter(!is.na(mood_pre_1) & !is.na(post_mood_post_1)) |>
  mutate(change = post_mood_post_1 - mood_pre_1)

mood_long <- mood_data |>
  pivot_longer(cols = c(mood_pre_1, post_mood_post_1),
               names_to = "Time", values_to = "Mood") |>
  mutate(Time = factor(Time, levels = c("mood_pre_1", "post_mood_post_1"),
                       labels = c("Before", "After")))

mood_stats <- mood_long |>
  group_by(Time) |>
  summarise(Mean = mean(Mood), SD = sd(Mood), n = n())

ggplot(mood_long, aes(x = Time, y = Mood, fill = Time)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.5) +
  geom_jitter(aes(colour = Time), width = 0.1, size = 2, alpha = 0.6) +
  geom_line(data = mood_data |>
              pivot_longer(cols = c(mood_pre_1, post_mood_post_1),
                          names_to = "Time", values_to = "Mood") |>
              mutate(Time = factor(Time, levels = c("mood_pre_1", "post_mood_post_1"),
                                  labels = c("Before", "After"))),
            aes(group = participant_id), alpha = 0.2, colour = "gray40") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, colour = "black") +
  scale_fill_manual(values = c(navy, orange)) +
  scale_colour_manual(values = c(navy, orange)) +
  labs(x = NULL, y = "Mood Rating (1-10)",
       title = "Mood Before vs After the DataLab Session",
       subtitle = "Lines connect the same person | Black diamonds = means") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, colour = "gray50"))
```

**The faint lines connect the SAME PERSON.** This is what makes it a within-subjects design!

```{r}
#| label: mood-table
kable(mood_stats, digits = 2, col.names = c("Time", "Mean", "SD", "n"),
      caption = "Mood Ratings Before and After")
```

### Design Detective Questions

**1. Were SAME people measured at both time points, or DIFFERENT people?**

`r mcq(c(answer = "Same people at both time points", "Different people at each time point"))`

**2. What type of design is this?**

`r mcq(c("Between-subjects", answer = "Within-subjects", "Correlational", "Matched pairs"))`

**3. What is the IV?**

`r mcq(c(answer = "Time (before vs after)", "Mood rating", "Participant ID", "No IV"))`

**4. What is the DV?**

`r mcq(c("Time", answer = "Mood rating", "Participant ID", "Change score"))`

**5. Why is within-subjects design considered "powerful"?**

`r mcq(c("More participants", answer = "Each person is their own control", "Easier to run", "No IV needed"))`

::: {.callout-note}
## The Key Insight
In the graph above, the lines connect the same person before and after. By comparing each person to THEMSELVES, we control for all individual differences (age, personality, baseline mood, etc.).
:::

---

## Activity 3: Descriptive Statistics Practice (30 min)

### Understanding Mean and SD

Look at these two groups with the SAME MEAN but different spreads:

```{r}
#| label: spread-demo
#| fig-height: 4
#| fig-align: center
set.seed(42)
spread_data <- tibble(
  Group = rep(c("Group A\n(Clustered)", "Group B\n(Spread Out)"), each = 20),
  Score = c(rnorm(20, 6, 0.8), rnorm(20, 6, 2.5))
)

spread_stats <- spread_data |>
  group_by(Group) |>
  summarise(Mean = round(mean(Score), 2), SD = round(sd(Score), 2))

ggplot(spread_data, aes(x = Group, y = Score, colour = Group)) +
  geom_jitter(width = 0.15, size = 4, alpha = 0.8) +
  geom_hline(yintercept = 6, linetype = "dashed", colour = "gray40", linewidth = 1) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 6, colour = "black") +
  scale_colour_manual(values = c(navy, orange)) +
  annotate("text", x = 2.5, y = 6.2, label = "Same Mean ≈ 6", colour = "gray40", size = 5) +
  labs(y = "Score", x = NULL,
       title = "Same Mean, Different Spread",
       subtitle = paste0("Group A: SD = ", spread_stats$SD[1], " | Group B: SD = ", spread_stats$SD[2])) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, colour = "gray50"))
```

**The mean alone doesn't tell the whole story!**

**Which group shows more VARIABILITY?**

`r mcq(c("Group A (Clustered)", answer = "Group B (Spread Out)", "They're the same", "Can't tell"))`

**Which group has the LARGER standard deviation?**

`r mcq(c("Group A (Clustered)", answer = "Group B (Spread Out)", "They're the same", "Can't tell"))`

---

### Reading R Output

Here's what R produces when we ask for descriptive statistics:

```{r}
#| label: r-output-demo
#| echo: true
#| code-fold: false

# A sample of memory scores
scores <- c(6, 8, 5, 7, 9, 6, 7, 8)

# Get descriptive statistics
mean(scores)
sd(scores)
summary(scores)
```

**Reading the output above:**

**1. What is the mean of these scores?**

`r mcq(c("6", "6.5", answer = "7", "7.5"))`

**2. What is the standard deviation (rounded to 1 decimal)?**

`r mcq(c("0.5", "1.0", answer = "1.2", "2.0"))`

**3. In the `summary()` output, what does "Min." tell you?**

`r mcq(c(answer = "The lowest score in the data", "The average score", "The most common score", "The middle score"))`

**4. What does "Median" tell you?**

`r mcq(c("The lowest score", "The highest score", answer = "The middle value when data is ordered", "The most common score"))`

::: {.callout-tip}
## Reading R Output
The key functions for descriptive statistics:

- `mean()` gives you the average
- `sd()` gives you the standard deviation
- `summary()` gives you Min, Quartiles, Median, Mean, Max — a complete picture!
:::

True or False: A larger SD means scores are MORE spread out. `r torf(TRUE)`

True or False: A SD of 0 would mean everyone got the exact same score. `r torf(TRUE)`

---

### Real Data: Comparing Labs

```{r}
#| label: lab-stats-display
kable(lab_stats, digits = 2, col.names = c("Lab", "n", "Mean", "SD", "Min", "Max"),
      caption = "Memory Scores by Lab")
```

Looking at the table above:

**Which lab had the higher mean recall?**

`r mcq(c("Lab A (Abstract)", answer = "Lab B (Concrete)", "They were the same"))`
```{r}
higher_lab <- if(lab_stats$Mean[2] > lab_stats$Mean[1]) "Lab B" else "Lab A"
```

**Is the difference between means (about `r round(diff_means, 1)` words) large or small compared to the SD (about `r round(mean(lab_stats$SD), 1)`)?**

`r mcq(c(answer = "About the same size as the SD", "Much larger than the SD", "Much smaller than the SD"))`

::: {.callout-important}
## The Key Insight
A difference between means only matters if it's **big relative to the spread (SD)**.

- **Signal** = difference between means
- **Noise** = variability within groups (SD)

Is the signal bigger than the noise?
:::

---

## Activity 4: Hypothesis Writing Practice (20 min)

### Writing Hypotheses for the Memory Experiment

**The experimental hypothesis (H₁)** predicts there WILL be an effect.

**The null hypothesis (H₀)** predicts there will be NO effect.

For the Lab A vs Lab B memory comparison:

**Which is a correct EXPERIMENTAL hypothesis?**

`r mcq(c(answer = "Concrete words will be recalled better than abstract words", "There will be no difference between word types", "Word type cannot affect memory", "All participants will recall 5 words"))`

**Which is a correct NULL hypothesis?**

`r mcq(c("Concrete words will be recalled better", "Abstract words will be recalled better", answer = "There will be no difference in recall between word types", "Memory doesn't exist"))`

---

### One-tailed vs Two-tailed

**One-tailed**: Predicts the DIRECTION ("Group A will be HIGHER than Group B")

**Two-tailed**: Predicts a difference but NOT the direction ("Groups will DIFFER")

**"Concrete words will be easier to remember than abstract words" is:**

`r mcq(c(answer = "One-tailed (predicts direction)", "Two-tailed (no direction)", "Null hypothesis", "Not a hypothesis"))`

**"Word type will affect memory" is:**

`r mcq(c("One-tailed (predicts direction)", answer = "Two-tailed (no direction)", "Null hypothesis", "Not a hypothesis"))`

---

### Write Your Own

For the pre-post mood comparison, which hypothesis is correctly written?

**One-tailed hypothesis:**

`r mcq(c(answer = "Mood ratings will be higher after the session than before", "Mood ratings will change after the session", "There will be no change in mood", "Mood cannot be measured"))`

**Two-tailed hypothesis:**

`r mcq(c("Mood will improve after the session", answer = "Mood ratings will differ between before and after the session", "There will be no change in mood", "Mood will decrease"))`

---

## Activity 5: MCQ Practice (20 min)

::: {.webex-check .webex-box}

### Investigation 1: Throwing Accuracy

Students threw balls with their dominant hand (10 throws) and their non-dominant hand (5 throws). We measured accuracy (% successful throws).

**1. Design type:**

`r mcq(c("Between-subjects", answer = "Within-subjects", "Correlational"))`

**2. Why is this within-subjects?**

`r mcq(c("Different people used different hands", answer = "The same person threw with BOTH hands", "Hands were randomly assigned"))`

**3. What is the IV?**

`r mcq(c(answer = "Hand used (dominant vs non-dominant)", "Throwing accuracy", "Number of throws", "No IV"))`

**4. What is the DV?**

`r mcq(c("Hand used", answer = "Throwing accuracy (%)", "Number of throws", "Participant age"))`
:::

::: {.webex-check .webex-box}

### Investigation 2: Self-Rating vs Performance

Before throwing, students rated how good they thought they'd be (1-10). We then measured their actual accuracy.

**1. Design type:**

`r mcq(c("Between-subjects", "Within-subjects", answer = "Correlational"))`

**2. Is there an IV?**

`r mcq(c("Yes - self-rating", "Yes - actual accuracy", answer = "No - neither variable was manipulated"))`

**3. Why can't we say self-rating CAUSES good performance?**

`r mcq(c("Sample too small", "Wrong test", answer = "No manipulation means no causal claim"))`
:::

::: {.webex-check .webex-box}

### Investigation 3: Sixth Sense

Students tried to detect if someone was staring at them (6 trials). If just guessing, you'd expect 3/6 (50%) correct.

**1. If someone got 5/6 correct, does that PROVE they have a sixth sense?**

`r torf(FALSE)`

**2. Why not?**

`r mcq(c("5/6 is impossible", answer = "Some people will get 5/6 just by luck", "The experiment was flawed", "Sixth sense doesn't exist"))`

**3. What would 50% accuracy suggest?**

`r mcq(c("They definitely have a sixth sense", "They definitely don't have a sixth sense", answer = "Performance at chance level - just guessing"))`
:::

---

## Summary

::: {.callout-important}
## Key Takeaways

**Design Types:**

- **Between-subjects** = Different people in each condition (Lab A vs Lab B)
- **Within-subjects** = Same people in all conditions (Pre vs Post mood)
- **Correlational** = No manipulation, measuring relationships (Self-rating vs accuracy)

**Variables:**

- **IV** = What the researcher manipulates (word type, time)
- **DV** = What the researcher measures (recall, mood rating)
- Correlational designs have **NO IV**

**Descriptive Statistics:**

- **Mean** = Average value
- **SD** = How spread out values are
- A difference between means matters if it's **big relative to the SD**
:::

---

## Next Steps

1. Complete the [Week 12 Self-Assessment Quiz](/textbook/quizzes/w12-quiz.qmd)
2. Review the [Lecture Reading](/textbook/chapters/w12-lecture.qmd)
3. Start thinking about your **Research Plan** (due Week 16)!
