---
title: "Design Detectives: DataLab"
subtitle: "Week 12 Lab Activity"
author: "PS50008A Research Methods"
date: "2026-01-22"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Tasks"
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(webexercises)
library(tidyverse)
library(knitr)
library(scales)

# Load DataLab data
data_path <- "../../DATALAB_MATERIALS/data/processed/"

memory <- read_csv(paste0(data_path, "memory_clean.csv"), show_col_types = FALSE)
reflex <- read_csv(paste0(data_path, "task1_reflex_clean.csv"), show_col_types = FALSE)
timewarp <- read_csv(paste0(data_path, "task2_timewarp_clean.csv"), show_col_types = FALSE)
sixthsense <- read_csv(paste0(data_path, "task3_sixthsense_clean.csv"), show_col_types = FALSE)
ladyluck <- read_csv(paste0(data_path, "task4_ladyluck_clean.csv"), show_col_types = FALSE)
frontallob <- read_csv(paste0(data_path, "task5_frontallob_clean.csv"), show_col_types = FALSE)
merged <- read_csv(paste0(data_path, "foundations_merged.csv"), show_col_types = FALSE)

# Brand colours
navy <- "#275882"
orange <- "#EA8439"
teal <- "#2A9D8F"
coral <- "#E76F51"

theme_set(theme_minimal(base_size = 14))
```

::: {.callout-note}
## How This Works

This activity accompanies the **instructor slides**. Your instructor will guide you through each phase:

1. **Task overview** (instructor presents)
2. **Group discussion** (you work through questions below)
3. **Whole-class debrief** (we discuss answers together)

Form groups of **2-3 people** and work through each section when prompted.
:::

::: {.callout-tip}
## Glossary Links
Key terms are linked to the [glossary](/glossary/). If you're unsure about terminology, click the links to review definitions.
:::

---

## Task 1: Reflex Test {.unnumbered}

### What You Did

You caught a falling 30cm ruler five times. The distance you caught it at reflects your reaction time - shorter distance means faster reactions.

### Your Results

```{r}
#| label: reflex-plot
#| fig-height: 4
#| fig-align: center

reflex_cols <- names(reflex)[grep("trial", names(reflex), ignore.case = TRUE)]

if(length(reflex_cols) >= 5) {
  reflex_long <- reflex |>
    select(participant_id, all_of(reflex_cols[1:5])) |>
    pivot_longer(cols = -participant_id, names_to = "trial", values_to = "distance") |>
    filter(!is.na(distance) & distance > 0 & distance <= 30) |>
    mutate(trial_num = as.numeric(str_extract(trial, "\\d+")))
  
  trial_means <- reflex_long |>
    group_by(trial_num) |>
    summarise(mean_dist = mean(distance, na.rm = TRUE), se = sd(distance, na.rm = TRUE)/sqrt(n()))
  
  ggplot(trial_means, aes(x = trial_num, y = mean_dist)) +
    geom_line(colour = coral, linewidth = 1.5) +
    geom_point(colour = coral, size = 4) +
    geom_ribbon(aes(ymin = mean_dist - se, ymax = mean_dist + se), alpha = 0.2, fill = coral) +
    labs(x = "Trial", y = "Mean Distance (cm)", title = "Performance Across Trials") +
    scale_x_continuous(breaks = 1:5) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim <- tibble(trial = 1:5, mean_dist = c(22, 19, 16, 14, 13))
  ggplot(sim, aes(x = trial, y = mean_dist)) +
    geom_line(colour = coral, linewidth = 1.5) + geom_point(colour = coral, size = 4) +
    labs(x = "Trial", y = "Mean Distance (cm)", title = "Performance Across Trials")
}
```

### Group Discussion Questions

**Q1.1** Look at the graph. What happens to performance from Trial 1 to Trial 5?

`r mcq(c("Performance gets worse", answer = "Performance improves", "Performance stays the same", "Performance is random"))`

**Q1.2** Were the **same people** tested in all 5 trials, or **different people** in each trial?

`r mcq(c("Different people in each trial", answer = "Same people in all trials"))`

**Q1.3** What design type is this? (See [within-subjects design](/glossary/design/within-groups-design.qmd))

`r mcq(c("Between-subjects", answer = "Within-subjects", "Correlational", "Matched pairs"))`

**Q1.4** What is the [**Independent Variable (IV)**](/glossary/design/independent-variable.qmd)?

`r mcq(c(answer = "Trial number (1, 2, 3, 4, 5)", "Distance caught", "Reaction time", "Nothing - no IV"))`

**Q1.5** What is the [**Dependent Variable (DV)**](/glossary/design/dependent-variable.qmd)?

`r mcq(c("Trial number", answer = "Distance caught (or reaction time)", "The participant", "The ruler"))`

**Q1.6** Write a **one-tailed hypothesis** for this experiment:

::: {.callout-tip collapse="true"}
## Example Answer
"Reaction time will decrease (improve) from Trial 1 to Trial 5" OR "Catch distance will decrease from Trial 1 to Trial 5"

This is a **practice effect** - performance improves with repetition.
:::

---

## Task 2: Time Warp {.unnumbered}

### What You Did

You estimated 60 seconds without counting. We wanted to see if people systematically over- or underestimate time.

### Your Results

```{r}
#| label: time-plot
#| fig-height: 4
#| fig-align: center

time_col <- names(timewarp)[grep("estimate|time", names(timewarp), ignore.case = TRUE)]

if(length(time_col) > 0) {
  time_data <- timewarp |>
    select(participant_id, estimate = !!sym(time_col[1])) |>
    filter(!is.na(estimate) & estimate > 0 & estimate < 180)
  
  mean_est <- mean(time_data$estimate, na.rm = TRUE)
  
  ggplot(time_data, aes(x = estimate)) +
    geom_histogram(binwidth = 5, fill = teal, colour = "white", alpha = 0.8) +
    geom_vline(xintercept = 60, linetype = "dashed", colour = coral, linewidth = 1.5) +
    geom_vline(xintercept = mean_est, colour = navy, linewidth = 1.5) +
    annotate("text", x = 62, y = Inf, vjust = 2, label = "Target: 60s", colour = coral, fontface = "bold", hjust = 0) +
    annotate("text", x = mean_est + 2, y = Inf, vjust = 4, label = paste0("Mean: ", round(mean_est,0), "s"), colour = navy, fontface = "bold", hjust = 0) +
    labs(x = "Your Estimate (seconds)", y = "Count", title = "Time Estimation") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(123)
  sim_time <- tibble(estimate = rnorm(30, 55, 12))
  ggplot(sim_time, aes(x = estimate)) +
    geom_histogram(binwidth = 5, fill = teal, colour = "white") +
    geom_vline(xintercept = 60, linetype = "dashed", colour = coral, linewidth = 1.5) +
    labs(x = "Estimate (seconds)", y = "Count", title = "Time Estimation")
}
```

### Group Discussion Questions

**Q2.1** On average, did people over- or underestimate 60 seconds?

`r mcq(c("Overestimate (said 'stop' too late)", answer = "Underestimate (said 'stop' too early)", "Perfectly accurate"))`

**Q2.2** What does it mean if someone underestimates 60 seconds?

`r mcq(c(answer = "Time felt like it passed quickly (they thought 60s was up before it actually was)", "Time felt like it passed slowly", "They were counting accurately"))`

**Q2.3** This experiment compares people's estimates to a known target (60 seconds). The [**null hypothesis**](/glossary/statistics/) would be:

`r mcq(c(answer = "Mean estimate will equal 60 seconds (no systematic bias)", "Mean estimate will be less than 60 seconds", "Mean estimate will be greater than 60 seconds", "There is no null hypothesis"))`

---

## Task 3: Memory {.unnumbered}

### The Secret Manipulation

Here's what you didn't know: **Lab A** and **Lab B** saw different words!

- **Lab A**: Abstract words (justice, freedom, wisdom...)
- **Lab B**: Concrete words (chair, river, mountain...)

### Your Results

```{r}
#| label: memory-plot
#| fig-height: 4.5
#| fig-align: center

lab_col <- names(memory)[grep("lab|room|Q11", names(memory), ignore.case = TRUE)]
total_col <- names(memory)[grep("total|mem_total", names(memory), ignore.case = TRUE)]

if(length(lab_col) > 0 && length(total_col) > 0) {
  memory_labs <- memory |>
    select(participant_id, lab = !!sym(lab_col[1]), total = !!sym(total_col[1])) |>
    filter(!is.na(lab) & !is.na(total)) |>
    mutate(Lab = factor(lab, labels = c("Lab A (Abstract)", "Lab B (Concrete)")))
  
  lab_stats <- memory_labs |>
    group_by(Lab) |>
    summarise(n = n(), Mean = round(mean(total), 2), SD = round(sd(total), 2))
  
  ggplot(memory_labs, aes(x = Lab, y = total, fill = Lab)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
    geom_jitter(width = 0.15, size = 3, alpha = 0.7, colour = "gray30") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 5, colour = "black") +
    scale_fill_manual(values = c(navy, orange)) +
    labs(x = NULL, y = "Words Recalled (out of 10)", title = "Memory Performance by Lab", subtitle = "Black diamonds = group means") +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim_mem <- tibble(Lab = rep(c("Lab A (Abstract)", "Lab B (Concrete)"), each = 16), total = c(rnorm(16, 5.8, 1.5), rnorm(16, 7.2, 1.4)))
  ggplot(sim_mem, aes(x = Lab, y = total, fill = Lab)) +
    geom_boxplot(alpha = 0.6) + scale_fill_manual(values = c(navy, orange)) +
    labs(y = "Words Recalled", title = "Memory Performance by Lab") + theme(legend.position = "none")
}
```

### Descriptive Statistics

```{r}
#| label: memory-table
if(exists("lab_stats")) {
  kable(lab_stats, col.names = c("Lab", "n", "Mean", "SD"), caption = "Memory scores by lab group")
} else {
  tibble(Lab = c("Lab A (Abstract)", "Lab B (Concrete)"), n = c(16, 16), Mean = c(5.81, 7.19), SD = c(1.52, 1.38)) |>
    kable(caption = "Memory scores by lab group")
}
```

### Group Discussion Questions

**Q3.1** What is the [**Independent Variable (IV)**](/glossary/design/independent-variable.qmd)?

`r mcq(c(answer = "Word type (abstract vs concrete)", "Number of words recalled", "The participant", "Memory ability"))`

**Q3.2** What is the [**Dependent Variable (DV)**](/glossary/design/dependent-variable.qmd)?

`r mcq(c("Word type", answer = "Number of words recalled", "Lab room", "Time to recall"))`

**Q3.3** Were the same people in both labs?

`r mcq(c("Yes, same people saw both word types", answer = "No, different people in each lab"))`

**Q3.4** What design type is this? (See [between-subjects design](/glossary/design/between-groups-design.qmd))

`r mcq(c(answer = "Between-subjects", "Within-subjects", "Correlational", "Matched pairs"))`

**Q3.5** Write an **experimental hypothesis (H1)** - one-tailed:

`r mcq(c(answer = "Concrete words will be recalled better than abstract words", "There will be no difference between word types", "Abstract words will be recalled better", "Word type will have some effect"))`

**Q3.6** Write a **null hypothesis (H0)**:

`r mcq(c("Concrete words will be easier to remember", "Abstract words will be easier to remember", answer = "There will be no difference in recall between word types", "Memory is always the same"))`

**Q3.7** Why might concrete words be easier to remember?

::: {.callout-tip collapse="true"}
## The Answer: Dual Coding Theory
Concrete words can be stored in TWO ways: as a verbal code (the word) AND as a visual image (you can picture "chair"). Abstract words only have the verbal code. Two codes = better memory.
:::

---

## Task 4: Frontal Lob {.unnumbered}

### What You Did

You threw balls at a bucket:

- 10 throws with your **dominant hand**
- 5 throws with your **non-dominant hand**

### Your Results

```{r}
#| label: throwing-plot
#| fig-height: 4.5
#| fig-align: center

dom_col <- names(frontallob)[grep("dom.*success|dom.*hit", names(frontallob), ignore.case = TRUE)]
nondom_col <- names(frontallob)[grep("non.*success|non.*hit", names(frontallob), ignore.case = TRUE)]

if(length(dom_col) > 0 && length(nondom_col) > 0) {
  throw_data <- frontallob |>
    select(participant_id, dom = !!sym(dom_col[1]), nondom = !!sym(nondom_col[1])) |>
    filter(!is.na(dom) & !is.na(nondom)) |>
    mutate(dom_pct = dom / 10 * 100, nondom_pct = nondom / 5 * 100)
  
  throw_long <- throw_data |>
    pivot_longer(cols = c(dom_pct, nondom_pct), names_to = "hand", values_to = "accuracy") |>
    mutate(hand = factor(hand, levels = c("dom_pct", "nondom_pct"), labels = c("Dominant", "Non-Dominant")))
  
  throw_stats <- throw_long |>
    group_by(hand) |>
    summarise(n = n(), Mean = round(mean(accuracy), 1), SD = round(sd(accuracy), 1))
  
  ggplot(throw_long, aes(x = hand, y = accuracy, fill = hand)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    scale_fill_manual(values = c(navy, orange)) +
    labs(x = NULL, y = "Accuracy (%)", title = "Throwing Accuracy by Hand") +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim_throw <- tibble(hand = rep(c("Dominant", "Non-Dominant"), each = 25), accuracy = c(rnorm(25, 78, 12), rnorm(25, 60, 15)))
  ggplot(sim_throw, aes(x = hand, y = accuracy, fill = hand)) +
    geom_boxplot(alpha = 0.6) + scale_fill_manual(values = c(navy, orange)) +
    labs(y = "Accuracy (%)", title = "Throwing Accuracy by Hand") + theme(legend.position = "none")
}
```

### Group Discussion Questions

**Q4.1** Were the same people tested with both hands?

`r mcq(c("No, different people used different hands", answer = "Yes, same people used both hands"))`

**Q4.2** What design type is this? (See [within-subjects design](/glossary/design/within-groups-design.qmd))

`r mcq(c("Between-subjects", answer = "Within-subjects", "Correlational", "Matched pairs"))`

**Q4.3** What is the [**IV**](/glossary/design/independent-variable.qmd)?

`r mcq(c(answer = "Hand used (dominant vs non-dominant)", "Accuracy", "Number of throws", "The participant"))`

**Q4.4** What is the [**DV**](/glossary/design/dependent-variable.qmd)?

`r mcq(c("Hand used", answer = "Throwing accuracy (%)", "Number of attempts", "The bucket"))`

**Q4.5** Why is within-subjects design considered "more powerful"?

`r mcq(c("You need fewer participants", "The statistics are simpler", answer = "Each person is their own control - individual differences are eliminated", "The results are always significant"))`

**Q4.6** Which statistical test would you use to compare dominant vs non-dominant hand?

`r mcq(c("Independent t-test", answer = "Paired t-test", "Correlation", "Chi-square"))`

---

## Task 5: Sleep & Memory {.unnumbered}

### The Question

Before the session, you told us how many hours you slept last night. We also measured how many words you recalled.

**Is there a relationship between sleep and memory?**

### Your Results

```{r}
#| label: sleep-memory-plot
#| fig-height: 4.5
#| fig-align: center

# Get sleep and memory data from merged
sleep_mem <- merged |>
  select(participant_id, sleep_hours, memory_mem_total_1) |>
  filter(!is.na(sleep_hours) & !is.na(memory_mem_total_1) & 
         sleep_hours > 0 & sleep_hours < 15)

if(nrow(sleep_mem) > 5) {
  cor_val <- cor(sleep_mem$sleep_hours, sleep_mem$memory_mem_total_1, use = "complete.obs")
  
  ggplot(sleep_mem, aes(x = sleep_hours, y = memory_mem_total_1)) +
    geom_point(size = 4, colour = navy, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, colour = teal, fill = teal, alpha = 0.2) +
    labs(x = "Hours of Sleep Last Night", 
         y = "Words Recalled (out of 10)",
         title = "Sleep and Memory Performance",
         subtitle = paste0("Correlation: r = ", round(cor_val, 2))) +
    scale_x_continuous(breaks = seq(4, 12, 2)) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim_sleep <- tibble(
    sleep = runif(28, 5, 9),
    memory = 4 + 0.4 * sleep + rnorm(28, 0, 1.2)
  )
  cor_val <- round(cor(sim_sleep$sleep, sim_sleep$memory), 2)
  ggplot(sim_sleep, aes(x = sleep, y = memory)) +
    geom_point(size = 4, colour = navy, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, colour = teal, fill = teal, alpha = 0.2) +
    labs(x = "Hours of Sleep", y = "Words Recalled", 
         title = "Sleep and Memory Performance",
         subtitle = paste0("Correlation: r = ", cor_val))
}
```

### Group Discussion Questions

**Q5.1** Is there an IV in this study?

`r mcq(c("Yes - sleep hours", "Yes - memory score", answer = "No - we measured both variables, we didn't manipulate either", "Yes - the participant"))`

**Q5.2** What design type is this? (See [correlational design](/glossary/design/correlational-design.qmd))

`r mcq(c("Between-subjects", "Within-subjects", answer = "Correlational", "Experimental"))`

**Q5.3** What does a positive correlation mean here?

`r mcq(c("Sleep causes better memory", answer = "People who slept more tended to recall more words", "People who slept less recalled more words", "There is no relationship"))`

**Q5.4** Can we say that sleep **causes** better memory based on this data?

`r torf(FALSE)`

**Q5.5** Why can't we make causal claims from correlational data?

`r mcq(c("The sample was too small", "The correlation wasn't strong enough", answer = "Third variables might explain the relationship (e.g., health, motivation)", "Correlations are always meaningless"))`

::: {.callout-warning}
## The Critical Point: Correlation ≠ Causation

Even if we find a relationship between sleep and memory, we **cannot** say sleep *causes* better memory. Why?

- Maybe healthier people sleep more AND have better memory
- Maybe motivated students prepare better (sleep AND study)
- Maybe stress reduces both sleep AND memory

This is why **experiments** (with manipulation) are needed to establish causation.
:::

---

## Task 6: Sixth Sense {.unnumbered}

### What You Did

6 trials: Someone either stared at you or didn't. You guessed which.

**If just guessing randomly**: You'd expect to get 3 out of 6 correct (50%).

### Your Results

```{r}
#| label: sixth-sense-plot
#| fig-height: 4
#| fig-align: center

# Calculate accuracy from actual vs guess columns
# Data has t1_actual, t1_guess, t2_actual, t2_guess, etc.
acc_data <- sixthsense |>
  mutate(
    correct_1 = as.numeric(t1_actual == t1_guess),
    correct_2 = as.numeric(t2_actual == t2_guess),
    correct_3 = as.numeric(t3_actual == t3_guess),
    correct_4 = as.numeric(t4_actual == t4_guess),
    correct_5 = as.numeric(t5_actual == t5_guess),
    correct_6 = as.numeric(t6_actual == t6_guess),
    total_correct = correct_1 + correct_2 + correct_3 + correct_4 + correct_5 + correct_6,
    accuracy = total_correct / 6
  ) |>
  select(participant_id, accuracy) |>
  filter(!is.na(accuracy))

if(nrow(acc_data) > 0) {
  ggplot(acc_data, aes(x = accuracy)) +
    geom_histogram(binwidth = 1/6, fill = "#9b59b6", colour = "white", alpha = 0.8) +
    geom_vline(xintercept = 0.5, linetype = "dashed", colour = coral, linewidth = 1.5) +
    annotate("text", x = 0.52, y = Inf, vjust = 2, label = "Chance (50%)", colour = coral, fontface = "bold", hjust = 0) +
    scale_x_continuous(labels = percent, limits = c(0, 1)) +
    labs(x = "Accuracy", y = "Count", title = "Sixth Sense Detection Accuracy") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim_sixth <- tibble(accuracy = rbinom(28, 6, 0.52) / 6)
  ggplot(sim_sixth, aes(x = accuracy)) +
    geom_histogram(binwidth = 1/6, fill = "#9b59b6", colour = "white") +
    geom_vline(xintercept = 0.5, linetype = "dashed", colour = coral, linewidth = 1.5) +
    scale_x_continuous(labels = percent) +
    labs(x = "Accuracy", y = "Count", title = "Sixth Sense Detection Accuracy")
}
```

### Group Discussion Questions

**Q6.1** If someone got 5 out of 6 correct, does that **prove** they have a sixth sense?

`r torf(FALSE)`

**Q6.2** Why not?

`r mcq(c("5/6 is impossible", answer = "Some people will get 5/6 just by luck, even without any real ability", "The experiment was designed wrong", "Sixth sense definitely doesn't exist"))`

**Q6.3** What does 50% accuracy suggest?

`r mcq(c("They definitely have a sixth sense", "They definitely don't have a sixth sense", answer = "Performance is at chance level - consistent with random guessing", "The task was too hard"))`

**Q6.4** What is the **null hypothesis** for this study?

`r mcq(c("People can detect staring", "People cannot detect staring", answer = "Detection accuracy equals 50% (chance level)", "There is no null hypothesis"))`

::: {.callout-tip collapse="true"}
## Why Statistics Matter

With 28 people, **some will get lucky** just by chance. If everyone flipped a coin 6 times, someone would get 5 or 6 heads - but that doesn't mean their coin is biased!

**Statistics help us ask**: Is this result more extreme than we'd expect from chance alone?
:::

---

## Summary {.unnumbered}

::: {.callout-important}
## Key Concepts from Today

**Design Types** (See [glossary](/glossary/design/))

| Type | Same people? | IV? | Test to use | Example |
|------|-------------|-----|-------------|---------|
| [Between-subjects](/glossary/design/between-groups-design.qmd) | No | Yes | Independent t-test | Memory (Lab A vs B) |
| [Within-subjects](/glossary/design/within-groups-design.qmd) | Yes | Yes | Paired t-test | Throwing (Dom vs Non-dom) |
| [Correlational](/glossary/design/correlational-design.qmd) | N/A | No | Correlation | Sleep & Memory |

**Variables**

- [**IV (Independent Variable)**](/glossary/design/independent-variable.qmd): What the researcher manipulates or controls
- [**DV (Dependent Variable)**](/glossary/design/dependent-variable.qmd): What the researcher measures

**Hypotheses**

- **Experimental (H1)**: Predicts an effect or difference
- **Null (H0)**: Predicts no effect (any difference is due to chance)
- **One-tailed**: Specifies the direction ("A will be greater than B")
- **Two-tailed**: Just predicts a difference ("A and B will differ")
:::

---

## Next Steps {.unnumbered}

1. Complete the **Week 12 Self-Assessment Quiz** to test your understanding

2. Review the **Lecture Reading** for more detail on these concepts

3. Start thinking about your **Research Plan** (due Week 16) - you now have all the vocabulary you need!

---

::: {.callout-note collapse="true"}
## Check Your Progress

How many questions did you get correct? Use this as a guide:

- **75-100%**: Great understanding - ready for the self-assessment
- **50-75%**: Good start - review the concepts you missed
- **Below 50%**: Spend time with the lecture reading before the quiz

**Remember**: This is formative - mistakes are learning opportunities!
:::
