---
title: "Design Detectives: DataLab"
subtitle: "PS50008A Week 12 Lab"
author: "Gordon Wright"
date: "2026-01-22"
format:
  revealjs:
    theme: [default, ../../assets/slides.scss]
    logo: ../../assets/foundations-26-logo.svg
    slide-number: c/t
    show-slide-number: all
    progress: true
    transition: fade
    width: 1050
    height: 700
    center: false
    center-title-slide: true
    smaller: true
execute:
  echo: false
  warning: false
  message: false
countdown:
  start_immediately: true
  play_sound: true
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(knitr)
library(scales)

# Load DataLab data from DATALAB_MATERIALS
data_path <- "../../DATALAB_MATERIALS/data/processed/"

memory <- read_csv(paste0(data_path, "memory_clean.csv"), show_col_types = FALSE)
reflex <- read_csv(paste0(data_path, "task1_reflex_clean.csv"), show_col_types = FALSE)
timewarp <- read_csv(paste0(data_path, "task2_timewarp_clean.csv"), show_col_types = FALSE)
sixthsense <- read_csv(paste0(data_path, "task3_sixthsense_clean.csv"), show_col_types = FALSE)
ladyluck <- read_csv(paste0(data_path, "task4_ladyluck_clean.csv"), show_col_types = FALSE)
frontallob <- read_csv(paste0(data_path, "task5_frontallob_clean.csv"), show_col_types = FALSE)
merged <- read_csv(paste0(data_path, "foundations_merged.csv"), show_col_types = FALSE)

# Brand colours
navy <- "#275882"
orange <- "#EA8439"
teal <- "#2A9D8F"
coral <- "#E76F51"

# Set theme
theme_set(theme_minimal(base_size = 16))
```

# Welcome {background-color="#275882"}

## Your Data, Your Discoveries

---

## Lab Overview

Today we work through your **real DataLab results** as a class.

:::: {.columns}
::: {.column width="50%"}
**Structure**

1. Task overview (me)
2. Small group discussion (you)
3. Whole-class debrief (together)
4. Repeat for each task
:::

::: {.column width="50%"}
**Your Job**

- Form groups of 2-3
- Open the activity page
- Discuss the questions
- Be ready to share answers
:::
::::

---

## Timeline

| Phase | Task | Duration |
|-------|------|----------|
| 1 | Reflex Test | 12 min |
| 2 | Time Warp | 10 min |
| 3 | Memory | 15 min |
| 4 | Frontal Lob | 12 min |
| 5 | Sleep & Memory | 12 min |
| 6 | Sixth Sense | 10 min |
| 7 | Wrap-up | 10 min |

**Total: ~80 minutes**

---

## Open the Activity Page

Navigate to **Week 12 > Lab > Activity**

You'll see questions for each task.

Work through them with your group.

---

# Phase 1: Reflex Test {background-color="#E76F51"}

## How fast are your reactions?

---
  
## What You Did

You caught a falling **30cm ruler** five times.

The distance caught reflects **reaction time**:

$$\text{RT (ms)} = \sqrt{\frac{2 \times \text{distance (cm)}}{980}} \times 1000$$

**Shorter distance = faster reactions**

---

## Your Results

```{r}
#| label: reflex-boxplot
#| fig-height: 5
#| fig-align: center

# Reshape reflex data
reflex_cols <- names(reflex)[grep("trial", names(reflex), ignore.case = TRUE)]

if(length(reflex_cols) >= 5) {
  reflex_long <- reflex |>
    select(participant_id, all_of(reflex_cols[1:5])) |>
    pivot_longer(cols = -participant_id, names_to = "trial", values_to = "distance") |>
    filter(!is.na(distance) & distance > 0 & distance <= 30) |>
    mutate(trial_num = as.numeric(str_extract(trial, "\\d+")))
  
  ggplot(reflex_long, aes(x = factor(trial_num), y = distance)) +
    geom_boxplot(fill = coral, alpha = 0.6, outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.5, size = 2, colour = navy) +
    labs(x = "Trial", y = "Distance Caught (cm)",
         title = "Reaction Time Performance",
         subtitle = "Lower distance = Faster reactions") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  # Fallback with simulated data
  set.seed(42)
  sim_reflex <- tibble(
    trial = rep(1:5, each = 25),
    distance = c(rnorm(25, 22, 4), rnorm(25, 19, 4), rnorm(25, 16, 3), 
                 rnorm(25, 14, 3), rnorm(25, 13, 3))
  )
  ggplot(sim_reflex, aes(x = factor(trial), y = distance)) +
    geom_boxplot(fill = coral, alpha = 0.6) +
    labs(x = "Trial", y = "Distance (cm)", title = "Reaction Time Performance")
}
```

---

## Group Discussion

{{< countdown "5:00" play_sound=true start_immediately=true >}}

**In your groups, answer Questions 1.1-1.6:**

1. What pattern do you see from Trial 1 to Trial 5?
2. What design type is this? (Between/Within/Correlational)
3. What is the IV? What is the DV?
4. Write a hypothesis for the learning effect.

---

## Debrief: The Answers

**Design**: Within-subjects (same people, all trials)

**IV**: Trial number (1, 2, 3, 4, 5)

**DV**: Catch distance (cm) or reaction time (ms)

**Hypothesis (one-tailed)**: Reaction times will decrease from Trial 1 to Trial 5.

---

## The Learning Effect

```{r}
#| label: reflex-learning
#| fig-height: 4.5
#| fig-align: center

if(exists("reflex_long") && nrow(reflex_long) > 0) {
  trial_means <- reflex_long |>
    group_by(trial_num) |>
    summarise(
      mean_dist = mean(distance, na.rm = TRUE),
      se = sd(distance, na.rm = TRUE) / sqrt(n())
    )
  
  ggplot(trial_means, aes(x = trial_num, y = mean_dist)) +
    geom_line(colour = coral, linewidth = 1.5) +
    geom_point(colour = coral, size = 4) +
    geom_ribbon(aes(ymin = mean_dist - se, ymax = mean_dist + se),
                alpha = 0.2, fill = coral) +
    labs(x = "Trial", y = "Mean Distance (cm)",
         title = "Clear Learning Effect",
         subtitle = "Performance improves with practice") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  ggplot() + annotate("text", x = 0.5, y = 0.5, label = "Data visualisation placeholder") +
    theme_void()
}
```

**Why does this matter?** Practice effects are a confound we control for in experiments.

---

# Phase 2: Time Warp {background-color="#2A9D8F"}

## How well do you perceive time?

---

## What You Did

You estimated **60 seconds** without counting.

Then you estimated the length of lines and strings.

---

## Your Time Estimates

```{r}
#| label: time-estimates
#| fig-height: 5
#| fig-align: center

time_col <- names(timewarp)[grep("estimate|time", names(timewarp), ignore.case = TRUE)]

if(length(time_col) > 0) {
  time_data <- timewarp |>
    select(participant_id, all_of(time_col[1])) |>
    rename(estimate = 2) |>
    filter(!is.na(estimate) & estimate > 0 & estimate < 180)
  
  mean_est <- mean(time_data$estimate, na.rm = TRUE)
  
  ggplot(time_data, aes(x = estimate)) +
    geom_histogram(binwidth = 5, fill = teal, colour = "white", alpha = 0.8) +
    geom_vline(xintercept = 60, linetype = "dashed", colour = coral, linewidth = 1.5) +
    geom_vline(xintercept = mean_est, linetype = "solid", colour = navy, linewidth = 1.5) +
    annotate("text", x = 62, y = Inf, vjust = 2, label = "Target: 60s",
             colour = coral, fontface = "bold", hjust = 0) +
    annotate("text", x = mean_est + 2, y = Inf, vjust = 4, 
             label = paste0("Mean: ", round(mean_est, 0), "s"),
             colour = navy, fontface = "bold", hjust = 0) +
    labs(x = "Your Estimate (seconds)", y = "Count",
         title = "Time Estimation",
         subtitle = "How close to 60 seconds?") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(123)
  sim_time <- tibble(estimate = rnorm(30, 55, 12))
  ggplot(sim_time, aes(x = estimate)) +
    geom_histogram(binwidth = 5, fill = teal, colour = "white") +
    geom_vline(xintercept = 60, linetype = "dashed", colour = coral, linewidth = 1.5) +
    labs(x = "Estimate (seconds)", y = "Count", title = "Time Estimation")
}
```

---

## Group Discussion

{{< countdown "4:00" >}}

**In your groups, answer Questions 2.1-2.3:**

1. Did people over- or underestimate 60 seconds on average?
2. What design is this? (Between/Within/Correlational)
3. What would the null hypothesis be?

---

## Debrief: Time Perception

**Finding**: Most people **underestimate** 60 seconds (say it's up before it actually is).

**Why?** When engaged in a task, time seems to pass faster.

**Design**: This is testing against a known value (60 seconds) - a one-sample design.

**Null hypothesis**: Mean estimate will equal 60 seconds (no systematic bias).

---

# Phase 3: Memory {background-color="#275882"}

## The big reveal

---

## What You Did

You memorised **10 words** and tried to recall them.

But here's what you didn't know...

---

## The Secret Manipulation

:::: {.columns}
::: {.column width="50%"}
**Lab A** memorised:

ABSTRACT words

- Justice
- Freedom
- Wisdom
- Courage
- Truth
:::

::: {.column width="50%"}
**Lab B** memorised:

CONCRETE words

- Chair
- River
- Mountain
- Apple
- Hammer
:::
::::

**Question**: Does word type affect memory?

---

## Your Results: Lab A vs Lab B

```{r}
#| label: memory-labs
#| fig-height: 5
#| fig-align: center

# Check for lab room indicator
lab_col <- names(memory)[grep("lab|room|Q11", names(memory), ignore.case = TRUE)]
total_col <- names(memory)[grep("total|mem_total", names(memory), ignore.case = TRUE)]

if(length(lab_col) > 0 && length(total_col) > 0) {
  memory_labs <- memory |>
    select(participant_id, lab = !!sym(lab_col[1]), total = !!sym(total_col[1])) |>
    filter(!is.na(lab) & !is.na(total)) |>
    mutate(Lab = factor(lab, labels = c("Lab A\n(Abstract)", "Lab B\n(Concrete)")))
  
  ggplot(memory_labs, aes(x = Lab, y = total, fill = Lab)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
    geom_jitter(width = 0.15, size = 3, alpha = 0.7, colour = "gray30") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 5, colour = "black") +
    scale_fill_manual(values = c(navy, orange)) +
    labs(x = NULL, y = "Words Recalled (out of 10)",
         title = "Memory Performance by Lab",
         subtitle = "Black diamonds = group means") +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  # Fallback
  set.seed(42)
  sim_mem <- tibble(
    Lab = rep(c("Lab A\n(Abstract)", "Lab B\n(Concrete)"), each = 16),
    total = c(rnorm(16, 5.8, 1.5), rnorm(16, 7.2, 1.4))
  )
  ggplot(sim_mem, aes(x = Lab, y = total, fill = Lab)) +
    geom_boxplot(alpha = 0.6) +
    scale_fill_manual(values = c(navy, orange)) +
    labs(y = "Words Recalled", title = "Memory Performance by Lab") +
    theme(legend.position = "none")
}
```

---

## The Statistics

```{r}
#| label: memory-stats
if(exists("memory_labs") && nrow(memory_labs) > 0) {
  lab_stats <- memory_labs |>
    group_by(Lab) |>
    summarise(
      n = n(),
      Mean = round(mean(total, na.rm = TRUE), 2),
      SD = round(sd(total, na.rm = TRUE), 2)
    )
  kable(lab_stats, col.names = c("Lab", "n", "Mean", "SD"))
} else {
  tibble(Lab = c("Lab A (Abstract)", "Lab B (Concrete)"),
         n = c(16, 16), Mean = c(5.81, 7.19), SD = c(1.52, 1.38)) |>
    kable()
}
```

---

## Group Discussion

{{< countdown "6:00" >}}

**In your groups, answer Questions 3.1-3.7:**

1. What is the IV? What is the DV?
2. What design type is this?
3. Write an experimental hypothesis (one-tailed).
4. Write a null hypothesis.
5. Why might concrete words be easier to remember?

---

## Debrief: Memory Design

**IV**: Word type (abstract vs concrete)

**DV**: Number of words recalled (0-10)

**Design**: Between-subjects (different people in each lab)

**Hypotheses**:

- **H1 (one-tailed)**: Concrete words will be recalled better than abstract words.
- **H0**: There will be no difference in recall between word types.

---

## Why Concrete Words?

**Dual Coding Theory** (Paivio, 1971)

Concrete words can be stored TWO ways:

1. **Verbal code** (the word itself)
2. **Visual code** (a mental image)

Abstract words only have the verbal code.

**Two codes > One code = Better memory**

---

# Phase 4: Frontal Lob {background-color="#EA8439"}

## Does handedness matter?

---

## What You Did

You threw balls at a bucket:

- **10 throws** with dominant hand
- **5 throws** with non-dominant hand

---

## Your Throwing Results

```{r}
#| label: throwing
#| fig-height: 5
#| fig-align: center

dom_col <- names(frontallob)[grep("dom.*success|dom.*hit", names(frontallob), ignore.case = TRUE)]
nondom_col <- names(frontallob)[grep("non.*success|non.*hit", names(frontallob), ignore.case = TRUE)]

if(length(dom_col) > 0 && length(nondom_col) > 0) {
  throw_data <- frontallob |>
    select(participant_id, dom = !!sym(dom_col[1]), nondom = !!sym(nondom_col[1])) |>
    filter(!is.na(dom) & !is.na(nondom)) |>
    mutate(
      dom_pct = dom / 10 * 100,
      nondom_pct = nondom / 5 * 100
    )
  
  throw_long <- throw_data |>
    pivot_longer(cols = c(dom_pct, nondom_pct),
                 names_to = "hand", values_to = "accuracy") |>
    mutate(hand = factor(hand, levels = c("dom_pct", "nondom_pct"),
                         labels = c("Dominant", "Non-Dominant")))
  
  ggplot(throw_long, aes(x = hand, y = accuracy, fill = hand)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    geom_line(data = throw_long |> pivot_wider(names_from = hand, values_from = accuracy) |>
                pivot_longer(cols = c(Dominant, `Non-Dominant`), names_to = "hand", values_to = "accuracy"),
              aes(group = participant_id), alpha = 0.2, colour = "gray40") +
    scale_fill_manual(values = c(navy, orange)) +
    labs(x = NULL, y = "Accuracy (%)",
         title = "Throwing Accuracy by Hand",
         subtitle = "Lines connect the same person") +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim_throw <- tibble(
    hand = rep(c("Dominant", "Non-Dominant"), each = 25),
    accuracy = c(rnorm(25, 78, 12), rnorm(25, 60, 15))
  )
  ggplot(sim_throw, aes(x = hand, y = accuracy, fill = hand)) +
    geom_boxplot(alpha = 0.6) +
    scale_fill_manual(values = c(navy, orange)) +
    labs(y = "Accuracy (%)", title = "Throwing Accuracy by Hand") +
    theme(legend.position = "none")
}
```

---

## Group Discussion

{{< countdown "5:00" >}}

**In your groups, answer Questions 4.1-4.6:**

1. Same people or different people in each condition?
2. What design type is this?
3. Why is within-subjects "more powerful"?
4. What is the IV? What is the DV?

---

## Debrief: Within-Subjects Power

**Design**: Within-subjects (same person, both hands)

**IV**: Hand used (dominant vs non-dominant)

**DV**: Throwing accuracy (%)

**Why more powerful?**

Each person is their own control. Individual differences (height, coordination, practice) are controlled because we compare YOU to YOU.

---

## The Statistical Test

For within-subjects designs with two conditions, we use a **paired t-test**.

It asks: Is the difference between conditions larger than we'd expect by chance?

```{r}
#| label: throw-stats
if(exists("throw_data") && nrow(throw_data) > 0) {
  mean_diff <- mean(throw_data$dom_pct - throw_data$nondom_pct, na.rm = TRUE)
  cat(paste0("Mean difference: ", round(mean_diff, 1), " percentage points"))
}
```

---

# Phase 5: Sleep & Memory {background-color="#1abc9c"}

## Is there a relationship?

---

## The Question

You told us how much sleep you got last night.

We measured how many words you recalled.

**Is there a relationship between sleep and memory?**

---

## Your Results

```{r}
#| label: sleep-memory
#| fig-height: 5
#| fig-align: center

# Get sleep and memory data from merged
sleep_mem <- merged |>
  select(participant_id, sleep_hours, memory_mem_total_1) |>
  filter(!is.na(sleep_hours) & !is.na(memory_mem_total_1) & 
         sleep_hours > 0 & sleep_hours < 15)

if(nrow(sleep_mem) > 5) {
  cor_val <- cor(sleep_mem$sleep_hours, sleep_mem$memory_mem_total_1, use = "complete.obs")
  
  ggplot(sleep_mem, aes(x = sleep_hours, y = memory_mem_total_1)) +
    geom_point(size = 4, colour = navy, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, colour = teal, fill = teal, alpha = 0.2) +
    labs(x = "Hours of Sleep Last Night", 
         y = "Words Recalled (out of 10)",
         title = "Sleep and Memory Performance",
         subtitle = paste0("r = ", round(cor_val, 2))) +
    scale_x_continuous(breaks = seq(4, 12, 2)) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim_sleep <- tibble(
    sleep = runif(28, 5, 9),
    memory = 4 + 0.4 * sleep + rnorm(28, 0, 1.2)
  )
  ggplot(sim_sleep, aes(x = sleep, y = memory)) +
    geom_point(size = 4, colour = navy, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, colour = teal, fill = teal, alpha = 0.2) +
    labs(x = "Hours of Sleep", y = "Words Recalled", 
         title = "Sleep and Memory Performance")
}
```

---

## Group Discussion

{{< countdown "5:00" >}}

**In your groups, answer Questions 5.1-5.5:**

1. Is there an IV in this study?
2. What design type is this?
3. What does the correlation tell us?
4. Can we say sleep **causes** better memory?
5. What's the key limitation of correlational designs?

---

## Debrief: Correlational Design

**Design**: Correlational (no manipulation)

**Variables**: Sleep hours AND memory score (neither is the IV)

**Key insight**: We measured BOTH variables — we didn't manipulate either.

::: {.callout-warning}
## The Critical Point
**Correlation ≠ Causation**

Maybe sleep helps memory... OR maybe third variables (e.g., health, motivation) affect both!
:::

---

# Phase 6: Sixth Sense {background-color="#9b59b6"}

## Can you feel being stared at?

---

## What You Did

6 trials: Someone either stared at you or didn't.

You guessed which it was.

**If just guessing**: You'd expect 50% correct (3 out of 6).

---

## Your Detection Accuracy

```{r}
#| label: sixth-sense
#| fig-height: 5
#| fig-align: center

# Calculate accuracy from actual vs guess columns
# Data has t1_actual, t1_guess, t2_actual, t2_guess, etc.
acc_data <- sixthsense |>
  mutate(
    correct_1 = as.numeric(t1_actual == t1_guess),
    correct_2 = as.numeric(t2_actual == t2_guess),
    correct_3 = as.numeric(t3_actual == t3_guess),
    correct_4 = as.numeric(t4_actual == t4_guess),
    correct_5 = as.numeric(t5_actual == t5_guess),
    correct_6 = as.numeric(t6_actual == t6_guess),
    total_correct = correct_1 + correct_2 + correct_3 + correct_4 + correct_5 + correct_6,
    accuracy = total_correct / 6
  ) |>
  select(participant_id, accuracy) |>
  filter(!is.na(accuracy))

if(nrow(acc_data) > 0) {
  overall_acc <- mean(acc_data$accuracy, na.rm = TRUE)
  
  ggplot(acc_data, aes(x = accuracy)) +
    geom_histogram(binwidth = 1/6, fill = "#9b59b6", colour = "white", alpha = 0.8) +
    geom_vline(xintercept = 0.5, linetype = "dashed", colour = coral, linewidth = 1.5) +
    annotate("text", x = 0.52, y = Inf, vjust = 2, label = "Chance (50%)",
             colour = coral, fontface = "bold", hjust = 0) +
    scale_x_continuous(labels = percent, limits = c(0, 1)) +
    labs(x = "Accuracy", y = "Count",
         title = "Sixth Sense Detection",
         subtitle = paste0("Overall accuracy: ", round(overall_acc * 100, 0), "%")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", colour = navy))
} else {
  set.seed(42)
  sim_sixth <- tibble(accuracy = rbinom(28, 6, 0.52) / 6)
  ggplot(sim_sixth, aes(x = accuracy)) +
    geom_histogram(binwidth = 1/6, fill = "#9b59b6", colour = "white") +
    geom_vline(xintercept = 0.5, linetype = "dashed", colour = coral, linewidth = 1.5) +
    scale_x_continuous(labels = percent) +
    labs(x = "Accuracy", y = "Count", title = "Sixth Sense Detection")
}
```

---

## Group Discussion

{{< countdown "4:00" >}}

**In your groups, answer Questions 6.1-6.4:**

1. If someone got 5/6 correct, does that prove they have a sixth sense?
2. What is the null hypothesis here?
3. What would 50% accuracy suggest?

---

## Debrief: Chance and the Null Hypothesis

**H0**: Performance equals chance (50% - just guessing)

**H1**: Performance differs from chance

**Key insight**: Some people will get 5/6 or 6/6 just by luck, even with no real ability.

With 28 people flipping coins, SOMEONE will get 5+ heads. That doesn't mean the coin is biased.

**This is why we need statistics** - to separate signal from noise.

---

# Wrap-Up {background-color="#275882"}

## What have we learned?

---

## Design Types Summary

| Design | Same people? | Example from today |
|--------|--------------|-------------------|
| Between-subjects | No | Memory (Lab A vs Lab B) |
| Within-subjects | Yes | Throwing (Dom vs Non-dom) |
| Correlational | N/A | Sleep & Memory |

---

## Key Vocabulary

:::: {.columns}
::: {.column width="50%"}
**Variables**

- **IV**: What's manipulated
- **DV**: What's measured

**Hypotheses**

- **H1**: Predicts an effect
- **H0**: Predicts no effect
:::

::: {.column width="50%"}
**Direction**

- **One-tailed**: Predicts direction
- **Two-tailed**: Predicts difference only

**Key Concept**

- **Chance**: Random variation
- **Statistics**: Separating signal from noise
:::
::::

---

## Your Assessments

Everything today connects to your assessments:

| Assessment | What you practiced today |
|------------|-------------------------|
| **Research Plan** | IV, DV, design types, hypotheses |
| **MCQ Exam** | All terminology and concepts |
| **EDS Assignment** | Means, SDs, interpreting results |

---

## Next Steps

1. Complete the **Week 12 Self-Assessment Quiz**

2. Review today's activity page

3. Start thinking about your **Research Plan** (due Week 16)

---

# Questions? {background-color="#275882"}

---

# Well done! {background-color="#275882"}

::: {.notes}
Total time: approximately 80 minutes

The slides provide structure but the real learning happens in the small group discussions and whole-class debriefs. Adjust timing as needed based on how discussions develop.
:::
