---
title: "Foundations Datalab"
subtitle: "Your Data, Your Results"
author: "PS50008A Research Methods"
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    transition: slide
    code-fold: true
    code-tools: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(knitr)

# Load all datasets (already cleaned by Python pipeline)
demog <- read_csv("../data/processed/demog_clean.csv", show_col_types = FALSE)
memory <- read_csv("../data/processed/memory_clean.csv", show_col_types = FALSE)
reflex <- read_csv("../data/processed/task1_reflex_clean.csv", show_col_types = FALSE)
timewarp <- read_csv("../data/processed/task2_timewarp_clean.csv", show_col_types = FALSE)
sixthsense <- read_csv("../data/processed/task3_sixthsense_clean.csv", show_col_types = FALSE)
ladyluck <- read_csv("../data/processed/task4_ladyluck_clean.csv", show_col_types = FALSE)
frontallob <- read_csv("../data/processed/task5_frontallob_clean.csv", show_col_types = FALSE)
post <- read_csv("../data/processed/post_clean.csv", show_col_types = FALSE)
```

# Welcome! {background-color="#2c3e50"}

## What We Did

You participated in **6 classic psychology tasks**:

1. **Memory Test** - Word recall and serial position
2. **Reflex Test** - Ruler catch reaction time
3. **Time Warp** - Time and length estimation
4. **Sixth Sense** - Can you detect being stared at?
5. **Lady Luck** - Random number generation
6. **Frontal Lob** - Motor control and handedness

## Our Sample

```{r}
#| label: sample-overview
#| fig-height: 4
demog |>
  ggplot(aes(x = factor(lab_room, labels = c("Lab A", "Lab B")))) +
  geom_bar(fill = "#3498db", width = 0.6) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 6) +
  labs(x = NULL, y = "Count", title = "Participants by Lab") +
  theme_minimal(base_size = 18) +
  ylim(0, 15)
```

**N = `r nrow(demog)`** participants completed demographics

## Who Are You?

:::: {.columns}
::: {.column width="50%"}
```{r}
#| label: age-dist
#| fig-height: 5
demog |>
  filter(!is.na(age)) |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 2, fill = "#9b59b6", color = "white") +
  labs(x = "Age", y = "Count", title = "Age Distribution") +
  theme_minimal(base_size = 16)
```
:::
::: {.column width="50%"}
```{r}
#| label: gender-dist
#| fig-height: 5
demog |>
  filter(!is.na(gender)) |>
  mutate(gender = factor(gender, labels = c("Male", "Female"))) |>
  ggplot(aes(x = gender)) +
  geom_bar(fill = "#e74c3c", width = 0.6) +
  labs(x = NULL, y = "Count", title = "Gender") +
  theme_minimal(base_size = 16)
```
:::
::::

# Memory Task {background-color="#27ae60"}

## The Serial Position Effect

You saw **10 words** and tried to recall them in order.

**Classic finding**: We remember the first words (primacy) and last words (recency) better than the middle.

## Your Results: Words Recalled

```{r}
#| label: memory-total
#| fig-height: 5
memory |>
  ggplot(aes(x = mem_total)) +
  geom_histogram(binwidth = 1, fill = "#27ae60", color = "white") +
  geom_vline(xintercept = mean(memory$mem_total, na.rm = TRUE),
             linetype = "dashed", color = "red", linewidth = 1.5) +
  annotate("text", x = mean(memory$mem_total, na.rm = TRUE) + 0.5, y = 4,
           label = paste("Mean =", round(mean(memory$mem_total, na.rm = TRUE), 1)),
           hjust = 0, size = 5, color = "red") +
  labs(x = "Words Correctly Recalled (out of 10)", y = "Count",
       title = "Memory Performance") +
  scale_x_continuous(breaks = 0:10) +
  theme_minimal(base_size = 18)
```

## Serial Position Curve

```{r}
#| label: serial-position
#| fig-height: 5
# Calculate accuracy at each position (1 = correct, 0 = incorrect)
position_data <- memory |>
  select(participant_id, pos1:pos10) |>
  pivot_longer(cols = pos1:pos10,
               names_to = "position",
               values_to = "correct") |>
  mutate(position = as.numeric(str_extract(position, "\\d+"))) |>
  group_by(position) |>
  summarise(prop_correct = mean(correct, na.rm = TRUE),
            se = sd(correct, na.rm = TRUE) / sqrt(n()))

position_data |>
  ggplot(aes(x = position, y = prop_correct)) +
  geom_line(color = "#27ae60", linewidth = 1.5) +
  geom_point(color = "#27ae60", size = 4) +
  geom_errorbar(aes(ymin = prop_correct - se, ymax = prop_correct + se),
                width = 0.2, color = "#27ae60") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "Word Position", y = "Proportion Correct Position",
       title = "Serial Position Effect",
       subtitle = "Do you see primacy and recency?") +
  theme_minimal(base_size = 18)
```

# Reaction Time {background-color="#e74c3c"}

## The Ruler Catch Test

You caught a falling ruler **5 times**.

The distance caught (in cm) reflects your **reaction time** - shorter distance = faster reactions!

## Your Reaction Times

```{r}
#| label: reflex-trials
#| fig-height: 5
reflex_long <- reflex |>
  select(participant_id, trial1_cm:trial5_cm) |>
  pivot_longer(cols = trial1_cm:trial5_cm,
               names_to = "trial",
               values_to = "distance_cm") |>
  filter(!is.na(distance_cm) & distance_cm > 0) |>
  mutate(trial = as.numeric(str_extract(trial, "\\d+")))

reflex_long |>
  ggplot(aes(x = factor(trial), y = distance_cm)) +
  geom_boxplot(fill = "#e74c3c", alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  labs(x = "Trial", y = "Distance Caught (cm)",
       title = "Reaction Time Across Trials",
       subtitle = "Lower = Faster reactions") +
  theme_minimal(base_size = 18)
```

## Do You Get Better With Practice?

```{r}
#| label: reflex-learning
#| fig-height: 5
reflex_means <- reflex_long |>
  group_by(trial) |>
  summarise(mean_dist = mean(distance_cm, na.rm = TRUE),
            se = sd(distance_cm, na.rm = TRUE) / sqrt(n()))

reflex_means |>
  ggplot(aes(x = trial, y = mean_dist)) +
  geom_line(color = "#e74c3c", linewidth = 1.5) +
  geom_point(color = "#e74c3c", size = 4) +
  geom_errorbar(aes(ymin = mean_dist - se, ymax = mean_dist + se),
                width = 0.2, color = "#e74c3c") +
  labs(x = "Trial", y = "Mean Distance (cm)",
       title = "Learning Effect?",
       subtitle = "Does reaction time improve with practice?") +
  theme_minimal(base_size = 18)
```

# Time Estimation {background-color="#f39c12"}

## Can You Estimate 60 Seconds?

Without counting, you estimated when **60 seconds** had passed.

```{r}
#| label: time-estimate
#| fig-height: 5
timewarp |>
  filter(!is.na(time_estimate_sec)) |>
  mutate(time_estimate_sec = as.numeric(time_estimate_sec)) |>
  filter(!is.na(time_estimate_sec) & time_estimate_sec < 120) |>
  ggplot(aes(x = time_estimate_sec)) +
  geom_histogram(binwidth = 5, fill = "#f39c12", color = "white") +
  geom_vline(xintercept = 60, linetype = "dashed", color = "blue", linewidth = 1.5) +
  annotate("text", x = 62, y = 3, label = "Target: 60s", hjust = 0,
           color = "blue", size = 5) +
  labs(x = "Your Estimate (seconds)", y = "Count",
       title = "Time Estimation",
       subtitle = "How close were you to 60 seconds?") +
  theme_minimal(base_size = 18)
```

# Sixth Sense {background-color="#9b59b6"}

## Can You Feel Being Stared At?

You completed **6 trials** where someone either stared at you or looked away.

You guessed whether you were being stared at and rated your confidence.

## Detection Accuracy

```{r}
#| label: stare-accuracy
#| fig-height: 5
# Calculate accuracy for each participant
stare_data <- sixthsense |>
  select(participant_id, contains("actual"), contains("guess")) |>
  pivot_longer(cols = -participant_id,
               names_to = c("trial", ".value"),
               names_pattern = "t(\\d+)_(.+)") |>
  mutate(correct = actual == guess)

accuracy_summary <- stare_data |>
  group_by(participant_id) |>
  summarise(accuracy = mean(correct, na.rm = TRUE))

accuracy_summary |>
  ggplot(aes(x = accuracy)) +
  geom_histogram(binwidth = 1/6, fill = "#9b59b6", color = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "red", linewidth = 1.5) +
  annotate("text", x = 0.52, y = 3, label = "Chance (50%)", hjust = 0,
           color = "red", size = 5) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "Accuracy", y = "Count",
       title = "Sixth Sense: Detection Accuracy",
       subtitle = "Can you do better than chance?") +
  theme_minimal(base_size = 18)
```

## Group Performance

```{r}
#| label: stare-group
overall_acc <- mean(stare_data$correct, na.rm = TRUE)
```

**Overall accuracy: `r scales::percent(overall_acc, accuracy = 0.1)`**

```{r}
#| label: stare-test
# One-sample t-test against chance
t_result <- t.test(accuracy_summary$accuracy, mu = 0.5)
```

Statistical test against chance (50%):

- **t(`r t_result$parameter`)** = `r round(t_result$statistic, 2)`
- **p** = `r round(t_result$p.value, 3)`

::: {.fragment}
**Conclusion**: `r if(t_result$p.value < 0.05) "We performed significantly different from chance!" else "We performed at chance level - no sixth sense detected!"`
:::

# Lady Luck {background-color="#1abc9c"}

## Random Number Generation

You generated **20 random numbers** between 1-10.

Humans are notoriously **bad** at being random - we avoid repetitions and prefer certain numbers.

## Number Frequency

```{r}
#| label: random-numbers
#| fig-height: 5
rand_data <- ladyluck |>
  select(participant_id, rand1:rand10) |>
  pivot_longer(cols = -participant_id,
               names_to = "position",
               values_to = "number") |>
  filter(!is.na(number) & number >= 1 & number <= 10)

rand_data |>
  ggplot(aes(x = factor(number))) +
  geom_bar(fill = "#1abc9c") +
  geom_hline(yintercept = nrow(rand_data) / 10, linetype = "dashed",
             color = "red", linewidth = 1) +
  annotate("text", x = 10, y = nrow(rand_data) / 10 + 2,
           label = "Expected if truly random", hjust = 1, color = "red", size = 4) +
  labs(x = "Number", y = "Frequency",
       title = "Which Numbers Did You Generate?",
       subtitle = "If truly random, each number should appear equally often") +
  theme_minimal(base_size = 18)
```

## Avoiding Repetitions?

```{r}
#| label: repetitions
#| fig-height: 4
# Check for consecutive repetitions
check_reps <- ladyluck |>
  select(participant_id, rand1:rand10) |>
  pivot_longer(cols = -participant_id, names_to = "pos", values_to = "num") |>
  group_by(participant_id) |>
  arrange(participant_id, pos) |>
  mutate(repetition = num == lag(num)) |>
  summarise(n_reps = sum(repetition, na.rm = TRUE))

mean_reps <- mean(check_reps$n_reps)
expected_reps <- 9 * 0.1  # 9 opportunities, 10% chance each

tibble(
  Type = c("Your Data", "Expected (Random)"),
  Repetitions = c(mean_reps, expected_reps)
) |>
  ggplot(aes(x = Type, y = Repetitions, fill = Type)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c("#e74c3c", "#1abc9c")) +
  labs(y = "Mean Consecutive Repetitions", x = NULL,
       title = "Do Humans Avoid Repetitions?") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
```

# Motor Control {background-color="#3498db"}

## Dominant vs Non-Dominant Hand

You threw balls at a target with **both hands**.

How much does handedness matter?

## Throwing Accuracy

```{r}
#| label: throwing
#| fig-height: 5
throw_data <- frontallob |>
  select(participant_id, dom_success, nondom_success) |>
  mutate(dom_pct = dom_success / 10 * 100,
         nondom_pct = nondom_success / 5 * 100) |>
  pivot_longer(cols = c(dom_pct, nondom_pct),
               names_to = "hand",
               values_to = "accuracy") |>
  mutate(hand = factor(hand, levels = c("dom_pct", "nondom_pct"),
                       labels = c("Dominant", "Non-Dominant")))

throw_data |>
  ggplot(aes(x = hand, y = accuracy, fill = hand)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("#3498db", "#e67e22")) +
  labs(x = NULL, y = "Accuracy (%)",
       title = "Throwing Accuracy by Hand",
       subtitle = "Dominant hand advantage?") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
```

# Wrap Up {background-color="#2c3e50"}

## What Did You Enjoy?

```{r}
#| label: enjoyment
#| fig-height: 5
post |>
  filter(!is.na(fav_task)) |>
  count(fav_task) |>
  ggplot(aes(x = reorder(fav_task, n), y = n)) +
  geom_col(fill = "#27ae60") +
  coord_flip() +
  labs(x = NULL, y = "Votes",
       title = "Favourite Task") +
  theme_minimal(base_size = 18)
```

## Pre vs Post Mood

```{r}
#| label: mood-change
#| fig-height: 4
# Merge pre and post mood
mood_data <- demog |>
  select(participant_id, mood_pre, energy_pre, stress_pre) |>
  inner_join(
    post |> select(participant_id, mood_post, energy_post, stress_post),
    by = "participant_id"
  ) |>
  pivot_longer(cols = -participant_id,
               names_to = c("measure", "time"),
               names_pattern = "(mood|energy|stress)_(pre|post)") |>
  mutate(time = factor(time, levels = c("pre", "post")))

mood_data |>
  group_by(measure, time) |>
  summarise(mean_val = mean(value, na.rm = TRUE),
            se = sd(value, na.rm = TRUE) / sqrt(n())) |>
  ggplot(aes(x = time, y = mean_val, fill = time)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_val - se, ymax = mean_val + se), width = 0.2) +
  facet_wrap(~measure) +
  scale_fill_manual(values = c("#95a5a6", "#3498db")) +
  labs(x = NULL, y = "Rating (1-10)", title = "Before vs After the Session") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none")
```

## Now It's Your Turn!

Explore the data yourself using the **interactive WebR pages**:

- Run the code
- Modify analyses
- Ask new questions

::: {.callout-tip}
## Key Skills
- Data wrangling with `dplyr`
- Visualization with `ggplot2`
- Statistical testing with t-tests
:::

# Questions? {background-color="#2c3e50"}

## Thank You!

Data and code available in the course materials.

**Happy Analysing!**
