---
title: "Your Data, Your Discoveries"
subtitle: "Foundations Datalab Results"
author: "PS50008A Research Methods"
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    transition: slide
    code-fold: true
    code-tools: true
    fig-align: center
    footer: "Foundations Datalab | Wednesday Lecture"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(knitr)
library(scales)

# Load all datasets
demog <- read_csv("../data/processed/demog_clean.csv", show_col_types = FALSE)
memory <- read_csv("../data/processed/memory_clean.csv", show_col_types = FALSE)
reflex <- read_csv("../data/processed/task1_reflex_clean.csv", show_col_types = FALSE)
timewarp <- read_csv("../data/processed/task2_timewarp_clean.csv", show_col_types = FALSE)
sixthsense <- read_csv("../data/processed/task3_sixthsense_clean.csv", show_col_types = FALSE)
ladyluck <- read_csv("../data/processed/task4_ladyluck_clean.csv", show_col_types = FALSE)
frontallob <- read_csv("../data/processed/task5_frontallob_clean.csv", show_col_types = FALSE)
post <- read_csv("../data/processed/post_clean.csv", show_col_types = FALSE)

theme_set(theme_minimal(base_size = 18))
```

# What Did You Actually Do? {background-color="#2c3e50"}

## DataLab: The Tasks

You completed **6 tasks** plus pre/post surveys:

1. **Memory Test** (Mind Palace)
2. **Reflex Test** (Ruler catch)
3. **Time Warp** (60 second estimation)
4. **Sixth Sense** (Staring detection)
5. **Lady Luck** (Randomness)
6. **Frontal Lob** (Throwing accuracy)

::: {.fragment}
Each task generates **data** - numbers we can analyse!
:::

## What Is Data?

**Data** = observations recorded as values

| What you did | What we recorded |
|--------------|------------------|
| Caught a ruler | Distance in cm |
| Estimated 60 seconds | Your estimate in seconds |
| Remembered words | Number correct |
| Threw balls at a bucket | How many went in |

::: {.fragment}
Every number tells part of a story.
:::

## You Are a Data Point

On every graph today, one point is **you**.

```{r}
#| label: you-are-here
#| fig-height: 4
mem_plot <- memory |>
  filter(!is.na(mem_total))

ggplot(mem_plot, aes(x = "", y = mem_total)) +
  geom_jitter(width = 0.2, size = 4, alpha = 0.6, color = "#3498db") +
  labs(x = NULL, y = "Words Recalled",
       title = "Memory Scores - Each Dot Is One Person") +
  theme(axis.text.x = element_blank())
```

::: {.fragment}
Can you spot yourself?
:::

# Looking at Your Data {background-color="#27ae60"}

## The Simplest Question: What's Typical?

When we have a bunch of numbers, we want to know: **what's normal?**

::: {.fragment}
**The Average (Mean)**: Add them up, divide by how many
:::

::: {.fragment}
```{r}
#| echo: true
# Memory scores
memory_scores <- c(5, 3, 6, 6, 6, 7, 7, 6, 6, 10, 8, 3, 3, 9, 8, 7, 2, 8, 8, 6)
mean(memory_scores)
```
:::

## But Averages Can Be Misleading...

Two groups with the **same average** can look very different!

```{r}
#| label: spread-demo
#| fig-height: 4
tibble(
  Group = rep(c("Group A", "Group B"), each = 10),
  Score = c(5, 5, 6, 6, 6, 6, 6, 7, 7, 7,   # Group A: clustered
            2, 3, 4, 5, 6, 7, 8, 9, 10, 7)   # Group B: spread out
) |>
  ggplot(aes(x = Group, y = Score, color = Group)) +
  geom_jitter(width = 0.1, size = 4) +
  geom_hline(yintercept = 6.1, linetype = "dashed") +
  scale_color_manual(values = c("#3498db", "#e74c3c")) +
  labs(y = "Score", title = "Same Average, Different Spread") +
  annotate("text", x = 2.5, y = 6.3, label = "Both groups: Mean ≈ 6") +
  theme(legend.position = "none")
```

::: {.fragment}
We also need to know about **spread** (variation)
:::

## Visualising Data: The Histogram

A histogram shows **how often** each value occurs.

```{r}
#| label: memory-hist
#| fig-height: 4.5
ggplot(memory, aes(x = mem_total)) +
  geom_histogram(binwidth = 1, fill = "#3498db", color = "white", alpha = 0.8) +
  labs(x = "Words Recalled (out of 10)", y = "Number of People",
       title = "How Many Words Did You Remember?") +
  scale_x_continuous(breaks = 0:10)
```

::: {.fragment}
Each bar = how many people got that score
:::

## Your Reaction Times

```{r}
#| label: rt-simple
#| fig-height: 4.5
reflex_long <- reflex |>
  pivot_longer(cols = starts_with("trial"),
               names_to = "trial", values_to = "distance") |>
  filter(!is.na(distance) & distance > 0 & distance < 50) |>
  mutate(trial_num = as.numeric(str_extract(trial, "\\d+")))

ggplot(reflex_long, aes(x = distance)) +
  geom_histogram(binwidth = 2, fill = "#27ae60", color = "white") +
  labs(x = "Distance Caught (cm)", y = "Count",
       title = "Reflex Test: How Far Did the Ruler Fall?")
```

::: {.fragment}
**Shorter distance = faster reaction**
:::

## Time Estimation

You tried to count **60 seconds** in your head.

```{r}
#| label: time-est
#| fig-height: 4.5
timewarp_clean <- timewarp |>
  filter(!is.na(time_estimate_sec) & time_estimate_sec > 0 & time_estimate_sec < 200)

ggplot(timewarp_clean, aes(x = time_estimate_sec)) +
  geom_histogram(binwidth = 5, fill = "#9b59b6", color = "white") +
  geom_vline(xintercept = 60, color = "red", linewidth = 2, linetype = "dashed") +
  labs(x = "Your Estimate (seconds)", y = "Count",
       title = "60 Second Estimation") +
  annotate("text", x = 62, y = 3, label = "Actual\n60 sec",
           color = "red", hjust = 0, size = 5)
```

::: {.fragment}
Were you an over-estimator or under-estimator?
:::

# The Memory Experiment {background-color="#f39c12"}

## What Was Different Between Labs?

This wasn't just about testing memory...

::: {.fragment}
**Lab A (with Gordon)**: Memorised **abstract** words

*things like: justice, freedom, wisdom...*
:::

::: {.fragment}
**Lab B (with Bence)**: Memorised **concrete** words

*things like: chair, river, mountain...*
:::

::: {.fragment}
**The Question**: Are concrete words easier to remember?
:::

## Why Would Word Type Matter?

::: {.fragment}
**Concrete words** = you can picture them

"Think of an **apple**" → You see an apple
:::

::: {.fragment}
**Abstract words** = harder to visualise

"Think of **justice**" → ???
:::

::: {.fragment}
The **dual coding hypothesis**: Pictures + words = better memory
:::

## Let's Look at the Data

```{r}
#| label: memory-by-lab
#| fig-height: 5
# lab_room indicates lab group (1 = A, 2 = B)
memory_by_lab <- memory |>
  filter(!is.na(lab_room) & !is.na(mem_total))

ggplot(memory_by_lab, aes(x = factor(lab_room, labels = c("Lab A\n(Abstract)", "Lab B\n(Concrete)")),
                          y = mem_total,
                          fill = factor(lab_room))) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.6) +
  scale_fill_manual(values = c("#3498db", "#e67e22")) +
  labs(x = NULL, y = "Words Recalled (out of 10)",
       title = "Memory by Lab Group") +
  theme(legend.position = "none")
```

## Comparing Groups

```{r}
#| label: group-means
lab_a <- memory_by_lab |> filter(lab_room == 1) |> pull(mem_total)
lab_b <- memory_by_lab |> filter(lab_room == 2) |> pull(mem_total)
```

| Lab | Word Type | Mean | How Many |
|-----|-----------|------|----------|
| Lab A | Abstract | `r round(mean(lab_a, na.rm=TRUE), 1)` words | `r length(lab_a)` people |
| Lab B | Concrete | `r round(mean(lab_b, na.rm=TRUE), 1)` words | `r length(lab_b)` people |

::: {.fragment}
**Difference**: `r round(mean(lab_b, na.rm=TRUE) - mean(lab_a, na.rm=TRUE), 1)` words

Is this difference **real** or just **luck**?
:::

# What Is "Statistically Significant"? {background-color="#9b59b6"}

## The Problem with Small Samples

Even if there's **no real difference**, random chance can create one.

::: {.fragment}
Imagine flipping a coin 10 times:

- Sometimes you get 7 heads, sometimes 3
- That doesn't mean the coin is biased!
:::

::: {.fragment}
Same with experiments: we need to ask **"could chance explain this?"**
:::

## The Sixth Sense Task

You tried to detect if someone was staring at you.

- 6 trials
- Guessed "staring" or "not staring"

::: {.fragment}
**If you're just guessing**: You'd expect to get **3 out of 6** right (50%)
:::

## Your Results

```{r}
#| label: sixth-results
#| fig-height: 4
stare_data <- sixthsense |>
  select(participant_id, starts_with("t")) |>
  pivot_longer(
    cols = starts_with("t"),
    names_to = c("trial", ".value"),
    names_pattern = "t(\\d+)_(.+)"
  ) |>
  mutate(correct = actual == guess)

accuracy <- stare_data |>
  group_by(participant_id) |>
  summarise(n_correct = sum(correct, na.rm = TRUE),
            accuracy = mean(correct, na.rm = TRUE))

ggplot(accuracy, aes(x = factor(n_correct))) +
  geom_bar(fill = "#9b59b6", alpha = 0.8) +
  geom_vline(xintercept = 3.5, linetype = "dashed", color = "red", linewidth = 1.5) +
  labs(x = "How Many Correct (out of 6)", y = "Number of People",
       title = "Sixth Sense Accuracy") +
  annotate("text", x = 4.5, y = 4, label = "3 = chance", color = "red", size = 5)
```

## Some People Got 5 or 6 Right!

Does that mean they have a sixth sense?

::: {.fragment}
**Not necessarily!**

If you flip a coin 6 times, sometimes you'll get 5 or 6 heads **just by luck**.
:::

::: {.fragment}
Statistics helps us figure out: **how surprising is this result?**
:::

## The Key Question

When we see a difference or pattern:

1. Is it a **real effect**?
2. Or could **random chance** produce this?

::: {.fragment}
This is what **statistical testing** helps us determine.

Tomorrow in the lab, you'll learn how to answer this question!
:::

# What Makes You... You? {background-color="#3498db"}

## Self-Ratings vs Reality

Before the throwing task, you rated how good you thought you'd be.

**Question**: Do people know themselves?

## Your Self-Ratings

```{r}
#| label: self-rating
#| fig-height: 5
throw_data <- frontallob |>
  mutate(dom_accuracy = dom_success / 10 * 100) |>
  filter(!is.na(self_rating) & !is.na(dom_accuracy))

ggplot(throw_data, aes(x = self_rating, y = dom_accuracy)) +
  geom_point(size = 5, color = "#3498db", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "#e74c3c", linewidth = 2) +
  labs(x = "How Good You Thought You'd Be (1-10)",
       y = "Actual Accuracy (%)",
       title = "Self-Rating vs Actual Performance") +
  scale_x_continuous(breaks = 1:10)
```

::: {.fragment}
Each dot is one person. The line shows the **relationship**.
:::

## What the Pattern Tells Us

If people know themselves well:

- High self-raters → High performance
- Low self-raters → Low performance
- Points cluster around the line

::: {.fragment}
If people **don't** know themselves:

- Self-ratings don't predict performance
- Points scattered everywhere
:::

# Your "Random" Numbers {background-color="#e74c3c"}

## Can Humans Be Random?

You tried to generate 20 random numbers between 1 and 10.

If truly random: each number should appear **equally often**.

## What You Actually Did

```{r}
#| label: random-numbers
#| fig-height: 4.5
rand_data <- ladyluck |>
  select(rand1:rand10) |>
  pivot_longer(everything(), values_to = "number") |>
  filter(!is.na(number) & number >= 1 & number <= 10)

expected <- nrow(rand_data) / 10

ggplot(rand_data, aes(x = factor(number))) +
  geom_bar(fill = "#e74c3c", alpha = 0.8) +
  geom_hline(yintercept = expected, linetype = "dashed",
             color = "blue", linewidth = 1.5) +
  labs(x = "Number", y = "How Often Chosen",
       title = "Your 'Random' Numbers") +
  annotate("text", x = 9.5, y = expected + 2,
           label = "Expected if random", color = "blue", hjust = 1)
```

::: {.fragment}
**Notice anything?** Humans avoid 1 and 10, favour middle numbers!
:::

## We're Bad at Random

This is a well-known finding in psychology:

- We avoid **extremes** (1 and 10 feel "not random")
- We avoid **patterns** (not 1,2,3,4...)
- We avoid **repeats** (same number twice feels wrong)

::: {.fragment}
Real randomness doesn't care about any of that!
:::

# Key Takeaways {background-color="#2c3e50"}

## What You've Learned Today

1. **Data = recorded observations** that we can analyse

2. **Averages** tell us what's typical, but **spread** matters too

3. **Histograms** show us the shape of our data

4. **Group differences** might be real or might be chance

5. **Statistical tests** help us tell the difference

## Tomorrow's Lab

You'll get hands-on with this data:

- **Visualise** your own results
- **Calculate** averages and compare groups
- **Test** whether differences are significant

::: {.fragment}
Bring your laptop and find **your data point** in every analysis!
:::

## Questions?

::: {.callout-tip}
## Remember
- You generated this data
- Every number came from something you did
- Tomorrow you'll analyse it yourself
:::

# See You in the Lab! {background-color="#2c3e50"}
