---
title: "Correlation in R"
subtitle: "PS50008A Week 15 Lab Activity"
author: "Gordon Wright"
date: "2026-02-11"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(webexercises)
library(tidyverse)
library(knitr)

# Load DataLab data
memory <- read_csv("memory_clean.csv", show_col_types = FALSE)
merged <- read_csv("foundations_merged.csv", show_col_types = FALSE)

navy <- "#275882"
orange <- "#EA8439"
theme_set(theme_minimal(base_size = 14))
```

## Overview

::: {.callout-important}
## Focus for Today
This lab explores **correlation** - measuring the relationship between two continuous variables:

- Creating effective scatterplots
- Calculating Pearson's r in R
- Interpreting correlation coefficients
- Applying correlation to your own data
:::

---

## Activity 1: Creating Scatterplots (15 min)

### Visualising Relationships

Before calculating correlations, we should always visualise the relationship with a scatterplot.

```{r}
#| label: frontallob-data
frontallob <- read_csv("task5_frontallob_clean.csv", show_col_types = FALSE)

frontallob_clean <- frontallob |>
  filter(!is.na(self_rating) & !is.na(dom_success))
```

```{r}
#| label: scatterplot-example
#| fig-height: 5
#| fig-align: center
ggplot(frontallob_clean, aes(x = self_rating, y = dom_success)) +
  geom_point(size = 3, colour = navy, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = orange, fill = orange, alpha = 0.2) +
  labs(x = "Self-Rating (1-10)",
       y = "Successful Throws (out of 10)",
       title = "Self-Confidence vs Throwing Accuracy",
       subtitle = "Each point = one participant") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

### Understanding Scatterplots

**1. In a scatterplot, each point represents:**

`r mcq(c("A variable", "A group", answer = "One participant", "An average"))`

**2. The x-axis typically shows:**

`r mcq(c(answer = "The predictor or first variable", "The outcome variable", "The sample size", "The correlation coefficient"))`

**3. What does the line of best fit show?**

`r mcq(c("The average of all points", answer = "The predicted relationship between the variables", "Where most points should fall", "The maximum possible score"))`

### Reading Scatterplot Patterns

**4. If points cluster tightly around an upward-sloping line, this suggests:**

`r mcq(c(answer = "A strong positive correlation", "A strong negative correlation", "No correlation", "A weak relationship"))`

**5. If points are scattered randomly with no clear pattern, this suggests:**

`r mcq(c("A strong positive correlation", "A strong negative correlation", answer = "No correlation (r close to 0)", "A curvilinear relationship"))`

**6. If points cluster around a downward-sloping line, this suggests:**

`r mcq(c("A strong positive correlation", answer = "A strong negative correlation", "No correlation", "The data is wrong"))`

True or False: We should always create a scatterplot before calculating a correlation. `r torf(TRUE)`

True or False: Scatterplots can reveal non-linear relationships that correlation coefficients might miss. `r torf(TRUE)`

---

## Activity 2: Calculating Correlations (15 min)

### Pearson's r

The Pearson correlation coefficient (r) measures the strength and direction of a **linear** relationship between two continuous variables.

### Properties of r

**1. Pearson's r ranges from:**

`r mcq(c("0 to 1", "-1 to 0", answer = "-1 to +1", "0 to 100"))`

**2. An r value of 0 means:**

`r mcq(c("A perfect positive relationship", "A perfect negative relationship", answer = "No linear relationship", "The calculation failed"))`

**3. An r value of -0.8 indicates:**

`r mcq(c("A weak positive relationship", answer = "A strong negative relationship", "A weak negative relationship", "No relationship"))`

### R Syntax

In R, we calculate correlation using `cor()` or `cor.test()`:

```r
# Simple correlation
cor(variable1, variable2)

# Correlation with p-value and CI
cor.test(variable1, variable2)
```

**4. What is the key difference between `cor()` and `cor.test()`?**

`r mcq(c("cor() is faster", answer = "cor.test() provides a p-value and confidence interval", "cor() gives more decimal places", "They calculate different types of correlation"))`

### Reading R Output

Here's example output from `cor.test()`:

```
	Pearson's product-moment correlation

data:  self_rating and accuracy
t = 3.45, df = 28, p-value = 0.002
alternative hypothesis: true correlation is not equal to 0
95 percent confidence interval:
 0.24  0.75
sample estimates:
      cor
0.55
```

**5. What is the correlation coefficient (r)?**

`r mcq(c("3.45", "0.002", "28", answer = "0.55"))`

**6. Is this correlation statistically significant at alpha = 0.05?**

`r mcq(c(answer = "Yes, because p = 0.002 < 0.05", "No, because the correlation is positive", "Can't tell", "No, because df = 28"))`

**7. The 95% CI is [0.24, 0.75]. What does this tell us?**

`r mcq(c("The correlation is exactly 0.55", answer = "We're 95% confident the true r is between 0.24 and 0.75", "The p-value is between 0.24 and 0.75", "The sample size should be 24 to 75"))`

---

## Activity 2b: Interpreting Scatterplots (10 min)

Look at each scatterplot below and estimate the correlation.

```{r}
#| label: correlation-examples
#| fig-height: 8
#| fig-width: 10
#| fig-align: center

set.seed(42)
n <- 50

study_hours <- runif(n, 1, 10)
exam_scores <- 40 + 5 * study_hours + rnorm(n, 0, 5)

screen_time <- runif(n, 1, 8)
sleep_hours <- 9 - 0.5 * screen_time + rnorm(n, 0, 1.2)

shoe_size <- rnorm(n, 9, 1.5)
iq_score <- rnorm(n, 100, 15)

stress <- runif(n, 1, 10)
wellbeing <- 95 - 7 * stress + rnorm(n, 0, 8)

coffee <- rpois(n, 2.5)
alertness <- 50 + 3 * coffee + rnorm(n, 0, 15)

exercise <- runif(n, 0, 7)
mood <- 40 + 5 * exercise + rnorm(n, 0, 12)

examples <- tibble(
  panel = rep(LETTERS[1:6], each = n),
  x = c(study_hours, screen_time, shoe_size, stress, coffee, exercise),
  y = c(exam_scores, sleep_hours, iq_score, wellbeing, alertness, mood),
  scenario = rep(c("Study Hours vs Exam Score",
                   "Screen Time vs Sleep Hours",
                   "Shoe Size vs IQ Score",
                   "Stress Level vs Wellbeing",
                   "Coffees per Day vs Alertness",
                   "Exercise (hrs/wk) vs Mood"), each = n)
)

ggplot(examples, aes(x = x, y = y)) +
  geom_point(colour = navy, alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = orange, linewidth = 1) +
  facet_wrap(~scenario, scales = "free", ncol = 3) +
  labs(x = NULL, y = NULL) +
  theme(strip.text = element_text(face = "bold", size = 11))
```

### Match the Correlations

The six scatterplots above have these r values (in random order): **-0.75, -0.48, 0.03, 0.28, 0.47, 0.85**

**1. Study Hours vs Exam Score:**

`r mcq(c("r = -0.75", "r = -0.48", "r = 0.03", "r = 0.28", "r = 0.47", answer = "r = 0.85"))`

**2. Screen Time vs Sleep Hours:**

`r mcq(c("r = -0.75", answer = "r = -0.48", "r = 0.03", "r = 0.28", "r = 0.47", "r = 0.85"))`

**3. Shoe Size vs IQ Score:**

`r mcq(c("r = -0.75", "r = -0.48", answer = "r = 0.03", "r = 0.28", "r = 0.47", "r = 0.85"))`

**4. Stress Level vs Wellbeing:**

`r mcq(c(answer = "r = -0.75", "r = -0.48", "r = 0.03", "r = 0.28", "r = 0.47", "r = 0.85"))`

**5. Coffees per Day vs Alertness:**

`r mcq(c("r = -0.75", "r = -0.48", "r = 0.03", answer = "r = 0.28", "r = 0.47", "r = 0.85"))`

**6. Exercise vs Mood:**

`r mcq(c("r = -0.75", "r = -0.48", "r = 0.03", "r = 0.28", answer = "r = 0.47", "r = 0.85"))`

::: {.callout-tip collapse="true"}
## Interpretation Tips

- **Strong correlations** (|r| > 0.5): Points cluster tightly around the line
- **Moderate correlations** (|r| = 0.3-0.5): Clear trend but more scatter
- **Weak correlations** (|r| < 0.3): Trend barely visible, lots of scatter
- **Near-zero correlations**: No discernible pattern at all
- **Negative correlations**: Line slopes downward (one goes up, other goes down)
:::

---

## Activity 3: Interpreting r Values (15 min)

### Cohen's Guidelines

Jacob Cohen provided rough guidelines for interpreting correlation strength:

| |r| value | Interpretation |
|-----------|----------------|
| 0.10 - 0.29 | Small/weak |
| 0.30 - 0.49 | Medium/moderate |
| 0.50 - 1.00 | Large/strong |

### Practice Interpretation

**1. r = 0.25 would be considered:**

`r mcq(c(answer = "Small/weak", "Medium/moderate", "Large/strong", "Negligible"))`

**2. r = -0.65 would be considered:**

`r mcq(c("Small/weak", "Medium/moderate", answer = "Large/strong", "Not applicable - it's negative"))`

**3. r = 0.42 would be considered:**

`r mcq(c("Small/weak", answer = "Medium/moderate", "Large/strong", "Negligible"))`

### Coefficient of Determination (r-squared)

Squaring r tells us the proportion of variance explained.

**4. If r = 0.50, then r-squared (r^2) equals:**

`r mcq(c("0.50", answer = "0.25", "0.75", "2.5"))`

**5. r^2 = 0.25 means:**

`r mcq(c(answer = "25% of the variance in Y is explained by X", "The correlation is 25%", "75% of participants show the relationship", "The effect is 25% larger than expected"))`

**6. If r = 0.70, what percentage of variance is explained?**

`r mcq(c("70%", answer = "49%", "30%", "100%"))`

### Critical Thinking

**7. A statistically significant correlation of r = 0.15 suggests:**

`r mcq(c("A strong, important relationship", answer = "A real but very weak relationship - likely not practically meaningful", "The data is unreliable", "We need to use a different test"))`

True or False: A significant p-value tells us the correlation is strong. `r torf(FALSE)`

True or False: Correlation does not imply causation. `r torf(TRUE)`

True or False: A negative correlation is weaker than a positive correlation of the same magnitude. `r torf(FALSE)`

---

## Activity 4: Correlation in YOUR Data (10 min)

### Applying to DataLab

Let's think about correlational questions we could ask with the DataLab data.

::: {.webex-check .webex-box}

### Question 1: Self-Rating and Performance

In the Frontal Lob task, participants rated their throwing ability before throwing.

**If high self-ratings are associated with better performance, what type of correlation would we expect?**

`r mcq(c(answer = "Positive correlation", "Negative correlation", "No correlation", "Cannot determine"))`

**Is there an IV and DV in this analysis?**

`r mcq(c("Yes - self-rating is IV, performance is DV", "Yes - performance is IV, self-rating is DV", answer = "No - correlational designs have no IV/DV, just two variables", "Depends on the hypothesis"))`
:::

::: {.webex-check .webex-box}

### Question 2: Sleep and Memory

Imagine we correlate hours of sleep the night before with memory test performance.

**What would a negative correlation (r = -0.30) mean?**

`r mcq(c("More sleep = better memory", answer = "More sleep = worse memory", "No relationship", "Sleep causes memory problems"))`

**Could we conclude that lack of sleep causes poor memory?**

`r mcq(c("Yes, the correlation proves it", answer = "No, correlation doesn't establish causation", "Only if p < 0.05", "Yes, if r is negative"))`
:::

::: {.webex-check .webex-box}

### Question 3: Third Variables

Sleep quality might be related to both coffee consumption AND memory performance.

**This illustrates:**

`r mcq(c("A direct causal relationship", answer = "A potential third variable (confound)", "That coffee improves memory", "That the correlation is invalid"))`

**If coffee consumption is related to both variables, it could:**

`r mcq(c(answer = "Create a spurious correlation between sleep and memory", "Prove that coffee causes poor sleep", "Have no effect on the correlation", "Make the correlation negative"))`
:::

### Thinking About Your Project

**Think about your group project research question:**

**1. Is your research question correlational or experimental?**

`r mcq(c(answer = "Correlational - looking at relationships between variables", "Experimental - manipulating an IV to see effects on a DV", "Descriptive - just describing one variable", "I'm not sure yet"))`

**2. If correlational, what two variables would you correlate?**

(Think about this for your own project - no MCQ here!)

**3. What would a positive correlation mean for your research question?**

`r mcq(c(answer = "Higher scores on variable 1 are associated with higher scores on variable 2", "The variables cause each other", "The relationship is strong", "The p-value is significant"))`

---

## Remaining Time: Research Plan Work (20+ min)

::: {.callout-warning}
## Research Plan Due Friday 27 February

Use the remaining lab time to work on your **Research Plan submission**. Your instructor is available to help!
:::

### What to Work On

If you haven't started:

1. Choose your research question from the approved topics
2. Identify your IV and DV
3. Decide on your design (between/within-subjects)

If you've already started:

1. Refine your hypotheses (H1 and H0)
2. Check your operationalisation is clear and measurable
3. Review the marking criteria on Learn

### Get Feedback Now

::: {.callout-tip}
## Ask Questions Today!

This is your last lab before Reading Week. Instructors can give feedback on:

- Whether your hypothesis is testable
- If your IV/DV are properly operationalised
- Whether your design makes sense for your question
:::

---

## Summary

::: {.callout-important}
## Key Takeaways

**Scatterplots:**

- Always visualise before calculating
- Each point = one participant
- Look for patterns, linearity, and outliers

**Pearson's r:**
- Ranges from -1 to +1
- Sign indicates direction (positive/negative)
- Magnitude indicates strength

**Interpretation:**
- Small: |r| = 0.10 - 0.29
- Medium: |r| = 0.30 - 0.49
- Large: |r| = 0.50+
- r^2 = proportion of variance explained

**Cautions:**
- Correlation does NOT equal causation
- Watch for third variables
- Statistical significance does not equal practical importance
:::

---

## Next Steps

1. Complete the [Week 15 Self-Assessment](self-assessment.qmd)
2. Review the [Lecture Reading](reading.qmd)
3. Prepare for Week 16: Project Groups - Literature Review
